### 2016-01-30 Documents/paper.Rmd
initial commit

---
title: "Issues in developing mulitvariable models for treatment selection"
author: "Michael C Sachs and Lisa M McShane"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
    html_document:
      pandoc_args: ["--bibliography", "references.bib"]
---

# Introduction

## Predictive biomarkers

In what ways can it change clinical practice?

@janes2011measuring, @janes2014approach,  @mcshane2013development, @polley2013statistical

## Overview of biomarker signatures

- Do I include all features or a subset?
    + Which subset?
    + How do I combine the features?
- Transformations?
- Weight and combine with regression coefficients?
- What coefficients?
- Thresholds/Cutoffs before or after combining?

References on developing signatures @simon2003pitfalls, @hastie2009elements, @moons2012riskI, @moons2012riskII

## Data analysis example

@zhu2010prognostic

# Issues

## Estimating Prediction Error

## Estimating Clinical Benefit

## Developing and validating a predictive biomarker for an effective treatment

Identify the non-responders and demonstrate that they do not benefit (hard)

## Developing and validating a predictive biomarker for an ineffective treatment

Identify the responders and demonstrate that they do benefit (no quite as hard)

# Conclusion

### 2016-01-30 Code/01-munge.R

@@ -1,0 +1,80 @@
+library(Biobase)
+library(GEOquery)
+library(rio)
+library(sva)
+# load series and platform data from GEO
+
+gset <- getGEO("GSE14814", GSEMatrix =TRUE)
+if (length(gset) > 1) idx <- grep("GPL96", attr(gset, "names")) else idx <- 1
+gset <- gset[[idx]]
+
+
+gdat <- as(gset, "data.frame")
+gdat$ID <- sampleNames(phenoData(gset))
+
+gexpr <- exprs(gset)
+fdata <- fData(gset)
+#fdex <- fdata$ID[fdata$ENTREZ_GENE_ID != "" & fdata$`Gene Ontology Cellular Component` != ""]
+#gexpr <- gexpr[fdex, ]
+
+anyna <- rowSums(is.na(gexpr)) > 0
+gexpr <- gexpr[!anyna, ]
+## 19,619 genes, close enough
+
+parse_pheno <- function(x){
+
+  pair <- strsplit(paste(x), ": ", fixed = TRUE)
+  if(length(pair[[1]]) == 1){
+    x
+  } else {
+    name <- pair[[1]]
+    name <- name[!is.na(name)][1]
+
+    val <- sapply(pair, "[", 2)
+
+  ## try converting to numeric
+
+    nval <- as.numeric(val)
+    if(all(is.na(nval))) nval <- val
+
+    out <- data.frame(nval)
+    colnames(out) <- name
+    out
+}
+
+}
+
+pheno0 <- gdat[, c("ID", grep("characteristics", colnames((gdat[, (ncol(gdat) - 55):ncol(gdat)])[1:4, ]), value = TRUE))]
+pheno0$characteristics_ch1.12 <- paste(pheno0$characteristics_ch1.12)
+pheno0$characteristics_ch1.12[pheno0$characteristics_ch1.12 == ""] <- paste(pheno0$characteristics_ch1.11[pheno0$characteristics_ch1.12 == ""])
+pheno0$characteristics_ch1.11 <- NULL
+
+pheno <- as.data.frame(lapply(pheno0, parse_pheno))
+
+#gdat <- merge(gdat, pheno, by = "ID")
+gdat$batch <- ifelse(gdat$data_processing == "The data were preprocessed using RMA v0.5.", -1, 1)
+
+## remove batch effect
+
+modcom <- model.matrix( ~ 1, data = gdat)
[…]

### 2016-01-30 Code/02-reproduce-Zhu.R

@@ -1,0 +1,61 @@
+library(survival)
+
+load("../Data/jbl10-data-2015-12-17.RData")
+
+## develop signature in 62 OBS patient subset
+
+psdex <- 2:(grep("tissue", colnames(gdat)) - 1)
+colnames(gdat)[psdex] <- paste0("X", colnames(gdat)[psdex])
+
+table(gdat$Post.Surgical.Treatment)
+obsgrp <- subset(gdat, Post.Surgical.Treatment == "OBS")
+
+## identify subset of 172 probes with P < .005 in cox model with outcome
+# results from Zhu, supplementary table 6S
+
+zhu <- readLines("../Data/sig172-probes-zhu.txt")
+header <- zhu[1:8]
+data0 <- zhu[-c(1:8)]
+data1 <- unlist(strsplit(data0, split = " ", fixed = TRUE))
+data2 <- as.data.frame(matrix(data1, ncol = 8, byrow = TRUE), stringsAsFactors = FALSE)
+data2[, 5:8] <- lapply(data2[, 5:8], as.numeric)
+colnames(data2) <- header
+
+## can i reproduce it?
+
+obsgrp[, psdex] <- lapply(obsgrp[, psdex], function(x) (x - mean(x, na.rm = TRUE))/sd(x, na.rm = TRUE))
+
+ps172dex <- NULL
+for(i in psdex){
+
+  form <- as.formula(paste("Surv(OS.time, OS.status == 'Dead') ~ ", colnames(obsgrp)[i]))
+  fit <- tryCatch(coxph(form, data = obsgrp, y = FALSE, ties = "breslow", robust = FALSE), error = function(e) NA)
+  if(is.na(fit[[1]])) next
+  fs <- summary(fit)$coefficients
+  ps172dex <- rbind(ps172dex, data.frame(probe = colnames(obsgrp)[i], index = i,
+                                p.value = fs[5], hr = fs[2], lhr = exp(fs[1] - 1.96*fs[3]),
+                                uhr = exp(fs[1] + 1.96*fs[3]), stringsAsFactors = FALSE))
+
+}
+
+
+ps172dex$probe2 <- ifelse(substr(ps172dex$probe, 1, 1) == "X",
+                         substr(ps172dex$probe, 2, nchar(ps172dex$probe)), ps172dex$probe)
+
+
[…]

### 2016-02-05 Documents/paper.Rmd
progress

@@
 output: 
-    html_document:
-      pandoc_args: ["--bibliography", "references.bib"]
+    pdf_document:
+      pandoc_args: ["--bibliography", "bookchapter-1.bib"]
 ---
@@
 
-## Predictive biomarkers
+```{r setup, include = FALSE}
+library(knitr)
+opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE)
+```
 
+Omics technologies that generate a large amount of molecular data about biospecimens have the potential to provide information about a patient's disease characteristics above and beyond standard clinical and pathological features.
+By combining the information from a large amount of molecular features into a multivariable model, hereafter referred to as a _biomarker signature_, there is the opportunity to identify distinct subgroups of patients for whom treatment decisions can be personalized.
+A multivariable biomarker can guide the decisions to treat or not to treat and help identify the patients who are most likely to survive.
+The key challenge we address in this paper is to estimate the precise combination of features from a high dimensional molecular assay suited to the clinical context.
+
+## Terminology and Notation
+
+A **biomarker signature** is a transformation of multiple individual features, typically molecular characterstics measured on a multiplex assay, to a one-dimensional space.
+Specifically, let $X$ denote the set of $p$ features under consideration.
+The signature is an unknown function $f(X): \mathbb{R}^p \mapsto \mathbb{R}^1$.
+The signature may be continuous, take multiple discrete values, or be dichotomous.
+
+Let $S$ denote the development dataset, which includes $X$, an outcome $Y$, a treatment $Z$, and possibly other variables.
+$S$ is a sample of size $n$ from distribution $\mathcal{P}$ with domain $\mathcal{X}$.
+Let $\mathcal{F}$ be a mapping from $\mathcal{X}$ to the space of continuous functions with domain $\mathbb{R}^p$ and range $\mathbb{R}$, $\mathcal{D}$.
+Thus $\mathcal{F}: \mathcal{X} \mapsto \mathcal{D}$ denotes the process or algorithm through which a particular $f$ is estimated.
+We do not place any other restrictions on $\mathcal{F}$, it could be a clustering approach, a regression approach, a combination of both, or something else entirely.
+We will use $\mathcal{F}$ to denote the manner in which $f$ is estimated and will write $f \in \mathcal{F}$ to denote that $f$ is estimated with the class of methods $\mathcal{F}$.
+
+Let $\phi: \mathcal{D} \times \mathcal{X} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC).
+This is a function of both $f$ and $S$.
+We are interested in estimating $E_\mathcal{P}[\phi_f(S)]$, which is the expected error under the data generation mechanism, for a particular $f \in \mathcal{F}$.
+This allows us to understand how the signature will perform on future observations generated from $\mathcal{P}$.
+We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ *for all* $f \in \mathcal{F}$, which is the generalization error for $f$ generated using mechanism $\mathcal{F}$.
+This doesn't guide outside researchers as to which specific $f$ to use, yet it is useful for development because it tells us how much signal is in the data.
+As shorthand we will write this as $E_\mathcal{P}[\phi_\mathcal{F}(S)]$
+
+A signature that **reliably predicts** an outcome $Y$ is one that has generalization error small enough for the clinical context.
+Such a signature may be useful for treatment selection, prognosis, or other type of clinical management.
+
+## Overview of biomarker signature development
+
+The goal of the development phase of a biomarker signature is to provide a valid estimate of the performance of $f \in \mathcal{F}$.
+Optionally, one can provide a specification of $f$ for others to use, and we want that particular $f$ to be estimated as precisely as possible.
+Typically, a specific $f$ is estimated using $\mathcal{F}$ based on some training data.
+This can be done using a variety of different methods.
+In recent years there have been an explosion in the literature of computational approaches to classification and prediction, and we do not intend to summarize them all here.
+Some excellent reviews of such approaches are @hastie2009elements and @moons2012riskI.
+The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply, determining how to combine the features, and whether to apply thresholds/cutoffs to the resulting signature.
+
+Signatures can be estimated by identifying naturally occuring clusters, or intrinsic subtypes using $X$ alone without regard to the outcome $Y$.
+This was the case with the PAM-50 gene signature [reference needed], and later, the intrinsic subtypes were shown to be strongly associated with clinical outcomes.
+Supervised learning techniques can also be used to identify signatures, as was the case with Oncotype DX [@paik2004multigene].
+In this case, regression-based methods are used to estimate a model that is highly predictive for the outcome $Y$.
+In the case of Oncotype DX, the outcome in question was recurrence of breast cancer.
+
+Another approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment.
+It is possible, and quite common in high-dimensional settings, to combine multiple approaches to estimating $f$.
+For instance, a data-reduction step by variable selection or clustering may be performed before doing regression analysis on the resulting components.
+
+No matter what the particular model building method is, our main concern and focus of this paper is with obtaining a valid estimate of its performance, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
+This depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used.
+An optional component of the development phase is to provide a specification of $f$ for others to use on independent data or in clinical practice.
+
+## Biomarker signatures in clinical practice
+
+A biomarker signature can inform clinical practice in a number of ways. 
 In what ways can it change clinical practice?
@@
 
-## Overview of biomarker signatures
-
-- Do I include all features or a subset?
-    + Which subset?
-    + How do I combine the features?
-- Transformations?
-- Weight and combine with regression coefficients?
-- What coefficients?
-- Thresholds/Cutoffs before or after combining?
-
-References on developing signatures @simon2003pitfalls, @hastie2009elements, @moons2012riskI, @moons2012riskII
-
 ## Data analysis example
 
-@zhu2010prognostic
+Throughout this paper, we reanalyze data from @zhu2010prognostic.
+Briefly, the data of interest are from the JBR.10 trial, which was a randomized controlled trial of adjuvant vinorelbine/cisplatin (ACT) versus observation alone (OBS) in 482 participants with non small cell lung cancer (NSCLC).
+Of those 482 participants, 169 had frozen tissue collected, and of those samples, 133 (71 in ACT and 62 in OBS) had gene-expression profiling done using U133A oligonucleotide microarrays (Affymetrix, Santa Clara, CA).
+
+The goal of the @zhu2010prognostic paper was to identify a multi-gene signature that strongly predicts prognosis, and the hypothesis was that the poor prognosis subgroup would benefit more from ACT compared to the good prognosis subgroup.
+The annotated gene expression data and clinical information are available from the Gene Expression Omnibus (identifier: GSE14814, @edgar2002gene).
+Batch effects were removed using the ComBat function in the sva R package [@leeksva] and then the gene expression values were centered by their means and scaled by their standard deviations.
+We reproduce their analysis while illustrating the key issues under discussion.
 
@@
 
-## Estimating Clinical Benefit
+```{r funcs}
+print("dick butt")
+```
 
-## Developing and validating a predictive biomarker for an effective treatment
+Recall that the main goal is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected error on future observations for $f \in \mathcal{F}$.
+This can be estimated with the in-sample empirical estimate: $\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)$ for a particular $f$.
+However, if $S$ is used to estimate $f$ then the estimate will be biased due to overfitting, that is, $|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|$ will be large.
+This is because $\phi$ depends on $f$, and thus the statistic $\phi$ is being adaptively defined based on the observed data $S$, hence causing the overfitting.
+
+A traditional remedy to this problem of overfitting is the split sample approach.
+First, randomly partition $S$ into the training sample $S_t$ and the holdout sample $S_h$ with sample sizes $n_t$ and $n_h$, respectively.
+Then, $S_h$ is hidden from the analyst while $\mathcal{F}$ is applied to $S_t$ to estimate the signature function $f_t$.
+Then, with $f_t$, and therefore $\phi$ fixed, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of both $E_{\mathcal{P}}[\phi_{f_t}(S)]$ and $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
+As a side-effect, the specific form of $f_t$ that is fixed using the $S_t$ partition can be reported as the function for others to use, therefore the aforementioned estimator is also an estimate of the error for that specific $f_t$.
+The drawback of the split-sample approach is that $f$ is not estimated as precisely as it would be compared to using the entire dataset $S$ and for the same reason neither is $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
+@dobbin2011optimally investigate how to optimally split a dataset into training and holdout partitions.
+
+Another approach to avoid overfitting is cross-validation, which is a resampling based approach.
+For a fixed integer $k$, which can be between 1 and $n$, we randomly select a partition of $k$ observations from $S$, denoted $S_k$.
+Then $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the the subset of $S$ that is disjoint from $S_k$.
+Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an unbiased estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
+This process is repeated $K$ times to yield $K$ estimates.
+Each of these estimates is unbiased, but noisy, because typically $k$ is very small relative to $n$.
+Thus, we average over $K$ to get a less noisy estimate.
+This process is called "leave $k$ out" cross-validation.
+Note that for each partition that is selected, we obtain a new estimate of $f$, therefore the estimator is an estimator only for $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
+Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$.
+
+A variation on the cross-validation approach is bootstrapping.
+In that case, a sample $S_b$ of size $n$ is sampled _with replacement_ from $S$.
+Then $f_{b}$ is estimated and fixed by applying $\mathcal{F}$ to $S_{b}$.
+The error $\phi$ is estimated on the subset of $S$ that is disjoint from $S_b$: $S_{-b}$.
+This process is repeated $K$ times to yield $K$ estimates.
+@efron1997improvements showed that these estimates are still biased due to the sampling with replacement.
+Thus they propose the 0.632 estimate:
+$$
+\hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
+$$
+where $\hat{E}[\phi_{f}(S)]$ is the naive estimate of $\phi_f$ using the entire dataset.
+
+To illustrate the different properties of these estimates and how they deal with overfitting, we conduct a simulation study.
+Data were generated with $n$ samples, each with a binary outcome $Y$ with prevalence 0.3, and $d$ features sampled from the standard normal distribution.
+This is the null case where no features are associated with $Y$.
+The signature estimation procedure entails a feature selection step, in which each feature is regressed against $Y$ in a logistic regression model.
+The 25 features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines our final signature.
+
+We compare each of the three methods described above along with the biased approaches of using the full sample to select the features, followed by either fitting the model on a split sample, or using fitting the full model inside the cross validation step.
+Note that @zhu2010prognostic used the latter approach, which explains the differences in our results.
+Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
+
+## Evaluating Clinical Benefit
+
+While a biomarker signature may accurately predict a clinical outcome, that does not neccessarily imply that the signature is clinically useful.
+
+### Developing a biomarker signature in a single-arm trial
+
+Classify the binary response to treatment.
+
+### Developing and validating a predictive biomarker for an effective treatment
 
@@
 
-## Developing and validating a predictive biomarker for an ineffective treatment
+### Developing and validating a predictive biomarker for an ineffective treatment
 
@@
 # Conclusion
+
+What I said and how I said it.
+
+# References

### 2016-02-05 Code/02-reproduce-Zhu.R

@@ -1,6 +1,6 @@
 library(survival)
 
-load("../Data/jbl10-data-2015-12-17.RData")
+load("../Data/jbl10-data-2016-02-04.RData")
 
 ## develop signature in 62 OBS patient subset
 
@@ -12,7 +12,7 @@ obsgrp <- subset(gdat, Post.Surgical.Treatment == "OBS")
 
 ## identify subset of 172 probes with P < .005 in cox model with outcome
 # results from Zhu, supplementary table 6S
-
+if(FALSE){
 zhu <- readLines("../Data/sig172-probes-zhu.txt")
 header <- zhu[1:8]
 data0 <- zhu[-c(1:8)]
@@ -20,7 +20,7 @@ data1 <- unlist(strsplit(data0, split = " ", fixed = TRUE))
 data2 <- as.data.frame(matrix(data1, ncol = 8, byrow = TRUE), stringsAsFactors = FALSE)
 data2[, 5:8] <- lapply(data2[, 5:8], as.numeric)
 colnames(data2) <- header
-
+}
 ## can i reproduce it?
 
 obsgrp[, psdex] <- lapply(obsgrp[, psdex], function(x) (x - mean(x, na.rm = TRUE))/sd(x, na.rm = TRUE))
@@ -43,17 +43,13 @@ ps172dex$probe2 <- ifelse(substr(ps172dex$probe, 1, 1) == "X",
                          substr(ps172dex$probe, 2, nchar(ps172dex$probe)), ps172dex$probe)
 
 
-matches <- merge(data2, ps172dex, by.x = "Probe Set ID", by.y = "probe2", all.x = TRUE, all.y = FALSE)
-
-plot(hr ~ HR, data = matches)
-
-
 ###
 
 table(ps172dex$p.value < .005)
 
+matches <- subset(ps172dex, p.value < .005)
+rownames(matches) <- matches$probe
 ## now do the forwards and backward selection
-
 subgrp <- obsgrp[, c("ID", paste(matches$probe), "OS.time", "OS.status")]
 
 

### 2016-02-10 Code/01-munge.R
signature building

@@ -60,10 +60,18 @@ modcom <- model.matrix( ~ 1, data = gdat)
 batchadj <- ComBat(gexpr, batch = gdat$batch, mod = modcom, prior.plots = TRUE)
 
 expradj <- as.data.frame(t(batchadj))
+
+## pca
+pca.fit <- prcomp(t(batchadj), center = TRUE, scale. = TRUE)
+pca.dat <- as.data.frame(pca.fit$x[, 1:25])
+pca.dat$ID <- rownames(pca.dat)
+##
+
 expradj$ID <- rownames(expradj)
 
 gdat <- merge(expradj, pheno, by = "ID")
 
+pca.dat2 <- merge(pca.dat, pheno, by = "ID")
 ## make sure no merging errors in data
 
 pheno.xls <- import("../Data/Zhu_2010_JCO_microarray_pts_clin_info.xls", sheet = 1)
@@ -75,6 +83,6 @@ check2 <- subset(check, OS.time == `OS time`)
 # we good
 
 save(gdat, file = paste0("../Data/jbl10-data-", Sys.Date(), ".RData"))
-
+save(pca.dat2, file = paste0("../Data/jbl10-pca-", Sys.Date(), ".RData"))
 
 

### 2016-02-10 Code/03-new-signature-zhu.R

@@ -1,0 +1,100 @@
+library(survival)
+library(tidyr)
+library(dplyr)
+
+load("../Data/jbl10-data-2016-02-09.RData")
+
+## supervised pca fit for heldout
+##    - select x most correlated genes
+##    - subset those, form 3 pcas
+##    - Fit multivariate model
+## cutpoint at median
+## predict for held in dataset
+
+scl.cen <- function(x){
+
+  (x - mean(x, na.rm = TRUE))/sd(x, na.rm = TRUE)
+
+}
+
+## dichotomize survival
+gdat$DSS.status[gdat$DSS.status == "dead"] <- "Dead"
+gdat$surv2.5 <- with(gdat, ifelse(DSS.time >= 5, 1, -1))
+psdex <- 2:(grep("tissue", colnames(gdat)) - 1)
+gdat[, psdex] <- lapply(gdat[, psdex], scl.cen)
+ldat <- gather_(gdat, "gene", "expression", colnames(gdat[, psdex]))
+ldat <- ldat %>% group_by(gene) %>% mutate(zexpression = (expression - mean(expression, na.rm = TRUE))/sd(expression, na.rm = TRUE))
+cldat <- gdat[, -psdex]
+
+fit.superpc <- function(dat0, cldat){
+
+  cors <- dat0 %>% group_by(gene) %>% summarize(cor = cor(zexpression, surv2.5))
+  ## select number of genes
+
+  n.gen.cands <- 3:100
+
+  p.ngenes <- sapply(n.gen.cands, function(n.genes){
+
+    sel.genes <- cors$gene[rank(-cors$cor) <= n.genes]
+
+    pcas <- dat0 %>% select(ID, gene, zexpression) %>% filter(gene %in% sel.genes) %>% spread(gene, zexpression)
+    pcamat <- as.matrix(pcas[, -1])
+    rownames(pcamat) <- pcas$ID
+
+    pc.fit <- prcomp(pcamat, center = FALSE, scale. = FALSE)
+    pc.dat <- as.data.frame(pc.fit$x[, 1:3])
+    pc.dat$ID <- rownames(pc.dat)
+
+    pdonk <- merge(cldat, pc.dat, by = "ID")
+    fit.cox <- coxph(Surv(DSS.time, DSS.status == "Dead") ~ PC1 + PC2 + PC3, data = pdonk)
[…]

### 2016-02-19 Documents/paper.Rmd
lisa review
pca selection is a bust

@@
       pandoc_args: ["--bibliography", "bookchapter-1.bib"]
+fontsize: 11pt
+geometry: margin=1in
 ---
+
+\linespread{2}
 
@@
 ```{r setup, include = FALSE}
+library(ggplot2)
 library(knitr)
 opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE)
+knit_hooks$set(plot = function(x, options) {
+  if ('tikz' %in% options$dev && !options$external) {
+    hook_plot_tex(x, options)
+  } else hook_plot_md(x, options)
+})
 ```
@@
 
-A biomarker signature can inform clinical practice in a number of ways. 
-In what ways can it change clinical practice?
+A biomarker signature can inform clinical practice in a number of ways, regardless of how the signature was developed.
+A highly prognostic signature may identify a subpopulation that has such a good chance of long term survival, that they do not need to undergo treatment that carries risks and side effects such as chemotherapy.
+In the context of a specific therapy that targets a particular molecular pathway, a signature may identify a subpopulation that does not benefit from that therapy, thereby guiding the decision to treat or not.
+A signature that was developed to identify intrinsic molecular subtypes could be both prognostic and informative about the effectiveness of a therapy.
+
+The key point is that signatures are useful if they can correctly and reliably classify patients into distinct subgroups for which different treatment decisions would be made.
+There are two distinct but related statistical concepts involved here: calibration or accuracy, and discrimination.
+Signatures are often optimized to be well-calibrated, that is, highly accurate for predicting outcomes.
+However if the signature does not separate a population into distinct subgroups, then it is unlikely to be informative enough to change clinical practice.
+
+In the development process it is important to evaluate both of these statistical concepts.
+Furthermore, it is not trivial to assess each of these in a valid manner when the data are used to define the signature itself.
+We illustrate the potential for bias, and remedies, in our examples.
 
@@
 The goal of the @zhu2010prognostic paper was to identify a multi-gene signature that strongly predicts prognosis, and the hypothesis was that the poor prognosis subgroup would benefit more from ACT compared to the good prognosis subgroup.
+The signature was trained to predict disease specific survival.
 The annotated gene expression data and clinical information are available from the Gene Expression Omnibus (identifier: GSE14814, @edgar2002gene).
@@
 
+The authors of @zhu2010prognostic present results that mainly focus on the discrimination ability of their estimated signature.
+They do that by demonstrating that the two risk subgroups predicted by their signature (high risk and low risk) have separation in their survival curves and that the hazard ratio for their signature is large and significant even when adjusting for other risk factors.
+They do not directly address calibration, that is, whether their signature accurately predicts survival times.
+
 # Issues
 
-## Estimating Prediction Error
-
-```{r funcs}
-print("dick butt")
-```
-
-Recall that the main goal is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected error on future observations for $f \in \mathcal{F}$.
+Recall that the main goal is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected value of a given statistic on future observations for $f \in \mathcal{F}$.
 This can be estimated with the in-sample empirical estimate: $\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)$ for a particular $f$.
@@
 This is because $\phi$ depends on $f$, and thus the statistic $\phi$ is being adaptively defined based on the observed data $S$, hence causing the overfitting.
+
+In many cases during signature development, the statistic $\phi$ in question is a statistic that relates to calibration, such as classification accuracy, correlation or mean squared error.
+While a biomarker signature may accurately predict a clinical outcome, that does not neccessarily imply that the signature is clinically useful.
+As we mentioned before, two statistical critera are necessary for determining clinical usefulness: calibration and discrimination.
+To assess discrimination, a different statistic $\phi$ may be used, such as an odds ratio, hazard ratio, or difference in survival probabilities.
+Evaluation of $\phi$ is also subject to bias due to overfitting, a fact that is commonly overlooked in the medical literature.
+
+In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups as defined by the JBL.10 signature $f$.
+The risk groups were determined by the signature $f$, which was estimated using $S$ the same data that is then used to estimate the hazard ratio which is estimated to be 15.
+They go on to assess the discrimination of the JBL.10 signature in a series of independent data sets, in which they find hazard ratios around 2.
+We will reanalyze the dataset and illustrate some remedies for avoiding bias in determining the discrimination (an additionally the calibration) of this signature.
+
+@altman2000we, @buyse2007towards, @moons2012riskI, @moons2012riskII
+
+## Remedies to Overfitting
+
+```{r funcs}
+print("Insert simulation results")
+```
 
@@
 This process is repeated $K$ times to yield $K$ estimates.
-@efron1997improvements showed that these estimates are still biased due to the sampling with replacement.
-Thus they propose the 0.632 estimate:
+These $K$ estimates are averaged to obtain the mean over the bootstrap replicates.
+@efron1997improvements suggest a variation, the 0.632 estimate:
 $$
@@
 Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
+In our simulation, we look at two different statistics, classification accuracy to assess calibration and the odds ratio for the outcome comparing the signature groups to assess discrimination.
 
-## Evaluating Clinical Benefit
+## Evaluating Calibration
 
-While a biomarker signature may accurately predict a clinical outcome, that does not neccessarily imply that the signature is clinically useful.
-
-### Developing a biomarker signature in a single-arm trial
-
-Classify the binary response to treatment.
+## Evaluating Discrimination and Clinical Benefit
 

### 2016-02-19 Code/03-new-signature-zhu.R

@@ -28,42 +28,52 @@ cldat <- gdat[, -psdex]
 
 fit.superpc <- function(dat0, cldat){
 
-  cors <- dat0 %>% group_by(gene) %>% summarize(cor = cor(zexpression, surv2.5))
+  cors <- dat0 %>% group_by(gene) %>%
+    do(data.frame(
+      t(summary(coxph(Surv(DSS.time, DSS.status == "Dead") ~ zexpression,
+                           data = .,
+                           eps = 1e-4))$coefficients[c(1,5)]))
+      )
+
   ## select number of genes
 
-  n.gen.cands <- 3:100
+  cors <- cors[unlist(cors$X2) < .005, ]
+  n.gen.cands <- 5:nrow(cors)
 
   p.ngenes <- sapply(n.gen.cands, function(n.genes){
 
-    sel.genes <- cors$gene[rank(-cors$cor) <= n.genes]
+    sel.genes <- cors$gene[rank(-abs(cors$X1)) <= n.genes]
+    coeffs <- cors$X1[rank(-abs(cors$X1)) <= n.genes]
 
-    pcas <- dat0 %>% select(ID, gene, zexpression) %>% filter(gene %in% sel.genes) %>% spread(gene, zexpression)
-    pcamat <- as.matrix(pcas[, -1])
+    pcas <- dat0 %>% select(ID, gene, zexpression) %>%
+      filter(gene %in% sel.genes) %>% spread(gene, zexpression)
+    pcamat <- as.matrix(pcas[, -1]) * matrix(rep(coeffs, nrow(pcas)), nrow = nrow(pcas), byrow = TRUE)
     rownames(pcamat) <- pcas$ID
 
-    pc.fit <- prcomp(pcamat, center = FALSE, scale. = FALSE)
-    pc.dat <- as.data.frame(pc.fit$x[, 1:3])
+    pc.fit <- prcomp(pcamat, center = TRUE, scale. = TRUE)
+    pc.dat <- as.data.frame(pc.fit$x[, 1:5])
     pc.dat$ID <- rownames(pc.dat)
 
     pdonk <- merge(cldat, pc.dat, by = "ID")
-    fit.cox <- coxph(Surv(DSS.time, DSS.status == "Dead") ~ PC1 + PC2 + PC3, data = pdonk)
-    summary(fit.cox)$logtest[3]
[…]

### 2016-02-23 Documents/paper.Rmd
new method working for sig devel

@@
 
-```{r funcs}
-print("Insert simulation results")
-```
-
 A traditional remedy to this problem of overfitting is the split sample approach.
@@
 
+```{r cvsims}
+load("../Code/cvsim-result.RData")
+cvres <- do.call("rbind", cvsims)
+library(tidyr)
+
+cvlong <- do.call("rbind", lapply(1:ncol(cvres), function(i){
+
+  cl <- cvres[, i]
+  splow <- strsplit(colnames(cvres)[i], ".", fixed = TRUE)
+  nm <- paste(unlist(sapply(splow, function(s) rev(rev(s)[-1]))), collapse = ".")
+  cls <- sapply(splow, function(s) rev(s)[1])
+  data.frame(value = cvres[, i], stat = cls, scen = nm, stringsAsFactors = FALSE)
+
+}))
+
+ggplot(subset(cvlong, stat == "AUC"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") + 
+  theme_bw(base_size = 13, base_family = "serif") + 
+  scale_x_discrete("Estimation Approach", labels = c("Bootstrap", "Leave 10 out CV", "Leave 100 out CV", 
+                                                     "30% Holdout", "50% Holdout", "Resubstitution", "Partial CV", "Partial Holdout")) + 
+  ylab("Area Under the ROC Curve")
+```
+
 ## Evaluating Calibration

### 2016-02-23 Code/03-new-signature-zhu.R

@@ -37,62 +37,73 @@ fit.superpc <- function(dat0, cldat){
 
   ## select number of genes
 
-  cors <- cors[unlist(cors$X2) < .005, ]
-  n.gen.cands <- 5:nrow(cors)
+  cors <- cors[cors$X2 < .005, ]
+  seldat0 <- dat0 %>% select(ID, gene, zexpression) %>%
+    filter(gene %in% cors$gene) %>% spread(gene, zexpression)
 
-  p.ngenes <- sapply(n.gen.cands, function(n.genes){
+  seldat0[, 2:ncol(seldat0)] <- lapply(2:ncol(seldat0), function(it){
 
-    sel.genes <- cors$gene[rank(-abs(cors$X1)) <= n.genes]
-    coeffs <- cors$X1[rank(-abs(cors$X1)) <= n.genes]
-
-    pcas <- dat0 %>% select(ID, gene, zexpression) %>%
-      filter(gene %in% sel.genes) %>% spread(gene, zexpression)
-    pcamat <- as.matrix(pcas[, -1]) * matrix(rep(coeffs, nrow(pcas)), nrow = nrow(pcas), byrow = TRUE)
-    rownames(pcamat) <- pcas$ID
-
-    pc.fit <- prcomp(pcamat, center = TRUE, scale. = TRUE)
-    pc.dat <- as.data.frame(pc.fit$x[, 1:5])
-    pc.dat$ID <- rownames(pc.dat)
-
-    pdonk <- merge(cldat, pc.dat, by = "ID")
-    fit.cox <- coxph(Surv(DSS.time, DSS.status == "Dead") ~ PC1 + PC2 + PC3 + PC4 + PC5, data = pdonk)
-    summary(fit.cox)$concordance[1]
+    unlist(cors %>% filter(gene == colnames(seldat0[, it])) %>% `$`("X1") * seldat0[, it])
 
   })
 
-  n.gene.sel <- n.gen.cands[which(order(-abs(p.ngenes)) == 1)]
-  sel.genes <- cors$gene[rank(-abs(cors$X1)) <= n.gene.sel]
+  seldat0 <- merge(seldat0, cldat, by = "ID", all.y = FALSE)
 
-  pcas <- dat0 %>% select(ID, gene, zexpression) %>%
-    filter(gene %in% sel.genes) %>% spread(gene, zexpression)
-  pcamat <- as.matrix(pcas[, -1])
-  rownames(pcamat) <- pcas$ID
+  ngen.max <- 25
 
[…]

### 2016-03-03 Documents/paper.Rmd
completed draft

--- 2016-02-19 Documents/paper.Rmd
@@
       pandoc_args: ["--bibliography", "bookchapter-1.bib"]
-fontsize: 11pt
+fontsize: 12pt
 geometry: margin=1in
@@
 ```{r setup, include = FALSE}
+library(dplyr)
 library(ggplot2)
 library(knitr)
-opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE)
+library(survival)
+library(ggkm)
+library(tidyr)
+
+opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE, warning = FALSE)
 knit_hooks$set(plot = function(x, options) {
@@
 Let $\phi: \mathcal{D} \times \mathcal{X} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC).
+This could also be a measure of discrimination, such as an odds ratio, hazard ratio, or log-rank statistic.
 This is a function of both $f$ and $S$.
@@
 
-@janes2011measuring, @janes2014approach,  @mcshane2013development, @polley2013statistical
+@janes2011measuring, @janes2014approach,  @mcshane2013development
 
@@
 Batch effects were removed using the ComBat function in the sva R package [@leeksva] and then the gene expression values were centered by their means and scaled by their standard deviations.
-We reproduce their analysis while illustrating the key issues under discussion.
+We expand on the main features of their analysis while illustrating the key issues under discussion.
 
@@
 They do not directly address calibration, that is, whether their signature accurately predicts survival times.
+
+Our signature estimation approach is similar but not identical to that in @zhu2010prognostic.
+Following the processing steps described above, we performed a gene selection step wherein we fit univariate Cox regression models with disease specific survival as the outcome and each gene as the single predictor.
+Genes with univariate p-values less than 0.005 were selected for further analysis.
+Then, each gene was weighted by its univariate Cox regression coefficient, and the resulting weighted gene expression values summed to form risk scores.
+Genes were selected in a forward selection manner, starting with the most significant genes, the gene that improved the concordance between survival times and the risk score was selected.
+If no gene improved the concordance, the process was stopped.
+The final list of selected genes were all included in a multivariable Cox regression model to fit the final risk score.
+The cutoff that yielded the smallest log-rank statistic p-value was used to dichotomize into two risk groups.
 
@@
 Then $f_{b}$ is estimated and fixed by applying $\mathcal{F}$ to $S_{b}$.
-The error $\phi$ is estimated on the subset of $S$ that is disjoint from $S_b$: $S_{-b}$.
+The statistic $\phi$ is estimated on the subset of $S$ that is disjoint from $S_b$: $S_{-b}$.
 This process is repeated $K$ times to yield $K$ estimates.
@@
 
+Another variation on all of these methods is the concept of pre-validation [@tibshirani2002pre].
+With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted predictor $\hat{f}(X_i)$ is estimated for $X_i \in S_{b}$ where $\hat{f}$ is estimated using $S_{-b}$.
+This process is repeated to obtain a set of pre-validated predictor estimates $\hat{f}$ which are then used to calculate $\phi$.
+For single-step holdout, this process is equivalent to what is described above.
+For cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on the held-out dataset.
+
+## Simulation Study
+
 To illustrate the different properties of these estimates and how they deal with overfitting, we conduct a simulation study.
@@
 This is the null case where no features are associated with $Y$.
-The signature estimation procedure entails a feature selection step, in which each feature is regressed against $Y$ in a logistic regression model.
+The signature estimation procedure entails a feature selection step, in which each feature is regressed against $Y$ in a univariate logistic regression model.
 The 25 features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines our final signature.
 
-We compare each of the three methods described above along with the biased approaches of using the full sample to select the features, followed by either fitting the model on a split sample, or using fitting the full model inside the cross validation step.
+We compare each of the methods described above along with the biased approaches of using the full sample to select the features, followed by either fitting the model on a split sample, or using fitting the full model inside the cross validation step.
+These are referred to as "partial holdout" and "partial CV" in the tables and figures.
 Note that @zhu2010prognostic used the latter approach, which explains the differences in our results.
+We also implemented the naive resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is trained on a holdout set and then evaluated on the combined complete dataset.
+The partial resubstitution approach is prevalent in the literature, such as the @zhu2010prognostic paper, wherein the model was trained on the one arm of the trial, and then the results were presented for the combined arms.
 Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
-In our simulation, we look at two different statistics, classification accuracy to assess calibration and the odds ratio for the outcome comparing the signature groups to assess discrimination.
+In our simulation, we look at two different statistics, the area under the ROC curve (AUC) to assess calibration and the odds ratio for the outcome comparing the signature groups to assess discrimination.
 
-## Evaluating Calibration
+```{r cvsims, fig.cap = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) in the setting where a dataset is used to both define and evaluate the signature.
+The violin plots show mirrored density estimates for the AUC for 1000 replicates of the numerical experiment.
+The true value of the AUC is 0.5.
+CV = Cross validation.
+"}
+load("../Code/cvsim-result.RData")
+cvres <- do.call("rbind", cvsims)
+cvlong <- do.call("rbind", lapply(1:ncol(cvres), function(i){
 
-## Evaluating Discrimination and Clinical Benefit
+  cl <- cvres[, i]
+  splow <- strsplit(colnames(cvres)[i], ".", fixed = TRUE)
+  nm <- paste(unlist(sapply(splow, function(s) rev(rev(s)[-1]))), collapse = ".")
+  cls <- sapply(splow, function(s) rev(s)[1])
+  data.frame(value = cvres[, i], stat = cls, scen = nm, stringsAsFactors = FALSE)
 
-### Developing and validating a predictive biomarker for an effective treatment
+}))
 
-Identify the non-responders and demonstrate that they do not benefit (hard)
+ggplot(subset(cvlong, stat == "AUC"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") + 
+  theme_bw(base_size = 13, base_family = "serif") + coord_flip() + 
+  scale_x_discrete("Estimation Approach", labels = c("Bootstrap", "Leave 10 out CV", "Leave 100 out CV", "Pre-validation",
+                                                     "30% Holdout", "50% Holdout", "Resubstitution", 
+                                                     "Partial CV", "Partial Holdout", "Partial Resubstitution")) + 
+  ylab("Area Under the ROC Curve")
+```
 
-### Developing and validating a predictive biomarker for an ineffective treatment
+```{r cvsims2, fig.cap = "Comparison of different approaches to estimating the odds ratio (OR) for discrimination in the setting where a dataset is used to both define and evaluate the signature.
+The violin plots show mirrored density estimates for the OR for 1000 replicates of the numerical experiment.
+The true value of the OR is 1.
+CV = Cross validation.
+"}
 
-Identify the responders and demonstrate that they do benefit (no quite as hard)
+ggplot(subset(cvlong, stat == "OR"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") + 
+  theme_bw(base_size = 13, base_family = "serif") + coord_flip() + 
+  scale_x_discrete("Estimation Approach", labels = c("Bootstrap", "Leave 2% out CV", "Leave 10% out CV", "Pre-validation",
+                                                     "30% Holdout", "50% Holdout", "Resubstitution", 
+                                                     "Partial CV (2%)", "Partial Holdout", "Partial Resubstitution")) + 
+  ylab("Odds Ratio")
+```
 
-# Conclusion
+```{r cvtab}
 
-What I said and how I said it.
+labs <- c(boot = "Bootstrap", cv.10 = "Leave 2% out CV", cv.100 = "Leave 10% out CV", 
+          cv.preval = "Pre-validation", 
+  holdout.3 = "30% Holdout", holdout.5 = "50% Holdout", naive = "Resubstitution", 
+  zhu.cv = "Partial CV (2%)", zhu.hold = "Partial Holdout", zhu.hold2 = "Partial Resubstitution")
+
+cvlong$Approach <- labs[cvlong$scen]
+cvlong$Truth <- ifelse(cvlong$stat == "AUC", .5, 1.0)
+cvlong$value[!is.finite(cvlong$value)] <- NA
+cvlong %>% group_by(stat, Approach) %>%
+  summarize(`mean` = mean(value, na.rm =TRUE), `std.dev` = sd(value, na.rm = TRUE), 
+            `Percent bias` = 100 * (mean(value - Truth[1], na.rm = TRUE))/Truth[1]) %>% 
+  kable(digits = 2, caption = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the odds ratio (OR) in the setting where a dataset is used to both define and evaluate the signature.
+The true value of the AUC is 0.5 and the true value of the OR is 1.0.
+Estimates are based on 1000 replicates of the numerical experiment.
+CV = Cross validation.
+")
+
+```
+
+Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the OR estimate is over 2 times higher than it should be, on average.
+Partial versions of the resubstitution (including partial holdout, and partial cross-validation) estimates do not ameliorate the bias very much.
+Investigators often feel as though partial holdout estimates are close to valid, as only half of the data are used to form the estimates, however here we see that these versions are still severely biased and should not be reported as valid assessments of the validity of biomarker signatures.
+
+All of the proposed remedies to resubstitution bias are successful and are unbiased, with their mean AUCs being nearly 0.5 and the mean ORs being nearly 1.
+We can compare the spread of the distributions to get a sense of the differences in precision of the estimates.
+The bootstrap approach appears to be the most precise of the unbiased estimates, followed by the cross-validation, holdout, and finally the pre-validation.
+The bootstrap, as intended, is a more efficient, smoothed version of the cross validation estimate.
+It provides the best balance between allocating data to precisely train the signature and having independent data remaining to precisely estimate the statistic $\phi$.
+
+## Data Analysis
+
+```{r examp1}
+source("../Code/03-new-signature-zhu.R")
+```
+
+```{r exampresub, cache = TRUE}
+## substitution estimate using full dataset
+obsgrp <- subset(ldat, Post.Surgical.Treatment == "OBS")
+obsgrpw <- subset(gdat, Post.Surgical.Treatment == "OBS")
+fit.all <- fit.superpc(obsgrp, cldat)
+fitted <- predict.superpc(gdat, fit.all)
+
+evalall <- eval.predict(obsgrpw, fit.all)
+
+cldat[, "riskgrp"] <- fitted$riskgrp
+cldat$DSS <- as.numeric(cldat$DSS.status == "Dead")
+#ggplot(subset(cldat, Post.Surgical.Treatment == "OBS"), 
+#       aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km() + 
+#  theme_bw(base_size = 13, base_family = "serif") + 
+#  scale_linetype_discrete("Resubstitution", labels = c("Low risk", "High risk")) + 
+#  ylab("Disease Specific Survival") + xlab("Time in years since randomization")
+
+```
+
+The signature development procedure was described in the introduction.
+There is both a feature selection step, and a multivariable estimation step.
+This results in a continuous signature which is the linear predictor of a Cox regression model.
+The signature is dichotomized by selecting the cutoff that yields the most significant log-rank statistic for comparing the resulting risk groups.
+Calibration of the signature is assessed using the concordance statistic as implemented in the \texttt{survival} package in \texttt{R} [@survival].
+To paraphrase the help file: this is defined as the probability of agreement for any two randomly chose observations, which in this case means that the observation with the shorter survival time also has the larger signature value.
+This is similar to an interpretation of the AUC for binary data.
+
+First we fit the signature using the entire observation cohort (n = 62).
+The signature was then evaluated on the same dataset.
+The survival plot shows extreme separation between the two risk groups (HR = `r round(exp(evalall[2]))`, p < 0.001), consistent with the reported JBL.10 signature, and the estimated concordance is `r round(evalall[1], 2)`.
+After correctly accounting for the selection process, our estimates of calibration and discrimination are much less impressive.
+
+The second plot shows the survival curves for the two risk groups using the pre-validated estimates of the risk score.
+We partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
+Then for each group $b$, we fit the model using $S_{-b}$ and obtained prevalidated estimates for $S_{b}$.
+The survival curves plot the survival times for comparing risk groups using the prevalidated estimates.
+The separation is much less impressive although it is clear that there is some signal there.
+
+```{r examppreval, cache = TRUE}
+set.seed(332)
+obssel <- unique(obsgrp$ID)
+# randomly reorder and select 8 groups of 6 and 2 groups of 7
+obssel <- obssel[sample(1:length(obssel), length(obssel), replace = FALSE)]
+sets <- c(rep(6, 8), 7, 7)
+scpet <- vector(mode = "list", length = 10)
+j <- 1
+for(i in 1:10){
+  scpet[[i]] <- obssel[j:(j + sets[i] - 1)]
+  j <- j + sets[i]
+}
+
+hat.f <- vector("list", 10)
+for(i in 1:10){
+
+  obsgrpb <- subset(ldat, Post.Surgical.Treatment == "OBS" & ID %in% setdiff(obssel, scpet[[i]]))
+  obsgrpwb <- subset(gdat, Post.Surgical.Treatment == "OBS" & ID %in% scpet[[i]])
+  fit.b <- fit.superpc(obsgrpb, cldat)
+  hat.f[[i]] <- cbind(ID = scpet[[i]], as.data.frame(predict.superpc(obsgrpwb, fit.b)))
+
+}
+
+preval.dat <- merge(do.call("rbind", hat.f), cldat[, - which(colnames(cldat) %in% c("riskgrp", "lps"))], by = "ID", all.y = FALSE)
+
+# ggplot(preval.dat, 
+#        aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km() + 
+#   theme_bw(base_size = 13, base_family = "serif") + 
+#   scale_linetype_discrete("Prevalidated", labels = c("Low risk", "High risk")) + 
+#   ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization")
+```
+
+```{r examppresub, fig.height = 9}
+trtgrp <- subset(gdat, Post.Surgical.Treatment == "ACT")
+
+trt.dat <- merge(data.frame(ID = trtgrp$ID, lps = predict.superpc(trtgrp, fit.all)$lps, 
+                            riskgrp = predict.superpc(trtgrp, fit.all)$riskgrp), 
+                 cldat[, - which(colnames(cldat) %in% c("riskgrp", "lps"))], by = "ID", all.y = FALSE)
+
+preval.dat2 <- rbind(preval.dat, trt.dat)
+
+cldat.trtplot <- cldat
+cldat.trtplot$Post.Surgical.Treatment <- "Combined"
+cldat.trtplot <- rbind(cldat, cldat.trtplot)
+ggplot(cldat.trtplot, 
+       aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km() + geom_kmticks() + 
+  theme_bw(base_size = 13, base_family = "serif") + 
+  scale_linetype_discrete("Partial Resubstitution", labels = c("Low risk", "High risk")) + 
+  ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
+  facet_wrap(~ Post.Surgical.Treatment, ncol = 1)
+```
+
+```{r inter1}
+cldat.trtplot$Risk.fac <- factor(cldat.trtplot$riskgrp, levels = c(FALSE, TRUE), labels = c("Low risk", "High risk"))
+ggplot(subset(cldat.trtplot, Post.Surgical.Treatment != "Combined") , 
+       aes(time = DSS.time, status = DSS, linetype = Post.Surgical.Treatment)) + geom_km()+ geom_kmticks() + 
+  theme_bw(base_size = 13, base_family = "serif") + 
+  scale_linetype_discrete("Treatment") + 
+  ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
+  facet_wrap(~ Risk.fac)
+```
+
+```{r exampresub2, fig.height = 9}
+preval.dat2.trtplot <- preval.dat2
+preval.dat2.trtplot$Post.Surgical.Treatment <- "Combined"
+preval.dat2.trtplot <- rbind(preval.dat2, preval.dat2.trtplot)
+preval.dat2.trtplot$Risk.fac <- factor(preval.dat2.trtplot$riskgrp, levels = c(FALSE, TRUE), labels = c("Low risk", "High risk"))
+
+ggplot(preval.dat2.trtplot, 
+       aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km()+ geom_kmticks() + 
+  theme_bw(base_size = 13, base_family = "serif") + 
+  scale_linetype_discrete("Prevalidated", labels = c("Low risk", "High risk")) + 
+  ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
+  facet_wrap(~ Post.Surgical.Treatment, ncol = 1)
+```
+
+```{r inter}
+ggplot(subset(preval.dat2.trtplot, Post.Surgical.Treatment != "Combined") , 
+       aes(time = DSS.time, status = DSS, linetype = Post.Surgical.Treatment)) + geom_km()+ geom_kmticks() + 
+  theme_bw(base_size = 13, base_family = "serif") + 
+  scale_linetype_discrete("Treatment") + 
+  ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
+  facet_wrap(~ Risk.fac)
+```
+
+```{r multivar, results = "asis"}
+cldat$Stage2 <- cldat$Stage %in% c("2A", "2B", "II")
+fit.train <- coxph(Surv(DSS.time, DSS) ~ riskgrp + Sex + Stage2 + age + Histology.type, 
+                   data = subset(cldat, Post.Surgical.Treatment == "OBS"))
+
+fit.train.inter <- coxph(Surv(DSS.time, DSS) ~ riskgrp * Post.Surgical.Treatment + Sex + Stage2 + age + Histology.type, 
+                   data = cldat)
+
+preval.dat2$Stage2 <- preval.dat2$Stage %in% c("2A", "2B", "II")
+fit.preval <- coxph(Surv(DSS.time, DSS) ~ riskgrp + Sex + Stage2 + age + Histology.type, 
+                    data = subset(preval.dat2, Post.Surgical.Treatment == "OBS"))
+
+fit.preval.inter <- coxph(Surv(DSS.time, DSS) ~ riskgrp * Post.Surgical.Treatment + Sex + Stage2 + age + Histology.type,
+                          data = preval.dat2)
+
+pCI <- function(x) {
+  sprintf("%.1f to %.1f", x[1], x[2])
+}
+
+myP <- function(x){
+  ifelse(x < .001, "< 0.001", paste(round(x, 3)))
+}
+
+HRmary <- data.frame(Comparison = "High Risk vs Low Risk", `Hazard Ratio` = round(exp(fit.train$coefficients[1]), 1), 
+           `95% CI` = pCI(exp(confint(fit.train)[1, ])), `Adjusted p` = myP(summary(fit.train)$coefficients[1, 5]), 
+           check.names = FALSE, stringsAsFactors = FALSE)
+
+HRmary <- rbind(HRmary, c("Trt/Risk interaction", round(exp(fit.train.inter$coefficients[8]), 1), 
+                          pCI(exp(confint(fit.train.inter)[8, ])), 
+                          myP(summary(fit.train.inter)$coefficients[8, 5])))
+
+HRmary <- rbind(HRmary, c("High Risk vs Low Risk", round(exp(fit.preval$coefficients[1]), 1), 
+                          pCI(exp(confint(fit.preval)[1, ])), 
+                          myP(summary(fit.preval)$coefficients[1, 5])),
+                c("Trt/Risk interaction", round(exp(fit.preval.inter$coefficients[8]), 1), 
+                          pCI(exp(confint(fit.preval.inter)[8, ])), 
+                          myP(summary(fit.preval.inter)$coefficients[8, 5])))
+
+rownames(HRmary) <- NULL
+
+HRmary <- cbind(Method = c("**Partial Resubstitution**", "", "**Prevalidation**", ""), HRmary)
+kable(HRmary, caption = "Hazard ratios and 95% confidence intervals from separate Cox regression models that adjust for tumor histologic subtype, stages, age, and sex.
+Rows labeled 'High risk vs low risk' show the hazard ratio for the signature-based risk group comparison.
+The rows labeled 'Trt/Risk interaction' show the hazard ratio for the interaction term of treatment by signature-based risk group.
+The partial substitution estimates are dramatically optimistically biased.
+\\label{adjhr}")
+```
+
+The second plot shows the survival curves for the two risk groups using the pre-validated estimates of the risk score.
+We partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
+Then for each group $b$, we fit the model using $S_{-b}$ and obtained prevalidated estimates for $S_{b}$.
+The survival curves plot the survival times for comparing risk groups using the prevalidated estimates.
+The separation is much less impressive although it is clear that there is some signal there.
+The concordance between the prevalidated signature and the survival times is `r round(survConcordance(Surv(DSS.time, DSS) ~ lps, data = preval.dat)$concordance, 2)`, indicating much worse calibration.
+
+We also show plots to assess the ability of the signature to be useful for treatment selection.
+These plots show the survival curves comparing treatment arms grouped in panels by the risk score.
+As described by @polley2013statistical, the idea is to determine whether the treatment is beneficial in one group and not benficial or harmful in another group, indicating that different treatment decisions would be made based on the signature.
+On one hand, the overfit signature shows dramatic differences in treatment efficacy between the low risk and high risk groups.
+In fact it appears that the treatment is harmful in the low risk group, but highly beneficial in the high risk group, a very rare finding.
+The prevalidated signature on the other hand, shows differences that are much less dramatic.
+It appears that the treatment is mildly beneficial in both groups, possibly to a higher degree in the high risk group.
+This suggests that the dramatic predictive value of the signature was merely an artifact of the overfitting process on the OBS arm, as pointed out by @simon2011re.
+
+In a multivariable Cox model we observe similar trends when comparing the prevalidated signature to the overfit signature.
+We fit two regression models.
+In the first, the aim is to assess the prognostic value of the signature by estimating the hazard ratio for the high risk versus low risk groups, adjusted for tumor histologic subtype, stage, age, and sex.
+In the second model, the idea is to assess the predictive value of the signature by estimating the treatment by signature interaction effect, adjusting for the same clinical covariates.
+The results are reported in Table \ref{adjhr}.
+For the partial resubstitution approach, we find an extreme hazard ratio of nearly 40 for the prognostic effect, and a strong and significant treatment by signature interaction.
+Using the prevalidated signature, the effect estimates are much smaller and insignificant in comparison to the standard clinical features.
+These results are unimpressive and would not be considered promising for clinical use.
+
+# Discussion
+
+All statistics, whether they assess calibration or discrimination or something else, are subject to bias due to overfitting.
+Remedies to this type of bias are well-studied in the statistical literature and here we have demonstrated how they can be implemented in a real scenario.
+Sadly, reports of strong associations with overfit biomarker signatures are all too common in the medical literature.
+The amount of bias that is possible is not known and can be difficult to decipher based on study reports.
+It is imperative that investigators and journal editors take overfitting bias seriously to ensure that signature estimates are valid.
+Unfortunately this often results in study reports that are far less optimistic than usual.
+
+The @zhu2010prognostic paper used the approach of identifying a signature using the control arm of the trial, evaluating it using the combined control and treatment arms, and then hoping that the signature would be useful for treatment selection.
+This approach has been shown to be invalid [@simon2011re] for identifying a predictive signature, and indeed we show that the resubstitution aspect of the evaluation likely led to overstatments of the size of the signature effect.
+Properly doing cross-validation for estimating the calibration of the signature and doing pre-validation for assessment of the discrimination of the signature show that the associations are much more modest.
+
+When we assess the signature on the independent treatment arm, we see that there is no significant difference between the risk groups.
+This raises the question of whether it is appropriate to use non-random splits of the dataset in order to obtain valid estimates of the calibration or discrimination.
+Instead of treatment arms, we could imagine a large multi-center study, and we could split the data into disjoint subgroups based on the center.
+Specifically, suppose we split the development dataset $S$ into $S_1$ and $S_2$ according to a discrete covariate $X$ that takes on levels 1 or 2.
+The we develop a signature $f_{S_1}$ using $S_1$ and evaluate it on $S_2$ by estimating $\hat{E}[\phi_{f_{S_1}}(S_2)]$.
+This is an estimate not of $E_\mathcal{P}[\phi_f(S)]$, but rather $E_{\mathcal{P}_2}[\phi_f(S)]$, where $\mathcal{P}_2$ is the distribution for the sub-population with $X = 2$ that $S_2$ is a sample from.
+This estimate would only be recommended if the signature is intended for use in that specific subpopulation, and if that were the case, then it doesn't make much sense to develop the signature using the subpopulation $\mathcal{P}_1$.
+If these groups differed substantially, then we would not expect the signature to perform adequately.
+The differences in performance may depend on many factors, including how $X$ is associated with the distribution of the features and the outcome.
+If a signature becomes broadly used in clinics, these center-to-center differences would be important to assess as a part of signature efficacy surveillance.
+
+# Note
+
+All analysis code and the source files for this manuscript is available from the authors' webpage.
 

### 2016-03-03 Code/03-new-signature-zhu.R

@@ -122,16 +122,6 @@ eval.predict <- function(dat0, las.fit){
 }
 
 
-## substitution estimate using full dataset
-
-fit.all <- fit.superpc(subset(ldat, Post.Surgical.Treatment == "OBS"), cldat)
-#fit.obs <- fit.superpc(ldat, cldat)
-fitted <- predict.superpc(subset(gdat, Post.Surgical.Treatment == "OBS"), fit.all)
-eval.predict(subset(gdat, Post.Surgical.Treatment == "OBS"), fit.all)
-
-
-
-## partial resubstitution
 
 
 

### 2016-05-09 Documents/paper.Rmd
redone simulations

@@
 ---
-title: "Issues in developing mulitvariable models for treatment selection"
+title: "Issues in developing multivariable models for treatment selection"
 author: "Michael C Sachs and Lisa M McShane"
@@
 \linespread{2}
+
+# Abstract
+
+Omics technologies that generate a large amount of molecular data about biospecimens have the potential to provide information about a patient's disease characteristics above and beyond standard clinical and pathological features.
+By combining the information from a large amount of molecular features into a multivariable model, called a biomarker signature, there is the opportunity to identify distinct subgroups of patients for whom treatment decisions can be personalized.
+A biomarker signature can guide the decisions to treat or not to treat and help identify the patients who are most likely to survive.
+The key challenge is to combine features from a high dimensional molecular assay to derive a signature with good clinical performance and appropriately characterize its performance.
+The inappropriate practice of using overlapping data to both build a signature and evaluate its performance can lead to severe over-optimism bias in performance estimates.
+We summarize the key statistical issues and methods for developing and validating biomarker signatures, using examples from the literature to illustrate them.
 
@@
 A multivariable biomarker can guide the decisions to treat or not to treat and help identify the patients who are most likely to survive.
-The key challenge we address in this paper is to estimate the precise combination of features from a high dimensional molecular assay suited to the clinical context.
+The key challenge we address in this paper is how to combine features from a high dimensional molecular assay to derive a signature that is fit for a specified clinical use and to provide valid estimates of performance characteristics.
 
@@
 
-@janes2011measuring, @janes2014approach,  @mcshane2013development
-
 ## Data analysis example
@@
 We will reanalyze the dataset and illustrate some remedies for avoiding bias in determining the discrimination (an additionally the calibration) of this signature.
-
-@altman2000we, @buyse2007towards, @moons2012riskI, @moons2012riskII
 
@@
 }))
+cvlong$value[cvlong$stat == "OR"] <- exp(cvlong$value[cvlong$stat == "OR"])
+
+cvlong$scen <- factor(cvlong$scen, levels = c("naive", "zhu.cv", "zhu.hold", "zhu.hold2", 
+                                              "cv.preval", "cv.10", "cv.100", "holdout.3", 
+                                              "holdout.5", "boot"), ordered = TRUE)
+
+labs <- c("Resubstitution", "Partial CV", 
+  "Partial Holdout", "Partial Resubstitution", 
+  "Pre-validation", "Leave 10 out CV", "Leave 100 out CV", 
+  "30% Holdout", "50% Holdout", "Bootstrap")
+
+names(labs) <- levels(cvlong$scen)
 
@@
   theme_bw(base_size = 13, base_family = "serif") + coord_flip() + 
-  scale_x_discrete("Estimation Approach", labels = c("Bootstrap", "Leave 10 out CV", "Leave 100 out CV", "Pre-validation",
-                                                     "30% Holdout", "50% Holdout", "Resubstitution", 
-                                                     "Partial CV", "Partial Holdout", "Partial Resubstitution")) + 
+  scale_x_discrete("Estimation Approach", labels = labs) + 
   ylab("Area Under the ROC Curve")
@@
 ```{r cvsims2, fig.cap = "Comparison of different approaches to estimating the odds ratio (OR) for discrimination in the setting where a dataset is used to both define and evaluate the signature.
-The violin plots show mirrored density estimates for the OR for 1000 replicates of the numerical experiment.
+The violin plots show mirrored density estimates for the log OR for 1000 replicates of the numerical experiment.
 The true value of the OR is 1.
@@
   theme_bw(base_size = 13, base_family = "serif") + coord_flip() + 
-  scale_x_discrete("Estimation Approach", labels = c("Bootstrap", "Leave 2% out CV", "Leave 10% out CV", "Pre-validation",
-                                                     "30% Holdout", "50% Holdout", "Resubstitution", 
-                                                     "Partial CV (2%)", "Partial Holdout", "Partial Resubstitution")) + 
-  ylab("Odds Ratio")
+  scale_x_discrete("Estimation Approach", labels = labs) + scale_y_log10(breaks = c(.5, 1, 2, 4)) +
+  ylab("Odds Ratio (log scale)")
 ```
@@
 
-labs <- c(boot = "Bootstrap", cv.10 = "Leave 2% out CV", cv.100 = "Leave 10% out CV", 
-          cv.preval = "Pre-validation", 
-  holdout.3 = "30% Holdout", holdout.5 = "50% Holdout", naive = "Resubstitution", 
-  zhu.cv = "Partial CV (2%)", zhu.hold = "Partial Holdout", zhu.hold2 = "Partial Resubstitution")
+cvlong$value[cvlong$stat == "OR"] <- log(cvlong$value[cvlong$stat == "OR"])
 
-cvlong$Approach <- labs[cvlong$scen]
-cvlong$Truth <- ifelse(cvlong$stat == "AUC", .5, 1.0)
+cvlong$Approach <- factor(labs[cvlong$scen], levels = labs, ordered = TRUE)
+cvlong$Truth <- ifelse(cvlong$stat == "AUC", .5, 0.0)
 cvlong$value[!is.finite(cvlong$value)] <- NA
@@
   summarize(`mean` = mean(value, na.rm =TRUE), `std.dev` = sd(value, na.rm = TRUE), 
-            `Percent bias` = 100 * (mean(value - Truth[1], na.rm = TRUE))/Truth[1]) %>% 
-  kable(digits = 2, caption = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the odds ratio (OR) in the setting where a dataset is used to both define and evaluate the signature.
-The true value of the AUC is 0.5 and the true value of the OR is 1.0.
+            `Bias` = mean(value - Truth[1], na.rm = TRUE)) %>%
+  kable(digits = 2, caption = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both define and evaluate the signature.
+The true value of the AUC is 0.5 and the true value of the Log OR is 0.0.
 Estimates are based on 1000 replicates of the numerical experiment.
@@
 
+## Additional Examples
+
+Examples of resubstitution estimates of performance are highly prevalent in the literature.
+@zhu2010prognostic was not unique in that sense, but rather they should be commended for their commitment to making their data, methods, and analysis code publically available, which allowed us to reproduce and reanalyze their study.
+More often, the methods are obfuscated through vague and non-standard descriptions, short methods sections, and hidden data analysis code.
+Most of the time, a resubstitution analysis can be discovered only through a very careful scrutiny of the methods and supplementary materials, if at all.
+Here we give some additional examples.
+
+@van2002gene used gene expression measurements to develop a signature and predict clinical outcome of patients with axillary lymph node-negative breast cancer (metastatic disease within 5 years versus disease-free at 5 years).
+The signature was developed using a stepwise selection and classification procedure, in which a subset of the genes was first selected using all of the subjects, then further subsets were selected using cross-validation until they arrived at a 70 gene signature.
+This is the classic form of the partial cross-validation as we've described above.
+The resubstitution estimate of the odds ratio to develop distant metastasis within 5-years was estimated to be 28, with $p < 0.0001$.
+A later study of a similar population used similar gene expression profiling approach, but with a valid split sample holdout validation approach [@wang2005gene].
+A univariate analysis of the locked-down signature evaluated on the holdout set gave a much more conservative odds ratio estimate of 5.7 for distant metastasis within 5 years.
+
+@zhang2001recursive describe a tree-based partitioning approach to use gene expression data to classify specimens as either cancer or normal tissue.
+Again, from a large number of candidate genes, a subset of three genes is identified using the entire set of 48 tissue specimens.
+Then, the cutoff values for the classifications are re-estimated using cross validation.
+This is yet again the classic form of partial cross validation that we have shown to be biased, contrary to their reports of "unbiased" estimates of 6-8% missclassification rates.
+This key detail of the analysis is only briefly described in the results section.
+
 # Discussion

### 2016-05-09 Documents/paper.Rmd
lisa edits round one

@@
 The signature is an unknown function $f(X): \mathbb{R}^p \mapsto \mathbb{R}^1$.
-The signature may be continuous, take multiple discrete values, or be dichotomous.
+The signature result may be continuous, take multiple discrete values, or be dichotomous.
 
-Let $S$ denote the development dataset, which includes $X$, an outcome $Y$, a treatment $Z$, and possibly other variables.
-$S$ is a sample of size $n$ from distribution $\mathcal{P}$ with domain $\mathcal{X}$.
+Let $S$ denote the development dataset which includes, for each of $n$ represented individuals, a feature vector $X$, an outcome $Y$, a treatment $Z$, and possibly other variables.
+$S$ is a sample of size $n$ from distribution $\mathcal{P}$ with domain $\mathcal{S}$.
 Let $\mathcal{F}$ be a mapping from $\mathcal{X}$ to the space of continuous functions with domain $\mathbb{R}^p$ and range $\mathbb{R}$, $\mathcal{D}$.
@@
 Let $\phi: \mathcal{D} \times \mathcal{X} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC).
-This could also be a measure of discrimination, such as an odds ratio, hazard ratio, or log-rank statistic.
+This could also be a measure of association, such as an odds ratio, hazard ratio, or log-rank statistic.
 This is a function of both $f$ and $S$.
-We are interested in estimating $E_\mathcal{P}[\phi_f(S)]$, which is the expected error under the data generation mechanism, for a particular $f \in \mathcal{F}$.
+We are interested in estimating $E_\mathcal{P}[\phi_{f^*}(S)]$, which is the expected error under the data generation mechanism, for a particular $f^* \in \mathcal{F}$.
 This allows us to understand how the signature will perform on future observations generated from $\mathcal{P}$.
@@
 
-The goal of the development phase of a biomarker signature is to provide a valid estimate of the performance of $f \in \mathcal{F}$.
-Optionally, one can provide a specification of $f$ for others to use, and we want that particular $f$ to be estimated as precisely as possible.
+The main goal is to produce a good signature.
+To establish this requires a valid estimate of the signature's performance.
+Assessment of the worth of a biomarker signature requires providing a valid estimate of the performance of $f \in \mathcal{F}$.
+Additionally, one can provide a specification of $f$ for others to use.
 Typically, a specific $f$ is estimated using $\mathcal{F}$ based on some training data.
@@
 Some excellent reviews of such approaches are @hastie2009elements and @moons2012riskI.
-The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply, determining how to combine the features, and whether to apply thresholds/cutoffs to the resulting signature.
+The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply, determining how to combine the feature measurements, and whether to apply thresholds/cutoffs to the resulting signature value.
+
+A biomarker signature can inform clinical practice in a number of ways.
+In early-stage disease, a highly prognostic signature may identify a subpopulation of patients that has such a good chance of long term survival, that they do not require additional treatment beyond some standard base therapy.
+Therefore these good risk patients can be spared the risks and side-effects associated with additional therapy.
+In the context of a specific therapy that targets a particular molecular pathway, a signature may identify a subpopulation of patients that does or does not benefit from that therapy, thereby guiding the decision to treat or not.
 
 Signatures can be estimated by identifying naturally occuring clusters, or intrinsic subtypes using $X$ alone without regard to the outcome $Y$.
-This was the case with the PAM-50 gene signature [reference needed], and later, the intrinsic subtypes were shown to be strongly associated with clinical outcomes.
-Supervised learning techniques can also be used to identify signatures, as was the case with Oncotype DX [@paik2004multigene].
+The PAM-50 gene signature was developed using clustering methods in an attempt to identify intrinsic subtypes with different prognosis [@parker2009supervised].
+Later, the intrinsic subtypes were shown to be strongly associated with clinical outcomes [@nielsen2010comparison].
+Supervised learning techniques can also be used to identify prognostic signatures, as was the case with Oncotype DX [@paik2004multigene], another signature developed for clinical decision making in early stage breast cancer.
 In this case, regression-based methods are used to estimate a model that is highly predictive for the outcome $Y$.
-In the case of Oncotype DX, the outcome in question was recurrence of breast cancer.
+In the case of Oncotype DX, the outcome in question was distant recurrence of breast cancer.
 
-Another approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment.
+An approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment.
 It is possible, and quite common in high-dimensional settings, to combine multiple approaches to estimating $f$.
@@
 
-A biomarker signature can inform clinical practice in a number of ways, regardless of how the signature was developed.
-A highly prognostic signature may identify a subpopulation that has such a good chance of long term survival, that they do not need to undergo treatment that carries risks and side effects such as chemotherapy.
-In the context of a specific therapy that targets a particular molecular pathway, a signature may identify a subpopulation that does not benefit from that therapy, thereby guiding the decision to treat or not.
-A signature that was developed to identify intrinsic molecular subtypes could be both prognostic and informative about the effectiveness of a therapy.
+Biomarker signatures are useful if they can correctly and reliably classify patients into distinct subgroups for which different treatment decisions would be made.
+There are two distinct but related statistical concepts involved here: calibration and discrimination.
+Signatures are often optimized to be well-calibrated, that is, accurate for predicting outcomes.
+However if the signature does not separate a population into distinct subgroups with therapeutic relevance, then it is unlikely to be informative enough to change clinical practice.
 
-The key point is that signatures are useful if they can correctly and reliably classify patients into distinct subgroups for which different treatment decisions would be made.
-There are two distinct but related statistical concepts involved here: calibration or accuracy, and discrimination.
-Signatures are often optimized to be well-calibrated, that is, highly accurate for predicting outcomes.
-However if the signature does not separate a population into distinct subgroups, then it is unlikely to be informative enough to change clinical practice.
-
-In the development process it is important to evaluate both of these statistical concepts.
-Furthermore, it is not trivial to assess each of these in a valid manner when the data are used to define the signature itself.
-We illustrate the potential for bias, and remedies, in our examples.
+In the development process it is important to evaluate both calibration and discrimination.
+It is not trivial to assess each of these in a valid manner when the data are used to define the signature itself.
+We illustrate the potential for bias through a series of examples, and we review some strategies to avoid the pitfalls that lead to bias.
 

### 2016-05-09 Documents/paper.Rmd
need to fix latex layout, and example figures

@@
     pdf_document:
+      fig_caption: yes
       pandoc_args: ["--bibliography", "bookchapter-1.bib"]
@@
 ---
-
-\linespread{2}
 
@@
 Briefly, the data of interest are from the JBR.10 trial, which was a randomized controlled trial of adjuvant vinorelbine/cisplatin (ACT) versus observation alone (OBS) in 482 participants with non small cell lung cancer (NSCLC).
-Of those 482 participants, 169 had frozen tissue collected, and of those samples, 133 (71 in ACT and 62 in OBS) had gene-expression profiling done using U133A oligonucleotide microarrays (Affymetrix, Santa Clara, CA).
+Of those 482 participants, 169 had frozen tissue collected, and of those samples, 133 (71 in ACT and 62 in OBS) had gene-expression profiling performed using U133A oligonucleotide microarrays (Affymetrix, Santa Clara, CA).
 
@@
 The annotated gene expression data and clinical information are available from the Gene Expression Omnibus (identifier: GSE14814, @edgar2002gene).
-Batch effects were removed using the ComBat function in the sva R package [@leeksva] and then the gene expression values were centered by their means and scaled by their standard deviations.
-We expand on the main features of their analysis while illustrating the key issues under discussion.
 
-The authors of @zhu2010prognostic present results that mainly focus on the discrimination ability of their estimated signature.
+@zhu2010prognostic present results that mainly focus on the discrimination ability of their estimated signature.
 They do that by demonstrating that the two risk subgroups predicted by their signature (high risk and low risk) have separation in their survival curves and that the hazard ratio for their signature is large and significant even when adjusting for other risk factors.
@@
 
-Our signature estimation approach is similar but not identical to that in @zhu2010prognostic.
+We used a similar approach to preprocessing as did @zhu2010prognostic, as we could not reproduce their workflow exactly due to outdated software.
+Batch effects were removed using the `ComBat` function in the `sva` `R` package [@leeksva] and then the gene expression values were centered by their means and scaled by their standard deviations.
+Our signature development approach is similar but not identical to that in @zhu2010prognostic.
+For purposes of illustrating the concepts we used a simplyfied approach to signature development that retains the main features of the original.
+The exact approach to signature development does not have a major impact on the main conclusions of our evaluation of the various approaches to signature performance assessment.
+
 Following the processing steps described above, we performed a gene selection step wherein we fit univariate Cox regression models with disease specific survival as the outcome and each gene as the single predictor.
@@
 However, if $S$ is used to estimate $f$ then the estimate will be biased due to overfitting, that is, $|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|$ will be large.
-This is because $\phi$ depends on $f$, and thus the statistic $\phi$ is being adaptively defined based on the observed data $S$, hence causing the overfitting.
+This bias results from the fact that $\phi$ depends on $f$, and thus the statistic $\phi$ is being adaptively defined based on the observed data $S$.
+Overfitting occurs when a model is fit to noise in the data.
+This often occurs when fitting a model that is overly complex relative to the amount of signal in the available data.
 
-In many cases during signature development, the statistic $\phi$ in question is a statistic that relates to calibration, such as classification accuracy, correlation or mean squared error.
+In many cases during signature development, the performance metric $\phi$ in question is a statistic that relates to calibration.
 While a biomarker signature may accurately predict a clinical outcome, that does not neccessarily imply that the signature is clinically useful.
-As we mentioned before, two statistical critera are necessary for determining clinical usefulness: calibration and discrimination.
-To assess discrimination, a different statistic $\phi$ may be used, such as an odds ratio, hazard ratio, or difference in survival probabilities.
-Evaluation of $\phi$ is also subject to bias due to overfitting, a fact that is commonly overlooked in the medical literature.
+To assess discrimination, a different statistic $\phi$ may be used, such as the area under the ROC curve.
+Measures of association such as the odds ratio, hazard ratio, or difference in survival probabilities are examples of performance metrics that are commonly used, although their value for biomarker signatures has been debated [@pepe2004limitations].
+Evaluation of $\phi$ is subject to bias due to overfitting regardless of what $\phi$ is used, a fact that is commonly overlooked in the medical literature.
 
 In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups as defined by the JBL.10 signature $f$.
-The risk groups were determined by the signature $f$, which was estimated using $S$ the same data that is then used to estimate the hazard ratio which is estimated to be 15.
-They go on to assess the discrimination of the JBL.10 signature in a series of independent data sets, in which they find hazard ratios around 2.
-We will reanalyze the dataset and illustrate some remedies for avoiding bias in determining the discrimination (an additionally the calibration) of this signature.
+The risk groups were determined by the signature $f$, which was estimated using $S$, the same data that was then used to produce the hazard ratio estimate of 15.
+They go on to assess the performance of the JBL.10 signature in a series of independent data sets, in which they estimate hazard ratios of approximately 2.
+Here we reanalyze the dataset and illustrate some remedies to avoid bias in estimating signature performance.
 
-## Remedies to Overfitting
+## Avoiding Overfitting
 
-A traditional remedy to this problem of overfitting is the split sample approach.
+A traditional method to avoid overfitting is the split sample approach.
 First, randomly partition $S$ into the training sample $S_t$ and the holdout sample $S_h$ with sample sizes $n_t$ and $n_h$, respectively.
 Then, $S_h$ is hidden from the analyst while $\mathcal{F}$ is applied to $S_t$ to estimate the signature function $f_t$.
-Then, with $f_t$, and therefore $\phi$ fixed, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of both $E_{\mathcal{P}}[\phi_{f_t}(S)]$ and $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
-As a side-effect, the specific form of $f_t$ that is fixed using the $S_t$ partition can be reported as the function for others to use, therefore the aforementioned estimator is also an estimate of the error for that specific $f_t$.
-The drawback of the split-sample approach is that $f$ is not estimated as precisely as it would be compared to using the entire dataset $S$ and for the same reason neither is $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
+For fixed $f_t$, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of both $E_{\mathcal{P}}[\phi_{f_t}(S)]$ and $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
+The specific form of $f_t$ that is fixed using the $S_t$ partition can be reported as the function for others to use, therefore the aforementioned estimator is also an estimate of the error for that specific $f_t$.
+The drawback of the split-sample approach is that the performance of $f_t$ is likely to be inferior to the performance of a function $f^*$ derived using the same approach applied to the entire data set.
 @dobbin2011optimally investigate how to optimally split a dataset into training and holdout partitions.
@@
 For a fixed integer $k$, which can be between 1 and $n$, we randomly select a partition of $k$ observations from $S$, denoted $S_k$.
-Then $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the the subset of $S$ that is disjoint from $S_k$.
+Then $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the subset of $S$ that is disjoint from $S_k$.
 Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an unbiased estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
@@
 Thus, we average over $K$ to get a less noisy estimate.
-This process is called "leave $k$ out" cross-validation.
-Note that for each partition that is selected, we obtain a new estimate of $f$, therefore the estimator is an estimator only for $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
-Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$.
+This process is called "leave $k$ out" or "$n/k$ fold" cross-validation.
+Note that for each partition that is selected, we obtain a new estimate of $f$, therefore we are only estimating $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$ and not the performance for a single specified signature.
+Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$ to yield $f^*$ as above.
 
@@
 Then $f_{b}$ is estimated and fixed by applying $\mathcal{F}$ to $S_{b}$.
-The statistic $\phi$ is estimated on the subset of $S$ that is disjoint from $S_b$: $S_{-b}$.
+The performance metric $\phi$ is calculated on the subset of $S$ that is disjoint from $S_b$: $S_{-b}$ to yield an estimate of $E[\phi_{f_b}(S_{-b})]$.
 This process is repeated $K$ times to yield $K$ estimates.
-These $K$ estimates are averaged to obtain the mean over the bootstrap replicates.
+These $K$ estimates of the performance metrics are averaged to obtain the mean over the bootstrap replicates.
 @efron1997improvements suggest a variation, the 0.632 estimate:
@@
 Another variation on all of these methods is the concept of pre-validation [@tibshirani2002pre].
-With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted predictor $\hat{f}(X_i)$ is estimated for $X_i \in S_{b}$ where $\hat{f}$ is estimated using $S_{-b}$.
+With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{-b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted predictor $\hat{f}(X_i)$ is estimated for $X_i \in S_{-b}$ where $\hat{f}$ is estimated using $S_{b}$.
 This process is repeated to obtain a set of pre-validated predictor estimates $\hat{f}$ which are then used to calculate $\phi$.
@@
 To illustrate the different properties of these estimates and how they deal with overfitting, we conduct a simulation study.
-Data were generated with $n$ samples, each with a binary outcome $Y$ with prevalence 0.3, and $d$ features sampled from the standard normal distribution.
+Data were generated with 1000 observations, each with a binary outcome $Y$ with prevalence 0.3, and 500 features sampled from the standard normal distribution.
 This is the null case where no features are associated with $Y$.
-The signature estimation procedure entails a feature selection step, in which each feature is regressed against $Y$ in a univariate logistic regression model.
-The 25 features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines our final signature.
+The signature development procedure entails a feature selection step, in which each feature is regressed against $Y$ in a univariate logistic regression model.
+The 25 features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines the final signature.
 
-We compare each of the methods described above along with the biased approaches of using the full sample to select the features, followed by either fitting the model on a split sample, or using fitting the full model inside the cross validation step.
-These are referred to as "partial holdout" and "partial CV" in the tables and figures.
-Note that @zhu2010prognostic used the latter approach, which explains the differences in our results.
+We compare each of the methods described above, split-sample holdout, cross-validation, bootstrap, and pre-validation, along with several commonly used but biased approaches.
+All of the approaches are described in \ref{descript}.
+Two of the biased approaches use the full sample to select the features, followed by fitting the multivariable model on the holdout subset.
+This is referred to as "parital holdout" or "partial CV" when using split-sample holdout or cross-validation as the validation step, respectively.
 We also implemented the naive resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is trained on a holdout set and then evaluated on the combined complete dataset.
-The partial resubstitution approach is prevalent in the literature, such as the @zhu2010prognostic paper, wherein the model was trained on the one arm of the trial, and then the results were presented for the combined arms.
 Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
-In our simulation, we look at two different statistics, the area under the ROC curve (AUC) to assess calibration and the odds ratio for the outcome comparing the signature groups to assess discrimination.
+In our simulation, we look at two different performance metrics, the area under the ROC curve (AUC) and the odds ratio for the outcome comparing the signature groups.
 
-```{r cvsims, fig.cap = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) in the setting where a dataset is used to both define and evaluate the signature.
+\begin{table}
+\begin{center}
+\begin{tabular}[ht]{l|p{3in}}
+Name & Description \\
+\hline
+Partial Holdout & Select features on full dataset $S$. Split data into $S_t$ and $S_h$. Build model on $S_t$ using only features pre-selected from full dataset $S$. Then test that model on $S_h$ \\
+Partial CV  & Select features on full dataset $S$.
+Fit regression model inside a cross-validation loop, where at each iteration $S_{-k}$ restricted to pre-selected features is used to build and $S_k$ is used to test.
+\\
+Naive Resubstitution & Select features on full dataset $S$ and build model on $S$ using features pre-selected from $S$. Then test that model on $S_h$. \\
+Partial Resubstitution & Split data into $S_t$ and $S_h$. Select features on $S_t$. Build model on $S_t$ using only features pre-selected from $S_t$. Then test that model on the full dataset $S$. \\
+\end{tabular}
+\end{center}
+\caption{Description of various commonly used but biased approaches to signature performance evaluation. This are used for illustration in the simulations. \label{descript} }
+\end{table}
+
+```{r cvsims, fig.cap = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) in the setting where a dataset is used to both develop the signature and evaluate its performance.
 The violin plots show mirrored density estimates for the AUC for 1000 replicates of the numerical experiment.
+In each replicate, there are 1000 observations and 500 features.
 The true value of the AUC is 0.5.
 CV = Cross validation.
-"}
+\\label{fig1}"}
 load("../Code/cvsim-result.RData")
@@
 
-```{r cvsims2, fig.cap = "Comparison of different approaches to estimating the odds ratio (OR) for discrimination in the setting where a dataset is used to both define and evaluate the signature.
+```{r cvsims2, fig.cap = "Comparison of different approaches to estimating the odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance.
 The violin plots show mirrored density estimates for the log OR for 1000 replicates of the numerical experiment.
-The true value of the OR is 1.
+In each replicate, there are 1000 observations and 500 features.
+The true value of the OR is 1.0.
 CV = Cross validation.
-"}
+\\label{fig2} "}
 
@@
 cvlong$value[!is.finite(cvlong$value)] <- NA
-cvlong %>% group_by(stat, Approach) %>%
+cvwide <- cvlong %>% group_by(stat, Approach) %>%
   summarize(`mean` = mean(value, na.rm =TRUE), `std.dev` = sd(value, na.rm = TRUE), 
-            `Bias` = mean(value - Truth[1], na.rm = TRUE)) %>%
-  kable(digits = 2, caption = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both define and evaluate the signature.
+            `Bias` = mean(value - Truth[1], na.rm = TRUE))
+
+cvwide2 <- left_join(cvwide[1:10, -1], cvwide[11:20, -1], by = c("Approach"))
+colnames(cvwide2) <- gsub(".x", " AUC", colnames(cvwide2))
+colnames(cvwide2) <- gsub(".y", " OR", colnames(cvwide2))
+
+cvwide2 %>%
+  kable(digits = 2, caption = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance.
 The true value of the AUC is 0.5 and the true value of the Log OR is 0.0.
 Estimates are based on 1000 replicates of the numerical experiment.
+In each replicate, there are 1000 observations and 500 features.
 CV = Cross validation.
-")
+", booktabs = TRUE)
 
@@
 Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the OR estimate is over 2 times higher than it should be, on average.
-Partial versions of the resubstitution (including partial holdout, and partial cross-validation) estimates do not ameliorate the bias very much.
-Investigators often feel as though partial holdout estimates are close to valid, as only half of the data are used to form the estimates, however here we see that these versions are still severely biased and should not be reported as valid assessments of the validity of biomarker signatures.
+Partial resubstitution, partial holdout, and partial cross-validation estimates do not ameliorate the bias very much.
+Investigators often feel as though partial holdout estimates are close to valid, as only half of the data are used to form the estimates, however here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.
 
-All of the proposed remedies to resubstitution bias are successful and are unbiased, with their mean AUCs being nearly 0.5 and the mean ORs being nearly 1.
+All of the proposed remedies to avoid these biases are successful and are unbiased, with their mean AUCs being nearly 0.5 and the mean ORs being nearly 1.
 We can compare the spread of the distributions to get a sense of the differences in precision of the estimates.
 The bootstrap approach appears to be the most precise of the unbiased estimates, followed by the cross-validation, holdout, and finally the pre-validation.
-The bootstrap, as intended, is a more efficient, smoothed version of the cross validation estimate.
+The bootstrap, as intended, is a more efficient, smoothed version of the cross validation estimate (Figure \ref{fig1}, \ref{fig2}).
 It provides the best balance between allocating data to precisely train the signature and having independent data remaining to precisely estimate the statistic $\phi$.

### 2016-05-10 Documents/paper.Rmd
lisas edits

@@
     pdf_document:
+      citation_package: natbib
+      keep_tex: yes
       fig_caption: yes
-      pandoc_args: ["--bibliography", "bookchapter-1.bib"]
+      pandoc_args: ["--bibliography", "paper.bib"]
 fontsize: 12pt
@@
 
+\begin{table}
+\caption{Description of various commonly used but biased approaches to signature performance evaluation. This are used for illustration in the simulations. \label{descript} }
+\begin{center}
+\begin{tabular}[!ht]{l|p{3in}}
+Name & Description \\
+\hline
+Partial Holdout & Select features on full dataset $S$. Split data into $S_t$ and $S_h$. Build model on $S_t$ using only features pre-selected from full dataset $S$. Then test that model on $S_h$ \\
+Partial CV  & Select features on full dataset $S$.
+Fit regression model inside a cross-validation loop, where at each iteration $S_{-k}$ restricted to pre-selected features is used to build and $S_k$ is used to test.
+\\
+Naive Resubstitution & Select features on full dataset $S$ and build model on $S$ using features pre-selected from $S$. Then test that model on $S_h$. \\
+Partial Resubstitution & Split data into $S_t$ and $S_h$. Select features on $S_t$. Build model on $S_t$ using only features pre-selected from $S_t$. Then test that model on the full dataset $S$. \\
+\end{tabular}
+\end{center}
+\end{table}
+
 To illustrate the different properties of these estimates and how they deal with overfitting, we conduct a simulation study.
@@
 In our simulation, we look at two different performance metrics, the area under the ROC curve (AUC) and the odds ratio for the outcome comparing the signature groups.
-
-\begin{table}
-\begin{center}
-\begin{tabular}[ht]{l|p{3in}}
-Name & Description \\
-\hline
-Partial Holdout & Select features on full dataset $S$. Split data into $S_t$ and $S_h$. Build model on $S_t$ using only features pre-selected from full dataset $S$. Then test that model on $S_h$ \\
-Partial CV  & Select features on full dataset $S$.
-Fit regression model inside a cross-validation loop, where at each iteration $S_{-k}$ restricted to pre-selected features is used to build and $S_k$ is used to test.
-\\
-Naive Resubstitution & Select features on full dataset $S$ and build model on $S$ using features pre-selected from $S$. Then test that model on $S_h$. \\
-Partial Resubstitution & Split data into $S_t$ and $S_h$. Select features on $S_t$. Build model on $S_t$ using only features pre-selected from $S_t$. Then test that model on the full dataset $S$. \\
-\end{tabular}
-\end{center}
-\caption{Description of various commonly used but biased approaches to signature performance evaluation. This are used for illustration in the simulations. \label{descript} }
-\end{table}
 
@@
 CV = Cross validation.
-", booktabs = TRUE)
+")
 
@@
 The signature is dichotomized by selecting the cutoff that yields the most significant log-rank statistic for comparing the resulting risk groups.
-Calibration of the signature is assessed using the concordance statistic as implemented in the \texttt{survival} package in \texttt{R} [@survival].
-To paraphrase the help file: this is defined as the probability of agreement for any two randomly chose observations, which in this case means that the observation with the shorter survival time also has the larger signature value.
+Discrimination of the signature is assessed using the concordance statistic as implemented in the \texttt{survival} package in \texttt{R} [@survival].
+To paraphrase the help file: this is defined as the probability of agreement for any two randomly chosen observations, which in this case means that the observation with the shorter survival time also has the larger signature value.
 This is similar to an interpretation of the AUC for binary data.
@@
 The signature was then evaluated on the same dataset.
-The survival plot shows extreme separation between the two risk groups (HR = `r round(exp(evalall[2]))`, p < 0.001), consistent with the reported JBL.10 signature, and the estimated concordance is `r round(evalall[1], 2)`.
-After correctly accounting for the selection process, our estimates of calibration and discrimination are much less impressive.
-
-The second plot shows the survival curves for the two risk groups using the pre-validated estimates of the risk score.
-We partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
-Then for each group $b$, we fit the model using $S_{-b}$ and obtained prevalidated estimates for $S_{b}$.
-The survival curves plot the survival times for comparing risk groups using the prevalidated estimates.
-The separation is much less impressive although it is clear that there is some signal there.
+The survival plot on the left side of Figure \ref{fig2} shows extreme separation between the two risk groups (HR = `r round(exp(evalall[2]))`, p < 0.001), consistent with the reported JBL.10 signature, and the estimated concordance is `r round(evalall[1], 2)`.
+After correctly accounting for the selection process, our estimates of association and discrimination are much less impressive.
 
@@
 
-```{r examppresub, fig.height = 9}
+The right plot in Figure \ref{fig2} shows the survival curves for the two risk groups using the pre-validated estimates of the risk score.
+We partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
+Then for each group $b$, we fit the model using $S_{-b}$ and obtained prevalidated estimates for $S_{b}$.
+The survival curves plot the survival times for comparing risk groups using the prevalidated estimates.
+The separation is much less impressive.
+The concordance between the prevalidated signature and the survival times is `r round(survConcordance(Surv(DSS.time, DSS) ~ lps, data = preval.dat)$concordance, 2)`, indicating much worse discrimination.
+
+```{r examppresub, fig.height = 6.5, fig.width = 8, fig.cap = "Comparison of survival by gene expression based risk signature.
+The left plot shows the resubstitution estimate, while the right plot shows the pre-validated estimate.
+\\label{fig2}"}
 trtgrp <- subset(gdat, Post.Surgical.Treatment == "ACT")
@@
 cldat.trtplot <- rbind(cldat, cldat.trtplot)
-ggplot(cldat.trtplot, 
-       aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km() + geom_kmticks() + 
-  theme_bw(base_size = 13, base_family = "serif") + 
-  scale_linetype_discrete("Partial Resubstitution", labels = c("Low risk", "High risk")) + 
-  ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
-  facet_wrap(~ Post.Surgical.Treatment, ncol = 1)
-```
 
-```{r inter1}
-cldat.trtplot$Risk.fac <- factor(cldat.trtplot$riskgrp, levels = c(FALSE, TRUE), labels = c("Low risk", "High risk"))
-ggplot(subset(cldat.trtplot, Post.Surgical.Treatment != "Combined") , 
-       aes(time = DSS.time, status = DSS, linetype = Post.Surgical.Treatment)) + geom_km()+ geom_kmticks() + 
-  theme_bw(base_size = 13, base_family = "serif") + 
-  scale_linetype_discrete("Treatment") + 
-  ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
-  facet_wrap(~ Risk.fac)
-```
+row1 <- subset(cldat.trtplot, Post.Surgical.Treatment == "OBS")
 
-```{r exampresub2, fig.height = 9}
 preval.dat2.trtplot <- preval.dat2
@@
 
-ggplot(preval.dat2.trtplot, 
-       aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km()+ geom_kmticks() + 
+row2 <- subset(preval.dat2.trtplot, Post.Surgical.Treatment == "OBS")
+row1$Set <- "Resubstitution"
+row2$Set <- "Pre-validation"
+
+rows <- bind_rows(row1, row2)
+rows$Set <- factor(rows$Set, levels = c("Resubstitution", "Pre-validation"), ordered = TRUE)
+
+ggplot(rows, aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km() + geom_kmticks() + 
   theme_bw(base_size = 13, base_family = "serif") + 
-  scale_linetype_discrete("Prevalidated", labels = c("Low risk", "High risk")) + 
+  scale_linetype_discrete("Signature Group: ", labels = c("Low risk", "High risk")) + 
   ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
-  facet_wrap(~ Post.Surgical.Treatment, ncol = 1)
+  facet_wrap(~ Set, ncol = 2) + theme(legend.position = "bottom")
 ```
 
-```{r inter}
-ggplot(subset(preval.dat2.trtplot, Post.Surgical.Treatment != "Combined") , 
-       aes(time = DSS.time, status = DSS, linetype = Post.Surgical.Treatment)) + geom_km()+ geom_kmticks() + 
+```{r inter1, fig.height = 9, fig.width = 7.5, fig.cap = "Survival curves comparing the treatment effect by gene expression signature risk group.
+The top set of plots shows the partial resubstitution based signature and the bottom row shows the pre-validated signature estimates.
+\\label{fig3}"}
+cldat.trtplot$Risk.fac <- factor(cldat.trtplot$riskgrp, levels = c(FALSE, TRUE), labels = c("Low risk", "High risk"))
+
+library(gtable)
+library(grid)
+
+inter1 <- subset(cldat.trtplot, Post.Surgical.Treatment != "Combined") 
+inter2 <- subset(preval.dat2.trtplot, Post.Surgical.Treatment != "Combined") 
+inter1$Set <- "Resubstitution"
+inter2$Set <- "Pre-validation"
+
+inters <- bind_rows(inter1, inter2)
+inters$Set <- factor(inters$Set, levels = c("Resubstitution", "Pre-validation"), ordered = TRUE)
+
+p1 <- ggplot(inters, aes(time = DSS.time, status = DSS, linetype = Post.Surgical.Treatment)) + geom_km()+ geom_kmticks() + 
   theme_bw(base_size = 13, base_family = "serif") + 
@@
   ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
-  facet_wrap(~ Risk.fac)
+  facet_grid( Set ~ Risk.fac)  + theme(strip.text.y = element_text(angle = 0))
+
+gt <- ggplotGrob(p1)
+# Get the positions of the panels in the layout: t = top, l = left, ...
+panels <-c(subset(gt$layout, name == "panel", select = t:r))
+
+# Add a row below the x-axis tick mark labels,
+# the same height as the strip
+gt <- gtable_add_rows(gt, gt$heights[3], 3)
+gt <- gtable_add_rows(gt, gt$heights[3], 6)
+
+# Get the strip grob
+stripGrob = gtable_filter(gt, "strip-right")
+
+# Insert the strip grob into the new row
+gt = gtable_add_grob(gt, stripGrob[[1]][[1]], t = 4, l = min(panels$l), r = max(panels$r))
+gt = gtable_add_grob(gt, stripGrob[[1]][[2]], t = 7, l = min(panels$l), r = max(panels$r))
+# remove the old strip
+gt = gt[, -(max(panels$r) + 1)]
+
+grid.newpage()
+grid.draw(gt)
 ```
@@
 
-The second plot shows the survival curves for the two risk groups using the pre-validated estimates of the risk score.
-We partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
-Then for each group $b$, we fit the model using $S_{-b}$ and obtained prevalidated estimates for $S_{b}$.
-The survival curves plot the survival times for comparing risk groups using the prevalidated estimates.
-The separation is much less impressive although it is clear that there is some signal there.
-The concordance between the prevalidated signature and the survival times is `r round(survConcordance(Surv(DSS.time, DSS) ~ lps, data = preval.dat)$concordance, 2)`, indicating much worse calibration.
-
-We also show plots to assess the ability of the signature to be useful for treatment selection.
+We also show plots to assess the ability of the signature to be useful for treatment selection (Figure \ref{fig3}).
 These plots show the survival curves comparing treatment arms grouped in panels by the risk score.
@@
 Using the prevalidated signature, the effect estimates are much smaller and insignificant in comparison to the standard clinical features.
-These results are unimpressive and would not be considered promising for clinical use.
+Note that the inference from the pre-validated model is not exactly correct either, because the procedure induces a correlation among the observations.
+Despite this, these results are unimpressive and would not be considered promising for clinical use.
 

### 2016-05-10 Documents/paper.tex
lisas edits

--- 2016-05-10 Documents/paper.Rmd
@@
----
-title: "Issues in developing multivariable models for treatment selection"
-author: "Michael C Sachs and Lisa M McShane"
-date: "`r format(Sys.Date(), '%B %d, %Y')`"
-output: 
-    pdf_document:
-      citation_package: natbib
-      keep_tex: yes
-      fig_caption: yes
-      pandoc_args: ["--bibliography", "paper.bib"]
-fontsize: 12pt
-geometry: margin=1in
----
+\documentclass[12pt,]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\usepackage[margin=1in]{geometry}
+\usepackage{natbib}
+\bibliographystyle{plainnat}
+\usepackage{longtable,booktabs}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={Michael C Sachs and Lisa M McShane},
+            pdftitle={Issues in developing multivariable models for treatment selection},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{0}
 
-# Abstract
+%%% Use protect on footnotes to avoid problems with footnotes in titles
+\let\rmarkdownfootnote\footnote%
+\def\footnote{\protect\rmarkdownfootnote}
 
-Omics technologies that generate a large amount of molecular data about biospecimens have the potential to provide information about a patient's disease characteristics above and beyond standard clinical and pathological features.
-By combining the information from a large amount of molecular features into a multivariable model, called a biomarker signature, there is the opportunity to identify distinct subgroups of patients for whom treatment decisions can be personalized.
-A biomarker signature can guide the decisions to treat or not to treat and help identify the patients who are most likely to survive.
-The key challenge is to combine features from a high dimensional molecular assay to derive a signature with good clinical performance and appropriately characterize its performance.
-The inappropriate practice of using overlapping data to both build a signature and evaluate its performance can lead to severe over-optimism bias in performance estimates.
-We summarize the key statistical issues and methods for developing and validating biomarker signatures, using examples from the literature to illustrate them.
+%%% Change title format to be more compact
+\usepackage{titling}
 
-# Introduction
+% Create subtitle command for use in maketitle
+\newcommand{\subtitle}[1]{
+  \posttitle{
+    \begin{center}\large#1\end{center}
+    }
+}
 
-```{r setup, include = FALSE}
-library(dplyr)
-library(ggplot2)
-library(knitr)
-library(survival)
-library(ggkm)
-library(tidyr)
+\setlength{\droptitle}{-2em}
+  \title{Issues in developing multivariable models for treatment selection}
+  \pretitle{\vspace{\droptitle}\centering\huge}
+  \posttitle{\par}
+  \author{Michael C Sachs and Lisa M McShane}
+  \preauthor{\centering\large\emph}
+  \postauthor{\par}
+  \predate{\centering\large\emph}
+  \postdate{\par}
+  \date{May 10, 2016}
 
-opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE, warning = FALSE)
-knit_hooks$set(plot = function(x, options) {
-  if ('tikz' %in% options$dev && !options$external) {
-    hook_plot_tex(x, options)
-  } else hook_plot_md(x, options)
-})
-```
+\begin{document}
 
-Omics technologies that generate a large amount of molecular data about biospecimens have the potential to provide information about a patient's disease characteristics above and beyond standard clinical and pathological features.
-By combining the information from a large amount of molecular features into a multivariable model, hereafter referred to as a _biomarker signature_, there is the opportunity to identify distinct subgroups of patients for whom treatment decisions can be personalized.
-A multivariable biomarker can guide the decisions to treat or not to treat and help identify the patients who are most likely to survive.
-The key challenge we address in this paper is how to combine features from a high dimensional molecular assay to derive a signature that is fit for a specified clinical use and to provide valid estimates of performance characteristics.
+\maketitle
 
-## Terminology and Notation
+\section{Abstract}\label{abstract}
 
-A **biomarker signature** is a transformation of multiple individual features, typically molecular characterstics measured on a multiplex assay, to a one-dimensional space.
-Specifically, let $X$ denote the set of $p$ features under consideration.
-The signature is an unknown function $f(X): \mathbb{R}^p \mapsto \mathbb{R}^1$.
-The signature result may be continuous, take multiple discrete values, or be dichotomous.
+Omics technologies that generate a large amount of molecular data about
+biospecimens have the potential to provide information about a patient's
+disease characteristics above and beyond standard clinical and
+pathological features. By combining the information from a large amount
+of molecular features into a multivariable model, called a biomarker
+signature, there is the opportunity to identify distinct subgroups of
+patients for whom treatment decisions can be personalized. A biomarker
+signature can guide the decisions to treat or not to treat and help
+identify the patients who are most likely to survive. The key challenge
+is to combine features from a high dimensional molecular assay to derive
+a signature with good clinical performance and appropriately
+characterize its performance. The inappropriate practice of using
+overlapping data to both build a signature and evaluate its performance
+can lead to severe over-optimism bias in performance estimates. We
+summarize the key statistical issues and methods for developing and
+validating biomarker signatures, using examples from the literature to
+illustrate them.
 
-Let $S$ denote the development dataset which includes, for each of $n$ represented individuals, a feature vector $X$, an outcome $Y$, a treatment $Z$, and possibly other variables.
-$S$ is a sample of size $n$ from distribution $\mathcal{P}$ with domain $\mathcal{S}$.
-Let $\mathcal{F}$ be a mapping from $\mathcal{X}$ to the space of continuous functions with domain $\mathbb{R}^p$ and range $\mathbb{R}$, $\mathcal{D}$.
-Thus $\mathcal{F}: \mathcal{X} \mapsto \mathcal{D}$ denotes the process or algorithm through which a particular $f$ is estimated.
-We do not place any other restrictions on $\mathcal{F}$, it could be a clustering approach, a regression approach, a combination of both, or something else entirely.
-We will use $\mathcal{F}$ to denote the manner in which $f$ is estimated and will write $f \in \mathcal{F}$ to denote that $f$ is estimated with the class of methods $\mathcal{F}$.
+\section{Introduction}\label{introduction}
 
-Let $\phi: \mathcal{D} \times \mathcal{X} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC).
-This could also be a measure of association, such as an odds ratio, hazard ratio, or log-rank statistic.
-This is a function of both $f$ and $S$.
-We are interested in estimating $E_\mathcal{P}[\phi_{f^*}(S)]$, which is the expected error under the data generation mechanism, for a particular $f^* \in \mathcal{F}$.
-This allows us to understand how the signature will perform on future observations generated from $\mathcal{P}$.
-We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ *for all* $f \in \mathcal{F}$, which is the generalization error for $f$ generated using mechanism $\mathcal{F}$.
-This doesn't guide outside researchers as to which specific $f$ to use, yet it is useful for development because it tells us how much signal is in the data.
-As shorthand we will write this as $E_\mathcal{P}[\phi_\mathcal{F}(S)]$
+Omics technologies that generate a large amount of molecular data about
+biospecimens have the potential to provide information about a patient's
+disease characteristics above and beyond standard clinical and
+pathological features. By combining the information from a large amount
+of molecular features into a multivariable model, hereafter referred to
+as a \emph{biomarker signature}, there is the opportunity to identify
+distinct subgroups of patients for whom treatment decisions can be
+personalized. A multivariable biomarker can guide the decisions to treat
+or not to treat and help identify the patients who are most likely to
+survive. The key challenge we address in this paper is how to combine
+features from a high dimensional molecular assay to derive a signature
+that is fit for a specified clinical use and to provide valid estimates
+of performance characteristics.
 
-A signature that **reliably predicts** an outcome $Y$ is one that has generalization error small enough for the clinical context.
-Such a signature may be useful for treatment selection, prognosis, or other type of clinical management.
+\subsection{Terminology and Notation}\label{terminology-and-notation}
 
-## Overview of biomarker signature development
+A \textbf{biomarker signature} is a transformation of multiple
+individual features, typically molecular characterstics measured on a
+multiplex assay, to a one-dimensional space. Specifically, let \(X\)
+denote the set of \(p\) features under consideration. The signature is
+an unknown function \(f(X): \mathbb{R}^p \mapsto \mathbb{R}^1\). The
+signature result may be continuous, take multiple discrete values, or be
+dichotomous.
 
-The main goal is to produce a good signature.
-To establish this requires a valid estimate of the signature's performance.
-Assessment of the worth of a biomarker signature requires providing a valid estimate of the performance of $f \in \mathcal{F}$.
-Additionally, one can provide a specification of $f$ for others to use.
-Typically, a specific $f$ is estimated using $\mathcal{F}$ based on some training data.
-This can be done using a variety of different methods.
-In recent years there have been an explosion in the literature of computational approaches to classification and prediction, and we do not intend to summarize them all here.
-Some excellent reviews of such approaches are @hastie2009elements and @moons2012riskI.
-The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply, determining how to combine the feature measurements, and whether to apply thresholds/cutoffs to the resulting signature value.
+Let \(S\) denote the development dataset which includes, for each of
+\(n\) represented individuals, a feature vector \(X\), an outcome \(Y\),
+a treatment \(Z\), and possibly other variables. \(S\) is a sample of
+size \(n\) from distribution \(\mathcal{P}\) with domain
+\(\mathcal{S}\). Let \(\mathcal{F}\) be a mapping from \(\mathcal{X}\)
+to the space of continuous functions with domain \(\mathbb{R}^p\) and
+range \(\mathbb{R}\), \(\mathcal{D}\). Thus
+\(\mathcal{F}: \mathcal{X} \mapsto \mathcal{D}\) denotes the process or
+algorithm through which a particular \(f\) is estimated. We do not place
+any other restrictions on \(\mathcal{F}\), it could be a clustering
+approach, a regression approach, a combination of both, or something
+else entirely. We will use \(\mathcal{F}\) to denote the manner in which
+\(f\) is estimated and will write \(f \in \mathcal{F}\) to denote that
+\(f\) is estimated with the class of methods \(\mathcal{F}\).
+
+Let \(\phi: \mathcal{D} \times \mathcal{X} \mapsto \mathbb{R}\) denote
+the statistic that quantifies the performance of the function \(f\),
+such as predictive accuracy, mean squared error, or area under the
+receiver operating characteristic (ROC) curve (AUC). This could also be
+a measure of association, such as an odds ratio, hazard ratio, or
+log-rank statistic. This is a function of both \(f\) and \(S\). We are
+interested in estimating \(E_\mathcal{P}[\phi_{f^*}(S)]\), which is the
+expected error under the data generation mechanism, for a particular
+\(f^* \in \mathcal{F}\). This allows us to understand how the signature
+will perform on future observations generated from \(\mathcal{P}\). We
+may also be interested in estimating \(E_\mathcal{P}[\phi_f(S)]\)
+\emph{for all} \(f \in \mathcal{F}\), which is the generalization error
+for \(f\) generated using mechanism \(\mathcal{F}\). This doesn't guide
+outside researchers as to which specific \(f\) to use, yet it is useful
+for development because it tells us how much signal is in the data. As
+shorthand we will write this as \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\)
+
+A signature that \textbf{reliably predicts} an outcome \(Y\) is one that
+has generalization error small enough for the clinical context. Such a
+signature may be useful for treatment selection, prognosis, or other
+type of clinical management.
+
+\subsection{Overview of biomarker signature
+development}\label{overview-of-biomarker-signature-development}
+
+The main goal is to produce a good signature. To establish this requires
+a valid estimate of the signature's performance. Assessment of the worth
+of a biomarker signature requires providing a valid estimate of the
+performance of \(f \in \mathcal{F}\). Additionally, one can provide a
+specification of \(f\) for others to use. Typically, a specific \(f\) is
+estimated using \(\mathcal{F}\) based on some training data. This can be
+done using a variety of different methods. In recent years there have
+been an explosion in the literature of computational approaches to
+classification and prediction, and we do not intend to summarize them
+all here. Some excellent reviews of such approaches are
+\citet{hastie2009elements} and \citet{moons2012riskI}. The main
+considerations in signature estimation are identifying the features to
+include, deciding what transformations to apply, determining how to
+combine the feature measurements, and whether to apply
+thresholds/cutoffs to the resulting signature value.
 
 A biomarker signature can inform clinical practice in a number of ways.
-In early-stage disease, a highly prognostic signature may identify a subpopulation of patients that has such a good chance of long term survival, that they do not require additional treatment beyond some standard base therapy.
-Therefore these good risk patients can be spared the risks and side-effects associated with additional therapy.
-In the context of a specific therapy that targets a particular molecular pathway, a signature may identify a subpopulation of patients that does or does not benefit from that therapy, thereby guiding the decision to treat or not.
+In early-stage disease, a highly prognostic signature may identify a
+subpopulation of patients that has such a good chance of long term
+survival, that they do not require additional treatment beyond some
+standard base therapy. Therefore these good risk patients can be spared
+the risks and side-effects associated with additional therapy. In the
+context of a specific therapy that targets a particular molecular
+pathway, a signature may identify a subpopulation of patients that does
+or does not benefit from that therapy, thereby guiding the decision to
+treat or not.
 
-Signatures can be estimated by identifying naturally occuring clusters, or intrinsic subtypes using $X$ alone without regard to the outcome $Y$.
-The PAM-50 gene signature was developed using clustering methods in an attempt to identify intrinsic subtypes with different prognosis [@parker2009supervised].
-Later, the intrinsic subtypes were shown to be strongly associated with clinical outcomes [@nielsen2010comparison].
-Supervised learning techniques can also be used to identify prognostic signatures, as was the case with Oncotype DX [@paik2004multigene], another signature developed for clinical decision making in early stage breast cancer.
-In this case, regression-based methods are used to estimate a model that is highly predictive for the outcome $Y$.
-In the case of Oncotype DX, the outcome in question was distant recurrence of breast cancer.
+Signatures can be estimated by identifying naturally occuring clusters,
+or intrinsic subtypes using \(X\) alone without regard to the outcome
+\(Y\). The PAM-50 gene signature was developed using clustering methods
+in an attempt to identify intrinsic subtypes with different prognosis
+\citep{parker2009supervised}. Later, the intrinsic subtypes were shown
+to be strongly associated with clinical outcomes
+\citep{nielsen2010comparison}. Supervised learning techniques can also
+be used to identify prognostic signatures, as was the case with Oncotype
+DX \citep{paik2004multigene}, another signature developed for clinical
+decision making in early stage breast cancer. In this case,
+regression-based methods are used to estimate a model that is highly
+predictive for the outcome \(Y\). In the case of Oncotype DX, the
+outcome in question was distant recurrence of breast cancer.
 
-An approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment.
-It is possible, and quite common in high-dimensional settings, to combine multiple approaches to estimating $f$.
-For instance, a data-reduction step by variable selection or clustering may be performed before doing regression analysis on the resulting components.
+An approach to identifying treatment-selection signatures is to use
+regression techniques to estimate a signature that has a strong
+interaction with a particular treatment. It is possible, and quite
+common in high-dimensional settings, to combine multiple approaches to
+estimating \(f\). For instance, a data-reduction step by variable
+selection or clustering may be performed before doing regression
+analysis on the resulting components.
 
-No matter what the particular model building method is, our main concern and focus of this paper is with obtaining a valid estimate of its performance, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
-This depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used.
-An optional component of the development phase is to provide a specification of $f$ for others to use on independent data or in clinical practice.
+No matter what the particular model building method is, our main concern
+and focus of this paper is with obtaining a valid estimate of its
+performance, that is, a good estimate of
+\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\). This depends on the true signal
+in the data and the specific algorithm \(\mathcal{F}\) used. An optional
+component of the development phase is to provide a specification of
+\(f\) for others to use on independent data or in clinical practice.
 
-## Biomarker signatures in clinical practice
+\subsection{Biomarker signatures in clinical
+practice}\label{biomarker-signatures-in-clinical-practice}
 
-Biomarker signatures are useful if they can correctly and reliably classify patients into distinct subgroups for which different treatment decisions would be made.
-There are two distinct but related statistical concepts involved here: calibration and discrimination.
-Signatures are often optimized to be well-calibrated, that is, accurate for predicting outcomes.
-However if the signature does not separate a population into distinct subgroups with therapeutic relevance, then it is unlikely to be informative enough to change clinical practice.
+Biomarker signatures are useful if they can correctly and reliably
+classify patients into distinct subgroups for which different treatment
+decisions would be made. There are two distinct but related statistical
+concepts involved here: calibration and discrimination. Signatures are
+often optimized to be well-calibrated, that is, accurate for predicting
+outcomes. However if the signature does not separate a population into
+distinct subgroups with therapeutic relevance, then it is unlikely to be
+informative enough to change clinical practice.
 
-In the development process it is important to evaluate both calibration and discrimination.
-It is not trivial to assess each of these in a valid manner when the data are used to define the signature itself.
-We illustrate the potential for bias through a series of examples, and we review some strategies to avoid the pitfalls that lead to bias.
+In the development process it is important to evaluate both calibration
+and discrimination. It is not trivial to assess each of these in a valid
+manner when the data are used to define the signature itself. We
+illustrate the potential for bias through a series of examples, and we
+review some strategies to avoid the pitfalls that lead to bias.
 
-## Data analysis example
+\subsection{Data analysis example}\label{data-analysis-example}
 
-Throughout this paper, we reanalyze data from @zhu2010prognostic.
-Briefly, the data of interest are from the JBR.10 trial, which was a randomized controlled trial of adjuvant vinorelbine/cisplatin (ACT) versus observation alone (OBS) in 482 participants with non small cell lung cancer (NSCLC).
-Of those 482 participants, 169 had frozen tissue collected, and of those samples, 133 (71 in ACT and 62 in OBS) had gene-expression profiling performed using U133A oligonucleotide microarrays (Affymetrix, Santa Clara, CA).
+Throughout this paper, we reanalyze data from \citet{zhu2010prognostic}.
+Briefly, the data of interest are from the JBR.10 trial, which was a
+randomized controlled trial of adjuvant vinorelbine/cisplatin (ACT)
+versus observation alone (OBS) in 482 participants with non small cell
+lung cancer (NSCLC). Of those 482 participants, 169 had frozen tissue
+collected, and of those samples, 133 (71 in ACT and 62 in OBS) had
+gene-expression profiling performed using U133A oligonucleotide
+microarrays (Affymetrix, Santa Clara, CA).
 
-The goal of the @zhu2010prognostic paper was to identify a multi-gene signature that strongly predicts prognosis, and the hypothesis was that the poor prognosis subgroup would benefit more from ACT compared to the good prognosis subgroup.
-The signature was trained to predict disease specific survival.
-The annotated gene expression data and clinical information are available from the Gene Expression Omnibus (identifier: GSE14814, @edgar2002gene).
+The goal of the \citet{zhu2010prognostic} paper was to identify a
+multi-gene signature that strongly predicts prognosis, and the
+hypothesis was that the poor prognosis subgroup would benefit more from
+ACT compared to the good prognosis subgroup. The signature was trained
+to predict disease specific survival. The annotated gene expression data
+and clinical information are available from the Gene Expression Omnibus
+(identifier: GSE14814, \citet{edgar2002gene}).
 
-@zhu2010prognostic present results that mainly focus on the discrimination ability of their estimated signature.
-They do that by demonstrating that the two risk subgroups predicted by their signature (high risk and low risk) have separation in their survival curves and that the hazard ratio for their signature is large and significant even when adjusting for other risk factors.
-They do not directly address calibration, that is, whether their signature accurately predicts survival times.
+\citet{zhu2010prognostic} present results that mainly focus on the
+discrimination ability of their estimated signature. They do that by
+demonstrating that the two risk subgroups predicted by their signature
+(high risk and low risk) have separation in their survival curves and
+that the hazard ratio for their signature is large and significant even
+when adjusting for other risk factors. They do not directly address
+calibration, that is, whether their signature accurately predicts
+survival times.
 
-We used a similar approach to preprocessing as did @zhu2010prognostic, as we could not reproduce their workflow exactly due to outdated software.
-Batch effects were removed using the `ComBat` function in the `sva` `R` package [@leeksva] and then the gene expression values were centered by their means and scaled by their standard deviations.
-Our signature development approach is similar but not identical to that in @zhu2010prognostic.
-For purposes of illustrating the concepts we used a simplyfied approach to signature development that retains the main features of the original.
-The exact approach to signature development does not have a major impact on the main conclusions of our evaluation of the various approaches to signature performance assessment.
+We used a similar approach to preprocessing as did
+\citet{zhu2010prognostic}, as we could not reproduce their workflow
+exactly due to outdated software. Batch effects were removed using the
+\texttt{ComBat} function in the \texttt{sva} \texttt{R} package
+\citep{leeksva} and then the gene expression values were centered by
+their means and scaled by their standard deviations. Our signature
+development approach is similar but not identical to that in
+\citet{zhu2010prognostic}. For purposes of illustrating the concepts we
+used a simplyfied approach to signature development that retains the
+main features of the original. The exact approach to signature
+development does not have a major impact on the main conclusions of our
+evaluation of the various approaches to signature performance
+assessment.
 
-Following the processing steps described above, we performed a gene selection step wherein we fit univariate Cox regression models with disease specific survival as the outcome and each gene as the single predictor.
-Genes with univariate p-values less than 0.005 were selected for further analysis.
-Then, each gene was weighted by its univariate Cox regression coefficient, and the resulting weighted gene expression values summed to form risk scores.
-Genes were selected in a forward selection manner, starting with the most significant genes, the gene that improved the concordance between survival times and the risk score was selected.
-If no gene improved the concordance, the process was stopped.
-The final list of selected genes were all included in a multivariable Cox regression model to fit the final risk score.
-The cutoff that yielded the smallest log-rank statistic p-value was used to dichotomize into two risk groups.
+Following the processing steps described above, we performed a gene
+selection step wherein we fit univariate Cox regression models with
+disease specific survival as the outcome and each gene as the single
+predictor. Genes with univariate p-values less than 0.005 were selected
+for further analysis. Then, each gene was weighted by its univariate Cox
+regression coefficient, and the resulting weighted gene expression
+values summed to form risk scores. Genes were selected in a forward
+selection manner, starting with the most significant genes, the gene
+that improved the concordance between survival times and the risk score
+was selected. If no gene improved the concordance, the process was
+stopped. The final list of selected genes were all included in a
+multivariable Cox regression model to fit the final risk score. The
+cutoff that yielded the smallest log-rank statistic p-value was used to
+dichotomize into two risk groups.
 
-# Issues
+\section{Issues}\label{issues}
 
-Recall that the main goal is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected value of a given statistic on future observations for $f \in \mathcal{F}$.
-This can be estimated with the in-sample empirical estimate: $\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)$ for a particular $f$.
-However, if $S$ is used to estimate $f$ then the estimate will be biased due to overfitting, that is, $|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|$ will be large.
-This bias results from the fact that $\phi$ depends on $f$, and thus the statistic $\phi$ is being adaptively defined based on the observed data $S$.
-Overfitting occurs when a model is fit to noise in the data.
-This often occurs when fitting a model that is overly complex relative to the amount of signal in the available data.
+Recall that the main goal is to estimate
+\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\), the expected value of a given
+statistic on future observations for \(f \in \mathcal{F}\). This can be
+estimated with the in-sample empirical estimate:
+\(\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)\) for a
+particular \(f\). However, if \(S\) is used to estimate \(f\) then the
+estimate will be biased due to overfitting, that is,
+\(|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|\) will be large. This
+bias results from the fact that \(\phi\) depends on \(f\), and thus the
+statistic \(\phi\) is being adaptively defined based on the observed
+data \(S\). Overfitting occurs when a model is fit to noise in the data.
+This often occurs when fitting a model that is overly complex relative
+to the amount of signal in the available data.
 
-In many cases during signature development, the performance metric $\phi$ in question is a statistic that relates to calibration.
-While a biomarker signature may accurately predict a clinical outcome, that does not neccessarily imply that the signature is clinically useful.
-To assess discrimination, a different statistic $\phi$ may be used, such as the area under the ROC curve.
-Measures of association such as the odds ratio, hazard ratio, or difference in survival probabilities are examples of performance metrics that are commonly used, although their value for biomarker signatures has been debated [@pepe2004limitations].
-Evaluation of $\phi$ is subject to bias due to overfitting regardless of what $\phi$ is used, a fact that is commonly overlooked in the medical literature.
+In many cases during signature development, the performance metric
+\(\phi\) in question is a statistic that relates to calibration. While a
+biomarker signature may accurately predict a clinical outcome, that does
+not neccessarily imply that the signature is clinically useful. To
+assess discrimination, a different statistic \(\phi\) may be used, such
+as the area under the ROC curve. Measures of association such as the
+odds ratio, hazard ratio, or difference in survival probabilities are
+examples of performance metrics that are commonly used, although their
+value for biomarker signatures has been debated
+\citep{pepe2004limitations}. Evaluation of \(\phi\) is subject to bias
+due to overfitting regardless of what \(\phi\) is used, a fact that is
+commonly overlooked in the medical literature.
 
-In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups as defined by the JBL.10 signature $f$.
-The risk groups were determined by the signature $f$, which was estimated using $S$, the same data that was then used to produce the hazard ratio estimate of 15.
-They go on to assess the performance of the JBL.10 signature in a series of independent data sets, in which they estimate hazard ratios of approximately 2.
-Here we reanalyze the dataset and illustrate some remedies to avoid bias in estimating signature performance.
+In \citet{zhu2010prognostic}, \(\phi\) was the hazard ratio comparing
+the high risk and low risk groups as defined by the JBL.10 signature
+\(f\). The risk groups were determined by the signature \(f\), which was
+estimated using \(S\), the same data that was then used to produce the
+hazard ratio estimate of 15. They go on to assess the performance of the
+JBL.10 signature in a series of independent data sets, in which they
+estimate hazard ratios of approximately 2. Here we reanalyze the dataset
+and illustrate some remedies to avoid bias in estimating signature
+performance.
 
-## Avoiding Overfitting
+\subsection{Avoiding Overfitting}\label{avoiding-overfitting}
 
 A traditional method to avoid overfitting is the split sample approach.
-First, randomly partition $S$ into the training sample $S_t$ and the holdout sample $S_h$ with sample sizes $n_t$ and $n_h$, respectively.
-Then, $S_h$ is hidden from the analyst while $\mathcal{F}$ is applied to $S_t$ to estimate the signature function $f_t$.
-For fixed $f_t$, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of both $E_{\mathcal{P}}[\phi_{f_t}(S)]$ and $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
-The specific form of $f_t$ that is fixed using the $S_t$ partition can be reported as the function for others to use, therefore the aforementioned estimator is also an estimate of the error for that specific $f_t$.
-The drawback of the split-sample approach is that the performance of $f_t$ is likely to be inferior to the performance of a function $f^*$ derived using the same approach applied to the entire data set.
-@dobbin2011optimally investigate how to optimally split a dataset into training and holdout partitions.
+First, randomly partition \(S\) into the training sample \(S_t\) and the
+holdout sample \(S_h\) with sample sizes \(n_t\) and \(n_h\),
+respectively. Then, \(S_h\) is hidden from the analyst while
+\(\mathcal{F}\) is applied to \(S_t\) to estimate the signature function
+\(f_t\). For fixed \(f_t\), \(\hat{E}[\phi_{f_t}(S_h)]\) is an unbiased
+estimator of both \(E_{\mathcal{P}}[\phi_{f_t}(S)]\) and
+\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\). The specific form of \(f_t\)
+that is fixed using the \(S_t\) partition can be reported as the
+function for others to use, therefore the aforementioned estimator is
+also an estimate of the error for that specific \(f_t\). The drawback of
+the split-sample approach is that the performance of \(f_t\) is likely
+to be inferior to the performance of a function \(f^*\) derived using
+the same approach applied to the entire data set.
+\citet{dobbin2011optimally} investigate how to optimally split a dataset
+into training and holdout partitions.
 
-Another approach to avoid overfitting is cross-validation, which is a resampling based approach.
-For a fixed integer $k$, which can be between 1 and $n$, we randomly select a partition of $k$ observations from $S$, denoted $S_k$.
-Then $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the subset of $S$ that is disjoint from $S_k$.
-Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an unbiased estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
-This process is repeated $K$ times to yield $K$ estimates.
-Each of these estimates is unbiased, but noisy, because typically $k$ is very small relative to $n$.
-Thus, we average over $K$ to get a less noisy estimate.
-This process is called "leave $k$ out" or "$n/k$ fold" cross-validation.
-Note that for each partition that is selected, we obtain a new estimate of $f$, therefore we are only estimating $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$ and not the performance for a single specified signature.
-Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$ to yield $f^*$ as above.
+Another approach to avoid overfitting is cross-validation, which is a
+resampling based approach. For a fixed integer \(k\), which can be
+between 1 and \(n\), we randomly select a partition of \(k\)
+observations from \(S\), denoted \(S_k\). Then \(f_{-k}\) is estimated
+and fixed by appling \(\mathcal{F}\) on \(S_{-k}\) which is the subset
+of \(S\) that is disjoint from \(S_k\). Then, we get an estimate
+\(\hat{E}[\phi_{f_{-k}}(S_k)]\) which is an unbiased estimate of
+\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). This process is repeated
+\(K\) times to yield \(K\) estimates. Each of these estimates is
+unbiased, but noisy, because typically \(k\) is very small relative to
+\(n\). Thus, we average over \(K\) to get a less noisy estimate. This
+process is called ``leave \(k\) out'' or ``\(n/k\) fold''
+cross-validation. Note that for each partition that is selected, we
+obtain a new estimate of \(f\), therefore we are only estimating
+\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\) and not the performance for a
+single specified signature. Typically, if a specific form for \(f\) is
+desired, it would be estimated using the entire dataset \(S\) to yield
+\(f^*\) as above.
 
-A variation on the cross-validation approach is bootstrapping.
-In that case, a sample $S_b$ of size $n$ is sampled _with replacement_ from $S$.
-Then $f_{b}$ is estimated and fixed by applying $\mathcal{F}$ to $S_{b}$.
-The performance metric $\phi$ is calculated on the subset of $S$ that is disjoint from $S_b$: $S_{-b}$ to yield an estimate of $E[\phi_{f_b}(S_{-b})]$.
-This process is repeated $K$ times to yield $K$ estimates.
-These $K$ estimates of the performance metrics are averaged to obtain the mean over the bootstrap replicates.
-@efron1997improvements suggest a variation, the 0.632 estimate:
-$$
+A variation on the cross-validation approach is bootstrapping. In that
+case, a sample \(S_b\) of size \(n\) is sampled \emph{with replacement}
+from \(S\). Then \(f_{b}\) is estimated and fixed by applying
+\(\mathcal{F}\) to \(S_{b}\). The performance metric \(\phi\) is
+calculated on the subset of \(S\) that is disjoint from \(S_b\):
+\(S_{-b}\) to yield an estimate of \(E[\phi_{f_b}(S_{-b})]\). This
+process is repeated \(K\) times to yield \(K\) estimates. These \(K\)
+estimates of the performance metrics are averaged to obtain the mean
+over the bootstrap replicates. \citet{efron1997improvements} suggest a
+variation, the 0.632 estimate: \[
 \hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
-$$
-where $\hat{E}[\phi_{f}(S)]$ is the naive estimate of $\phi_f$ using the entire dataset.
+\] where \(\hat{E}[\phi_{f}(S)]\) is the naive estimate of \(\phi_f\)
+using the entire dataset.
 
-Another variation on all of these methods is the concept of pre-validation [@tibshirani2002pre].
-With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{-b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted predictor $\hat{f}(X_i)$ is estimated for $X_i \in S_{-b}$ where $\hat{f}$ is estimated using $S_{b}$.
-This process is repeated to obtain a set of pre-validated predictor estimates $\hat{f}$ which are then used to calculate $\phi$.
-For single-step holdout, this process is equivalent to what is described above.
-For cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on the held-out dataset.
+Another variation on all of these methods is the concept of
+pre-validation \citep{tibshirani2002pre}. With pre-validation, instead
+of computing the statistic \(\phi\) for each of the held-out subsets
+(\(S_{-b}\) for the bootstrap or \(S_{k}\) for cross-validation), the
+fitted predictor \(\hat{f}(X_i)\) is estimated for \(X_i \in S_{-b}\)
+where \(\hat{f}\) is estimated using \(S_{b}\). This process is repeated
+to obtain a set of pre-validated predictor estimates \(\hat{f}\) which
+are then used to calculate \(\phi\). For single-step holdout, this
+process is equivalent to what is described above. For cross-validation
+and the bootstrap, this process avoids the problem of having too few
+cases to estimate the statistic \(\phi\) on the held-out dataset.
 
-## Simulation Study
+\subsection{Simulation Study}\label{simulation-study}
 
@@
 
-To illustrate the different properties of these estimates and how they deal with overfitting, we conduct a simulation study.
-Data were generated with 1000 observations, each with a binary outcome $Y$ with prevalence 0.3, and 500 features sampled from the standard normal distribution.
-This is the null case where no features are associated with $Y$.
-The signature development procedure entails a feature selection step, in which each feature is regressed against $Y$ in a univariate logistic regression model.
-The 25 features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines the final signature.
+To illustrate the different properties of these estimates and how they
+deal with overfitting, we conduct a simulation study. Data were
+generated with 1000 observations, each with a binary outcome \(Y\) with
+prevalence 0.3, and 500 features sampled from the standard normal
+distribution. This is the null case where no features are associated
+with \(Y\). The signature development procedure entails a feature
+selection step, in which each feature is regressed against \(Y\) in a
+univariate logistic regression model. The 25 features with the smallest
+p-values are selected for inclusion in a multivariable logistic
+regression model which defines the final signature.
 
-We compare each of the methods described above, split-sample holdout, cross-validation, bootstrap, and pre-validation, along with several commonly used but biased approaches.
-All of the approaches are described in \ref{descript}.
-Two of the biased approaches use the full sample to select the features, followed by fitting the multivariable model on the holdout subset.
-This is referred to as "parital holdout" or "partial CV" when using split-sample holdout or cross-validation as the validation step, respectively.
-We also implemented the naive resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is trained on a holdout set and then evaluated on the combined complete dataset.
-Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
-In our simulation, we look at two different performance metrics, the area under the ROC curve (AUC) and the odds ratio for the outcome comparing the signature groups.
+We compare each of the methods described above, split-sample holdout,
+cross-validation, bootstrap, and pre-validation, along with several
+commonly used but biased approaches. All of the approaches are described
+in \ref{descript}. Two of the biased approaches use the full sample to
+select the features, followed by fitting the multivariable model on the
+holdout subset. This is referred to as ``parital holdout'' or ``partial
+CV'' when using split-sample holdout or cross-validation as the
+validation step, respectively. We also implemented the naive
+resubstitution approach, wherein the model is trained and evaluated on
+the same dataset, and the partial resubstitution approach wherein the
+model is trained on a holdout set and then evaluated on the combined
+complete dataset. Our main interest is in comparing the bias and
+variance of the resulting estimates of
+\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). In our simulation, we look
+at two different performance metrics, the area under the ROC curve (AUC)
+and the odds ratio for the outcome comparing the signature groups.
 
-```{r cvsims, fig.cap = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) in the setting where a dataset is used to both develop the signature and evaluate its performance.
-The violin plots show mirrored density estimates for the AUC for 1000 replicates of the numerical experiment.
-In each replicate, there are 1000 observations and 500 features.
-The true value of the AUC is 0.5.
-CV = Cross validation.
-\\label{fig1}"}
-load("../Code/cvsim-result.RData")
-cvres <- do.call("rbind", cvsims)
-cvlong <- do.call("rbind", lapply(1:ncol(cvres), function(i){
+\begin{figure}[htbp]
+\centering
+\includegraphics{paper_files/figure-latex/cvsims-1.pdf}
+\caption{Comparison of different approaches to estimating the Area Under
+the ROC Curve (AUC) in the setting where a dataset is used to both
+develop the signature and evaluate its performance. The violin plots
+show mirrored density estimates for the AUC for 1000 replicates of the
+numerical experiment. In each replicate, there are 1000 observations and
+500 features. The true value of the AUC is 0.5. CV = Cross validation.
+\label{fig1}}
+\end{figure}
 
-  cl <- cvres[, i]
-  splow <- strsplit(colnames(cvres)[i], ".", fixed = TRUE)
-  nm <- paste(unlist(sapply(splow, function(s) rev(rev(s)[-1]))), collapse = ".")
-  cls <- sapply(splow, function(s) rev(s)[1])
-  data.frame(value = cvres[, i], stat = cls, scen = nm, stringsAsFactors = FALSE)
+\begin{figure}[htbp]
+\centering
+\includegraphics{paper_files/figure-latex/cvsims2-1.pdf}
+\caption{Comparison of different approaches to estimating the odds ratio
+(OR) in the setting where a dataset is used to both develop the
+signature and evaluate its performance. The violin plots show mirrored
+density estimates for the log OR for 1000 replicates of the numerical
+experiment. In each replicate, there are 1000 observations and 500
+features. The true value of the OR is 1.0. CV = Cross validation.
+\label{fig2}}
+\end{figure}
 
-}))
-cvlong$value[cvlong$stat == "OR"] <- exp(cvlong$value[cvlong$stat == "OR"])
+\begin{longtable}[c]{@{}lrrrrrr@{}}
+\caption{Comparison of different approaches to estimating the Area Under
+the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a
+dataset is used to both develop the signature and evaluate its
+performance. The true value of the AUC is 0.5 and the true value of the
+Log OR is 0.0. Estimates are based on 1000 replicates of the numerical
+experiment. In each replicate, there are 1000 observations and 500
+features. CV = Cross validation.}\tabularnewline
+\toprule
+Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR &
+Bias OR\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR &
+Bias OR\tabularnewline
+\midrule
+\endhead
+Resubstitution & 0.72 & 0.01 & 0.22 & 1.33 & 0.12 & 1.33\tabularnewline
+Partial CV & 0.68 & 0.01 & 0.18 & 1.02 & 0.12 & 1.02\tabularnewline
+Partial Holdout & 0.67 & 0.02 & 0.17 & 0.97 & 0.19 & 0.97\tabularnewline
+Partial Resubstitution & 0.65 & 0.02 & 0.15 & 0.88 & 0.12 &
+0.88\tabularnewline
+Pre-validation & 0.50 & 0.05 & 0.00 & -0.01 & 0.32 &
+-0.01\tabularnewline
+Leave 10 out CV & 0.50 & 0.04 & 0.00 & 0.00 & 0.25 & 0.00\tabularnewline
+Leave 100 out CV & 0.50 & 0.03 & 0.00 & 0.00 & 0.20 &
+0.00\tabularnewline
+30\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.24 & 0.00\tabularnewline
+50\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.20 & 0.00\tabularnewline
+Bootstrap & 0.50 & 0.01 & 0.00 & 0.00 & 0.08 & 0.00\tabularnewline
+\bottomrule
+\end{longtable}
 
-cvlong$scen <- factor(cvlong$scen, levels = c("naive", "zhu.cv", "zhu.hold", "zhu.hold2", 
-                                              "cv.preval", "cv.10", "cv.100", "holdout.3", 
-                                              "holdout.5", "boot"), ordered = TRUE)
+Not surprisingly, the resubstitution estimates are optimistically
+biased: the naive resubstitution estimate of the AUC is 44\% larger than
+it should be and the OR estimate is over 2 times higher than it should
+be, on average. Partial resubstitution, partial holdout, and partial
+cross-validation estimates do not ameliorate the bias very much.
+Investigators often feel as though partial holdout estimates are close
+to valid, as only half of the data are used to form the estimates,
+however here we see that these versions are still severely biased and
+should not be reported as valid assessments of the performance of
+biomarker signatures.
 
-labs <- c("Resubstitution", "Partial CV", 
-  "Partial Holdout", "Partial Resubstitution", 
-  "Pre-validation", "Leave 10 out CV", "Leave 100 out CV", 
-  "30% Holdout", "50% Holdout", "Bootstrap")
+All of the proposed remedies to avoid these biases are successful and
+are unbiased, with their mean AUCs being nearly 0.5 and the mean ORs
+being nearly 1. We can compare the spread of the distributions to get a
+sense of the differences in precision of the estimates. The bootstrap
+approach appears to be the most precise of the unbiased estimates,
+followed by the cross-validation, holdout, and finally the
+pre-validation. The bootstrap, as intended, is a more efficient,
+smoothed version of the cross validation estimate (Figure \ref{fig1},
+\ref{fig2}). It provides the best balance between allocating data to
+precisely train the signature and having independent data remaining to
+precisely estimate the statistic \(\phi\).
 
-names(labs) <- levels(cvlong$scen)
-
-ggplot(subset(cvlong, stat == "AUC"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") + 
-  theme_bw(base_size = 13, base_family = "serif") + coord_flip() + 
-  scale_x_discrete("Estimation Approach", labels = labs) + 
-  ylab("Area Under the ROC Curve")
-```
-
-```{r cvsims2, fig.cap = "Comparison of different approaches to estimating the odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance.
-The violin plots show mirrored density estimates for the log OR for 1000 replicates of the numerical experiment.
-In each replicate, there are 1000 observations and 500 features.
-The true value of the OR is 1.0.
-CV = Cross validation.
-\\label{fig2} "}
-
-ggplot(subset(cvlong, stat == "OR"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") + 
-  theme_bw(base_size = 13, base_family = "serif") + coord_flip() + 
-  scale_x_discrete("Estimation Approach", labels = labs) + scale_y_log10(breaks = c(.5, 1, 2, 4)) +
-  ylab("Odds Ratio (log scale)")
-```
-
-```{r cvtab}
-
-cvlong$value[cvlong$stat == "OR"] <- log(cvlong$value[cvlong$stat == "OR"])
-
-cvlong$Approach <- factor(labs[cvlong$scen], levels = labs, ordered = TRUE)
-cvlong$Truth <- ifelse(cvlong$stat == "AUC", .5, 0.0)
-cvlong$value[!is.finite(cvlong$value)] <- NA
-cvwide <- cvlong %>% group_by(stat, Approach) %>%
-  summarize(`mean` = mean(value, na.rm =TRUE), `std.dev` = sd(value, na.rm = TRUE), 
-            `Bias` = mean(value - Truth[1], na.rm = TRUE))
-
-cvwide2 <- left_join(cvwide[1:10, -1], cvwide[11:20, -1], by = c("Approach"))
-colnames(cvwide2) <- gsub(".x", " AUC", colnames(cvwide2))
-colnames(cvwide2) <- gsub(".y", " OR", colnames(cvwide2))
-
-cvwide2 %>%
-  kable(digits = 2, caption = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance.
-The true value of the AUC is 0.5 and the true value of the Log OR is 0.0.
-Estimates are based on 1000 replicates of the numerical experiment.
-In each replicate, there are 1000 observations and 500 features.
-CV = Cross validation.
-")
-
-```
-
-Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the OR estimate is over 2 times higher than it should be, on average.
-Partial resubstitution, partial holdout, and partial cross-validation estimates do not ameliorate the bias very much.
-Investigators often feel as though partial holdout estimates are close to valid, as only half of the data are used to form the estimates, however here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.
-
-All of the proposed remedies to avoid these biases are successful and are unbiased, with their mean AUCs being nearly 0.5 and the mean ORs being nearly 1.
-We can compare the spread of the distributions to get a sense of the differences in precision of the estimates.
-The bootstrap approach appears to be the most precise of the unbiased estimates, followed by the cross-validation, holdout, and finally the pre-validation.
-The bootstrap, as intended, is a more efficient, smoothed version of the cross validation estimate (Figure \ref{fig1}, \ref{fig2}).
-It provides the best balance between allocating data to precisely train the signature and having independent data remaining to precisely estimate the statistic $\phi$.
-
-## Data Analysis
-
-```{r examp1}
-source("../Code/03-new-signature-zhu.R")
-```
-
-```{r exampresub, cache = TRUE}
-## substitution estimate using full dataset
-obsgrp <- subset(ldat, Post.Surgical.Treatment == "OBS")
-obsgrpw <- subset(gdat, Post.Surgical.Treatment == "OBS")
-fit.all <- fit.superpc(obsgrp, cldat)
-fitted <- predict.superpc(gdat, fit.all)
-
-evalall <- eval.predict(obsgrpw, fit.all)
-
-cldat[, "riskgrp"] <- fitted$riskgrp
-cldat$DSS <- as.numeric(cldat$DSS.status == "Dead")
-#ggplot(subset(cldat, Post.Surgical.Treatment == "OBS"), 
-#       aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km() + 
-#  theme_bw(base_size = 13, base_family = "serif") + 
-#  scale_linetype_discrete("Resubstitution", labels = c("Low risk", "High risk")) + 
-#  ylab("Disease Specific Survival") + xlab("Time in years since randomization")
-
-```
+\subsection{Data Analysis}\label{data-analysis}
 
 The signature development procedure was described in the introduction.
-There is both a feature selection step, and a multivariable estimation step.
-This results in a continuous signature which is the linear predictor of a Cox regression model.
-The signature is dichotomized by selecting the cutoff that yields the most significant log-rank statistic for comparing the resulting risk groups.
-Discrimination of the signature is assessed using the concordance statistic as implemented in the \texttt{survival} package in \texttt{R} [@survival].
-To paraphrase the help file: this is defined as the probability of agreement for any two randomly chosen observations, which in this case means that the observation with the shorter survival time also has the larger signature value.
-This is similar to an interpretation of the AUC for binary data.
+There is both a feature selection step, and a multivariable estimation
+step. This results in a continuous signature which is the linear
+predictor of a Cox regression model. The signature is dichotomized by
+selecting the cutoff that yields the most significant log-rank statistic
+for comparing the resulting risk groups. Discrimination of the signature
+is assessed using the concordance statistic as implemented in the
+\texttt{survival} package in \texttt{R} \citep{survival}. To paraphrase
+the help file: this is defined as the probability of agreement for any
+two randomly chosen observations, which in this case means that the
+observation with the shorter survival time also has the larger signature
+value. This is similar to an interpretation of the AUC for binary data.
 
 First we fit the signature using the entire observation cohort (n = 62).
-The signature was then evaluated on the same dataset.
-The survival plot on the left side of Figure \ref{fig2} shows extreme separation between the two risk groups (HR = `r round(exp(evalall[2]))`, p < 0.001), consistent with the reported JBL.10 signature, and the estimated concordance is `r round(evalall[1], 2)`.
-After correctly accounting for the selection process, our estimates of association and discrimination are much less impressive.
+The signature was then evaluated on the same dataset. The survival plot
+on the left side of Figure \ref{fig2} shows extreme separation between
+the two risk groups (HR = 20, p \textless{} 0.001), consistent with the
+reported JBL.10 signature, and the estimated concordance is 0.87. After
+correctly accounting for the selection process, our estimates of
+association and discrimination are much less impressive.
 
-```{r examppreval, cache = TRUE}
-set.seed(332)
-obssel <- unique(obsgrp$ID)
-# randomly reorder and select 8 groups of 6 and 2 groups of 7
-obssel <- obssel[sample(1:length(obssel), length(obssel), replace = FALSE)]
-sets <- c(rep(6, 8), 7, 7)
-scpet <- vector(mode = "list", length = 10)
-j <- 1
-for(i in 1:10){
-  scpet[[i]] <- obssel[j:(j + sets[i] - 1)]
-  j <- j + sets[i]
-}
+The right plot in Figure \ref{fig2} shows the survival curves for the
+two risk groups using the pre-validated estimates of the risk score. We
+partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
+Then for each group \(b\), we fit the model using \(S_{-b}\) and
+obtained prevalidated estimates for \(S_{b}\). The survival curves plot
+the survival times for comparing risk groups using the prevalidated
+estimates. The separation is much less impressive. The concordance
+between the prevalidated signature and the survival times is 0.61,
+indicating much worse discrimination.
 
-hat.f <- vector("list", 10)
-for(i in 1:10){
+\begin{figure}[htbp]
+\centering
+\includegraphics{paper_files/figure-latex/examppresub-1.pdf}
+\caption{Comparison of survival by gene expression based risk signature.
+The left plot shows the resubstitution estimate, while the right plot
+shows the pre-validated estimate. \label{fig2}}
+\end{figure}
 
-  obsgrpb <- subset(ldat, Post.Surgical.Treatment == "OBS" & ID %in% setdiff(obssel, scpet[[i]]))
-  obsgrpwb <- subset(gdat, Post.Surgical.Treatment == "OBS" & ID %in% scpet[[i]])
-  fit.b <- fit.superpc(obsgrpb, cldat)
-  hat.f[[i]] <- cbind(ID = scpet[[i]], as.data.frame(predict.superpc(obsgrpwb, fit.b)))
+\begin{figure}[htbp]
+\centering
+\includegraphics{paper_files/figure-latex/inter1-1.pdf}
+\caption{Survival curves comparing the treatment effect by gene
+expression signature risk group. The top set of plots shows the partial
+resubstitution based signature and the bottom row shows the
+pre-validated signature estimates. \label{fig3}}
+\end{figure}
 
-}
+\begin{longtable}[c]{@{}lllll@{}}
+\caption{Hazard ratios and 95\% confidence intervals from separate Cox
+regression models that adjust for tumor histologic subtype, stages, age,
+and sex. Rows labeled `High risk vs low risk' show the hazard ratio for
+the signature-based risk group comparison. The rows labeled `Trt/Risk
+interaction' show the hazard ratio for the interaction term of treatment
+by signature-based risk group. The partial substitution estimates are
+dramatically optimistically biased. \label{adjhr}}\tabularnewline
+\toprule
+Method & Comparison & Hazard Ratio & 95\% CI & Adjusted p\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+Method & Comparison & Hazard Ratio & 95\% CI & Adjusted p\tabularnewline
+\midrule
+\endhead
+\textbf{Partial Resubstitution} & High Risk vs Low Risk & 38.9 & 9.2 to
+164.7 & \textless{} 0.001\tabularnewline
+& Trt/Risk interaction & 14.7 & 3.2 to 67.0 & \textless{}
+0.001\tabularnewline
+\textbf{Prevalidation} & High Risk vs Low Risk & 1.9 & 0.8 to 4.3 &
+0.122\tabularnewline
+& Trt/Risk interaction & 1.8 & 0.5 to 6.5 & 0.395\tabularnewline
+\bottomrule
+\end{longtable}
 
-preval.dat <- merge(do.call("rbind", hat.f), cldat[, - which(colnames(cldat) %in% c("riskgrp", "lps"))], by = "ID", all.y = FALSE)
+We also show plots to assess the ability of the signature to be useful
+for treatment selection (Figure \ref{fig3}). These plots show the
+survival curves comparing treatment arms grouped in panels by the risk
+score. As described by \citet{polley2013statistical}, the idea is to
+determine whether the treatment is beneficial in one group and not
+benficial or harmful in another group, indicating that different
+treatment decisions would be made based on the signature. On one hand,
+the overfit signature shows dramatic differences in treatment efficacy
+between the low risk and high risk groups. In fact it appears that the
+treatment is harmful in the low risk group, but highly beneficial in the
+high risk group, a very rare finding. The prevalidated signature on the
+other hand, shows differences that are much less dramatic. It appears
+that the treatment is mildly beneficial in both groups, possibly to a
+higher degree in the high risk group. This suggests that the dramatic
+predictive value of the signature was merely an artifact of the
+overfitting process on the OBS arm, as pointed out by
+\citet{simon2011re}.
 
-# ggplot(preval.dat, 
-#        aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km() + 
-#   theme_bw(base_size = 13, base_family = "serif") + 
-#   scale_linetype_discrete("Prevalidated", labels = c("Low risk", "High risk")) + 
-#   ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization")
-```
+In a multivariable Cox model we observe similar trends when comparing
+the prevalidated signature to the overfit signature. We fit two
+regression models. In the first, the aim is to assess the prognostic
+value of the signature by estimating the hazard ratio for the high risk
+versus low risk groups, adjusted for tumor histologic subtype, stage,
+age, and sex. In the second model, the idea is to assess the predictive
+value of the signature by estimating the treatment by signature
+interaction effect, adjusting for the same clinical covariates. The
+results are reported in Table \ref{adjhr}. For the partial
+resubstitution approach, we find an extreme hazard ratio of nearly 40
+for the prognostic effect, and a strong and significant treatment by
+signature interaction. Using the prevalidated signature, the effect
+estimates are much smaller and insignificant in comparison to the
+standard clinical features. Note that the inference from the
+pre-validated model is not exactly correct either, because the procedure
+induces a correlation among the observations. Despite this, these
+results are unimpressive and would not be considered promising for
+clinical use.
 
-The right plot in Figure \ref{fig2} shows the survival curves for the two risk groups using the pre-validated estimates of the risk score.
-We partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
-Then for each group $b$, we fit the model using $S_{-b}$ and obtained prevalidated estimates for $S_{b}$.
-The survival curves plot the survival times for comparing risk groups using the prevalidated estimates.
-The separation is much less impressive.
-The concordance between the prevalidated signature and the survival times is `r round(survConcordance(Surv(DSS.time, DSS) ~ lps, data = preval.dat)$concordance, 2)`, indicating much worse discrimination.
+\subsection{Additional Examples}\label{additional-examples}
 
-```{r examppresub, fig.height = 6.5, fig.width = 8, fig.cap = "Comparison of survival by gene expression based risk signature.
-The left plot shows the resubstitution estimate, while the right plot shows the pre-validated estimate.
-\\label{fig2}"}
-trtgrp <- subset(gdat, Post.Surgical.Treatment == "ACT")
+Examples of resubstitution estimates of performance are highly prevalent
+in the literature. \citet{zhu2010prognostic} was not unique in that
+sense, but rather they should be commended for their commitment to
+making their data, methods, and analysis code publically available,
+which allowed us to reproduce and reanalyze their study. More often, the
+methods are obfuscated through vague and non-standard descriptions,
+short methods sections, and hidden data analysis code. Most of the time,
+a resubstitution analysis can be discovered only through a very careful
+scrutiny of the methods and supplementary materials, if at all. Here we
+give some additional examples.
 
-trt.dat <- merge(data.frame(ID = trtgrp$ID, lps = predict.superpc(trtgrp, fit.all)$lps, 
-                            riskgrp = predict.superpc(trtgrp, fit.all)$riskgrp), 
-                 cldat[, - which(colnames(cldat) %in% c("riskgrp", "lps"))], by = "ID", all.y = FALSE)
+\citet{van2002gene} used gene expression measurements to develop a
+signature and predict clinical outcome of patients with axillary lymph
+node-negative breast cancer (metastatic disease within 5 years versus
+disease-free at 5 years). The signature was developed using a stepwise
+selection and classification procedure, in which a subset of the genes
+was first selected using all of the subjects, then further subsets were
+selected using cross-validation until they arrived at a 70 gene
+signature. This is the classic form of the partial cross-validation as
+we've described above. The resubstitution estimate of the odds ratio to
+develop distant metastasis within 5-years was estimated to be 28, with
+\(p < 0.0001\). A later study of a similar population used similar gene
+expression profiling approach, but with a valid split sample holdout
+validation approach \citep{wang2005gene}. A univariate analysis of the
+locked-down signature evaluated on the holdout set gave a much more
+conservative odds ratio estimate of 5.7 for distant metastasis within 5
+years.
 
-preval.dat2 <- rbind(preval.dat, trt.dat)
+\citet{zhang2001recursive} describe a tree-based partitioning approach
+to use gene expression data to classify specimens as either cancer or
+normal tissue. Again, from a large number of candidate genes, a subset
+of three genes is identified using the entire set of 48 tissue
+specimens. Then, the cutoff values for the classifications are
+re-estimated using cross validation. This is yet again the classic form
+of partial cross validation that we have shown to be biased, contrary to
+their reports of ``unbiased'' estimates of 6-8\% missclassification
+rates. This key detail of the analysis is only briefly described in the
+results section.
 
-cldat.trtplot <- cldat
-cldat.trtplot$Post.Surgical.Treatment <- "Combined"
-cldat.trtplot <- rbind(cldat, cldat.trtplot)
+\section{Discussion}\label{discussion}
 
-row1 <- subset(cldat.trtplot, Post.Surgical.Treatment == "OBS")
+All statistics, whether they assess calibration or discrimination or
+something else, are subject to bias due to overfitting. Remedies to this
+type of bias are well-studied in the statistical literature and here we
+have demonstrated how they can be implemented in a real scenario. Sadly,
+reports of strong associations with overfit biomarker signatures are all
+too common in the medical literature. The amount of bias that is
+possible is not known and can be difficult to decipher based on study
+reports. It is imperative that investigators and journal editors take
+overfitting bias seriously to ensure that signature estimates are valid.
+Unfortunately this often results in study reports that are far less
+optimistic than usual.
 
-preval.dat2.trtplot <- preval.dat2
-preval.dat2.trtplot$Post.Surgical.Treatment <- "Combined"
-preval.dat2.trtplot <- rbind(preval.dat2, preval.dat2.trtplot)
-preval.dat2.trtplot$Risk.fac <- factor(preval.dat2.trtplot$riskgrp, levels = c(FALSE, TRUE), labels = c("Low risk", "High risk"))
+The \citet{zhu2010prognostic} paper used the approach of identifying a
+signature using the control arm of the trial, evaluating it using the
+combined control and treatment arms, and then hoping that the signature
+would be useful for treatment selection. This approach has been shown to
+be invalid \citep{simon2011re} for identifying a predictive signature,
+and indeed we show that the resubstitution aspect of the evaluation
+likely led to overstatments of the size of the signature effect.
+Properly doing cross-validation for estimating the calibration of the
+signature and doing pre-validation for assessment of the discrimination
+of the signature show that the associations are much more modest.
 
-row2 <- subset(preval.dat2.trtplot, Post.Surgical.Treatment == "OBS")
-row1$Set <- "Resubstitution"
-row2$Set <- "Pre-validation"
+When we assess the signature on the independent treatment arm, we see
+that there is no significant difference between the risk groups. This
+raises the question of whether it is appropriate to use non-random
+splits of the dataset in order to obtain valid estimates of the
+calibration or discrimination. Instead of treatment arms, we could
+imagine a large multi-center study, and we could split the data into
+disjoint subgroups based on the center. Specifically, suppose we split
+the development dataset \(S\) into \(S_1\) and \(S_2\) according to a
+discrete covariate \(X\) that takes on levels 1 or 2. The we develop a
+signature \(f_{S_1}\) using \(S_1\) and evaluate it on \(S_2\) by
+estimating \(\hat{E}[\phi_{f_{S_1}}(S_2)]\). This is an estimate not of
+\(E_\mathcal{P}[\phi_f(S)]\), but rather
+\(E_{\mathcal{P}_2}[\phi_f(S)]\), where \(\mathcal{P}_2\) is the
+distribution for the sub-population with \(X = 2\) that \(S_2\) is a
+sample from. This estimate would only be recommended if the signature is
+intended for use in that specific subpopulation, and if that were the
+case, then it doesn't make much sense to develop the signature using the
+subpopulation \(\mathcal{P}_1\). If these groups differed substantially,
+then we would not expect the signature to perform adequately. The
+differences in performance may depend on many factors, including how
+\(X\) is associated with the distribution of the features and the
+outcome. If a signature becomes broadly used in clinics, these
+center-to-center differences would be important to assess as a part of
+signature efficacy surveillance.
 
-rows <- bind_rows(row1, row2)
-rows$Set <- factor(rows$Set, levels = c("Resubstitution", "Pre-validation"), ordered = TRUE)
+\section{Note}\label{note}
 
-ggplot(rows, aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km() + geom_kmticks() + 
-  theme_bw(base_size = 13, base_family = "serif") + 
-  scale_linetype_discrete("Signature Group: ", labels = c("Low risk", "High risk")) + 
-  ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
-  facet_wrap(~ Set, ncol = 2) + theme(legend.position = "bottom")
-```
+All analysis code and the source files for this manuscript is available
+from the authors' webpage.
 
-```{r inter1, fig.height = 9, fig.width = 7.5, fig.cap = "Survival curves comparing the treatment effect by gene expression signature risk group.
-The top set of plots shows the partial resubstitution based signature and the bottom row shows the pre-validated signature estimates.
-\\label{fig3}"}
-cldat.trtplot$Risk.fac <- factor(cldat.trtplot$riskgrp, levels = c(FALSE, TRUE), labels = c("Low risk", "High risk"))
+\renewcommand\refname{References}
+\bibliography{paper}
 
-library(gtable)
-library(grid)
-
-inter1 <- subset(cldat.trtplot, Post.Surgical.Treatment != "Combined") 
-inter2 <- subset(preval.dat2.trtplot, Post.Surgical.Treatment != "Combined") 
-inter1$Set <- "Resubstitution"
-inter2$Set <- "Pre-validation"
-
-inters <- bind_rows(inter1, inter2)
-inters$Set <- factor(inters$Set, levels = c("Resubstitution", "Pre-validation"), ordered = TRUE)
-
-p1 <- ggplot(inters, aes(time = DSS.time, status = DSS, linetype = Post.Surgical.Treatment)) + geom_km()+ geom_kmticks() + 
-  theme_bw(base_size = 13, base_family = "serif") + 
-  scale_linetype_discrete("Treatment") + 
-  ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
-  facet_grid( Set ~ Risk.fac)  + theme(strip.text.y = element_text(angle = 0))
-
-gt <- ggplotGrob(p1)
-# Get the positions of the panels in the layout: t = top, l = left, ...
-panels <-c(subset(gt$layout, name == "panel", select = t:r))
-
-# Add a row below the x-axis tick mark labels,
-# the same height as the strip
-gt <- gtable_add_rows(gt, gt$heights[3], 3)
-gt <- gtable_add_rows(gt, gt$heights[3], 6)
-
-# Get the strip grob
-stripGrob = gtable_filter(gt, "strip-right")
-
-# Insert the strip grob into the new row
-gt = gtable_add_grob(gt, stripGrob[[1]][[1]], t = 4, l = min(panels$l), r = max(panels$r))
-gt = gtable_add_grob(gt, stripGrob[[1]][[2]], t = 7, l = min(panels$l), r = max(panels$r))
-# remove the old strip
-gt = gt[, -(max(panels$r) + 1)]
-
-grid.newpage()
-grid.draw(gt)
-```
-
-```{r multivar, results = "asis"}
-cldat$Stage2 <- cldat$Stage %in% c("2A", "2B", "II")
-fit.train <- coxph(Surv(DSS.time, DSS) ~ riskgrp + Sex + Stage2 + age + Histology.type, 
-                   data = subset(cldat, Post.Surgical.Treatment == "OBS"))
-
-fit.train.inter <- coxph(Surv(DSS.time, DSS) ~ riskgrp * Post.Surgical.Treatment + Sex + Stage2 + age + Histology.type, 
-                   data = cldat)
-
-preval.dat2$Stage2 <- preval.dat2$Stage %in% c("2A", "2B", "II")
-fit.preval <- coxph(Surv(DSS.time, DSS) ~ riskgrp + Sex + Stage2 + age + Histology.type, 
-                    data = subset(preval.dat2, Post.Surgical.Treatment == "OBS"))
-
-fit.preval.inter <- coxph(Surv(DSS.time, DSS) ~ riskgrp * Post.Surgical.Treatment + Sex + Stage2 + age + Histology.type,
-                          data = preval.dat2)
-
-pCI <- function(x) {
-  sprintf("%.1f to %.1f", x[1], x[2])
-}
-
-myP <- function(x){
-  ifelse(x < .001, "< 0.001", paste(round(x, 3)))
-}
-
-HRmary <- data.frame(Comparison = "High Risk vs Low Risk", `Hazard Ratio` = round(exp(fit.train$coefficients[1]), 1), 
-           `95% CI` = pCI(exp(confint(fit.train)[1, ])), `Adjusted p` = myP(summary(fit.train)$coefficients[1, 5]), 
-           check.names = FALSE, stringsAsFactors = FALSE)
-
-HRmary <- rbind(HRmary, c("Trt/Risk interaction", round(exp(fit.train.inter$coefficients[8]), 1), 
-                          pCI(exp(confint(fit.train.inter)[8, ])), 
-                          myP(summary(fit.train.inter)$coefficients[8, 5])))
-
-HRmary <- rbind(HRmary, c("High Risk vs Low Risk", round(exp(fit.preval$coefficients[1]), 1), 
-                          pCI(exp(confint(fit.preval)[1, ])), 
-                          myP(summary(fit.preval)$coefficients[1, 5])),
-                c("Trt/Risk interaction", round(exp(fit.preval.inter$coefficients[8]), 1), 
-                          pCI(exp(confint(fit.preval.inter)[8, ])), 
-                          myP(summary(fit.preval.inter)$coefficients[8, 5])))
-
-rownames(HRmary) <- NULL
-
-HRmary <- cbind(Method = c("**Partial Resubstitution**", "", "**Prevalidation**", ""), HRmary)
-kable(HRmary, caption = "Hazard ratios and 95% confidence intervals from separate Cox regression models that adjust for tumor histologic subtype, stages, age, and sex.
-Rows labeled 'High risk vs low risk' show the hazard ratio for the signature-based risk group comparison.
-The rows labeled 'Trt/Risk interaction' show the hazard ratio for the interaction term of treatment by signature-based risk group.
-The partial substitution estimates are dramatically optimistically biased.
-\\label{adjhr}")
-```
-
-We also show plots to assess the ability of the signature to be useful for treatment selection (Figure \ref{fig3}).
-These plots show the survival curves comparing treatment arms grouped in panels by the risk score.
-As described by @polley2013statistical, the idea is to determine whether the treatment is beneficial in one group and not benficial or harmful in another group, indicating that different treatment decisions would be made based on the signature.
-On one hand, the overfit signature shows dramatic differences in treatment efficacy between the low risk and high risk groups.
-In fact it appears that the treatment is harmful in the low risk group, but highly beneficial in the high risk group, a very rare finding.
-The prevalidated signature on the other hand, shows differences that are much less dramatic.
-It appears that the treatment is mildly beneficial in both groups, possibly to a higher degree in the high risk group.
-This suggests that the dramatic predictive value of the signature was merely an artifact of the overfitting process on the OBS arm, as pointed out by @simon2011re.
-
-In a multivariable Cox model we observe similar trends when comparing the prevalidated signature to the overfit signature.
-We fit two regression models.
-In the first, the aim is to assess the prognostic value of the signature by estimating the hazard ratio for the high risk versus low risk groups, adjusted for tumor histologic subtype, stage, age, and sex.
-In the second model, the idea is to assess the predictive value of the signature by estimating the treatment by signature interaction effect, adjusting for the same clinical covariates.
-The results are reported in Table \ref{adjhr}.
-For the partial resubstitution approach, we find an extreme hazard ratio of nearly 40 for the prognostic effect, and a strong and significant treatment by signature interaction.
-Using the prevalidated signature, the effect estimates are much smaller and insignificant in comparison to the standard clinical features.
-Note that the inference from the pre-validated model is not exactly correct either, because the procedure induces a correlation among the observations.
-Despite this, these results are unimpressive and would not be considered promising for clinical use.
-
-## Additional Examples
-
-Examples of resubstitution estimates of performance are highly prevalent in the literature.
-@zhu2010prognostic was not unique in that sense, but rather they should be commended for their commitment to making their data, methods, and analysis code publically available, which allowed us to reproduce and reanalyze their study.
-More often, the methods are obfuscated through vague and non-standard descriptions, short methods sections, and hidden data analysis code.
-Most of the time, a resubstitution analysis can be discovered only through a very careful scrutiny of the methods and supplementary materials, if at all.
-Here we give some additional examples.
-
-@van2002gene used gene expression measurements to develop a signature and predict clinical outcome of patients with axillary lymph node-negative breast cancer (metastatic disease within 5 years versus disease-free at 5 years).
-The signature was developed using a stepwise selection and classification procedure, in which a subset of the genes was first selected using all of the subjects, then further subsets were selected using cross-validation until they arrived at a 70 gene signature.
-This is the classic form of the partial cross-validation as we've described above.
-The resubstitution estimate of the odds ratio to develop distant metastasis within 5-years was estimated to be 28, with $p < 0.0001$.
-A later study of a similar population used similar gene expression profiling approach, but with a valid split sample holdout validation approach [@wang2005gene].
-A univariate analysis of the locked-down signature evaluated on the holdout set gave a much more conservative odds ratio estimate of 5.7 for distant metastasis within 5 years.
-
-@zhang2001recursive describe a tree-based partitioning approach to use gene expression data to classify specimens as either cancer or normal tissue.
-Again, from a large number of candidate genes, a subset of three genes is identified using the entire set of 48 tissue specimens.
-Then, the cutoff values for the classifications are re-estimated using cross validation.
-This is yet again the classic form of partial cross validation that we have shown to be biased, contrary to their reports of "unbiased" estimates of 6-8% missclassification rates.
-This key detail of the analysis is only briefly described in the results section.
-
-# Discussion
-
-All statistics, whether they assess calibration or discrimination or something else, are subject to bias due to overfitting.
-Remedies to this type of bias are well-studied in the statistical literature and here we have demonstrated how they can be implemented in a real scenario.
-Sadly, reports of strong associations with overfit biomarker signatures are all too common in the medical literature.
-The amount of bias that is possible is not known and can be difficult to decipher based on study reports.
-It is imperative that investigators and journal editors take overfitting bias seriously to ensure that signature estimates are valid.
-Unfortunately this often results in study reports that are far less optimistic than usual.
-
-The @zhu2010prognostic paper used the approach of identifying a signature using the control arm of the trial, evaluating it using the combined control and treatment arms, and then hoping that the signature would be useful for treatment selection.
-This approach has been shown to be invalid [@simon2011re] for identifying a predictive signature, and indeed we show that the resubstitution aspect of the evaluation likely led to overstatments of the size of the signature effect.
-Properly doing cross-validation for estimating the calibration of the signature and doing pre-validation for assessment of the discrimination of the signature show that the associations are much more modest.
-
-When we assess the signature on the independent treatment arm, we see that there is no significant difference between the risk groups.
-This raises the question of whether it is appropriate to use non-random splits of the dataset in order to obtain valid estimates of the calibration or discrimination.
-Instead of treatment arms, we could imagine a large multi-center study, and we could split the data into disjoint subgroups based on the center.
-Specifically, suppose we split the development dataset $S$ into $S_1$ and $S_2$ according to a discrete covariate $X$ that takes on levels 1 or 2.
-The we develop a signature $f_{S_1}$ using $S_1$ and evaluate it on $S_2$ by estimating $\hat{E}[\phi_{f_{S_1}}(S_2)]$.
-This is an estimate not of $E_\mathcal{P}[\phi_f(S)]$, but rather $E_{\mathcal{P}_2}[\phi_f(S)]$, where $\mathcal{P}_2$ is the distribution for the sub-population with $X = 2$ that $S_2$ is a sample from.
-This estimate would only be recommended if the signature is intended for use in that specific subpopulation, and if that were the case, then it doesn't make much sense to develop the signature using the subpopulation $\mathcal{P}_1$.
-If these groups differed substantially, then we would not expect the signature to perform adequately.
-The differences in performance may depend on many factors, including how $X$ is associated with the distribution of the features and the outcome.
-If a signature becomes broadly used in clinics, these center-to-center differences would be important to assess as a part of signature efficacy surveillance.
-
-# Note
-
-All analysis code and the source files for this manuscript is available from the authors' webpage.
-
-# References
+\end{document}

### 2016-05-18 Documents/paper revised 5-10-16+LM.Rmd
incorporated lisas edits, waiting on final discussion revisions

--- 2016-05-10 Documents/paper.Rmd
@@
 ---
-title: "Issues in developing multivariable models for treatment selection"
+title: "Issues in developing multivariable molecular signatures for guiding clinical care decisions"
 author: "Michael C Sachs and Lisa M McShane"
@@
 
-Omics technologies that generate a large amount of molecular data about biospecimens have the potential to provide information about a patient's disease characteristics above and beyond standard clinical and pathological features.
-By combining the information from a large amount of molecular features into a multivariable model, called a biomarker signature, there is the opportunity to identify distinct subgroups of patients for whom treatment decisions can be personalized.
-A biomarker signature can guide the decisions to treat or not to treat and help identify the patients who are most likely to survive.
+Omics technologies that generate a large amount of molecular data characterizing patient biospecimens have the potential to provide information about patients' disease characteristics above and beyond standard clinical and pathological features.
+By combining the information from a large number of molecular features into a multivariable model or decision algorithm, called a biomarker signature, there is the opportunity to identify distinct subgroups of patients for whom treatment decisions can be personalized.
+A biomarker signature can guide decisions to treat or not to treat and help identify the patients who are most likely to have a more favororable outcome or benefit from a particular therapy.
 The key challenge is to combine features from a high dimensional molecular assay to derive a signature with good clinical performance and appropriately characterize its performance.
 The inappropriate practice of using overlapping data to both build a signature and evaluate its performance can lead to severe over-optimism bias in performance estimates.
-We summarize the key statistical issues and methods for developing and validating biomarker signatures, using examples from the literature to illustrate them.
+We summarize the key statistical issues and methods for developing and validating biomarker signatures, using examples from the literature for illustration.
 
@@
 
-Omics technologies that generate a large amount of molecular data about biospecimens have the potential to provide information about a patient's disease characteristics above and beyond standard clinical and pathological features.
-By combining the information from a large amount of molecular features into a multivariable model, hereafter referred to as a _biomarker signature_, there is the opportunity to identify distinct subgroups of patients for whom treatment decisions can be personalized.
-A multivariable biomarker can guide the decisions to treat or not to treat and help identify the patients who are most likely to survive.
-The key challenge we address in this paper is how to combine features from a high dimensional molecular assay to derive a signature that is fit for a specified clinical use and to provide valid estimates of performance characteristics.
+Omics technologies that generate a large amount of molecular data characterizing patient biospecimens have the potential to provide information about the patients' disease characteristics above and beyond standard clinical and pathological features.
+By combining the information from a large number of molecular features into a multivariable model, hereafter referred to as a _biomarker signature_, there is the opportunity to identify distinct subgroups of patients for whom clinical management decisions can be personalized.
+In early-stage disease, for example, a highly prognostic signature might identify a subpopulation of good risk patients that has such a high probability of long survival or disease-free survival that they do not require additional treatment beyond some standard base therapy.
+Therefore these good risk patients can be spared the risks and side-effects associated with additional therapy.
+In the context of a specific therapy that targets a particular molecular pathway, a signature may identify a subpopulation of patients that does or does not benefit from that therapy, thereby guiding the decision of whether or not to administer the therapy.
+
+The key challenge we address in this paper is how to combine feature mneasurements generated by a high dimensional multiplex molecular assay to derive a signature that is fit for a specified clinical and additionally provide valid estimates of its performance characteristics.
+A signature that has clinical value must have statistical performance (as quantified by one or more metrics) that is fit for the clinical context, and it must provide information that can be acted upon clinically to provide some benefit to a patient.
+For example, a biomarker signature that is well calibrated may accurately predict a clinical outcome, but that does not neccessarily imply that the prediction provides clinically useful information; the signature may divide patients into subgroups with different prognosis, but if there are no therapies available to improve the outcome of either subgroup, then the signature may have little clinical value.
+While the focus here is statistical estimation of a numerical performance metric, it should be kept in mind that clinical usefulness may depend on a variety of additional factors.
 
@@
 
-A **biomarker signature** is a transformation of multiple individual features, typically molecular characterstics measured on a multiplex assay, to a one-dimensional space.
-Specifically, let $X$ denote the set of $p$ features under consideration.
+A **biomarker signature** is a transformation of multiple individual feature measurements, typically molecular characteristics measured on a multiplex assay, to a one-dimensional space.
+Specifically, let $X$ denote the set of $p$ feature measurements under consideration.
 The signature is an unknown function $f(X): \mathbb{R}^p \mapsto \mathbb{R}^1$.
 The signature result may be continuous, take multiple discrete values, or be dichotomous.
-
-Let $S$ denote the development dataset which includes, for each of $n$ represented individuals, a feature vector $X$, an outcome $Y$, a treatment $Z$, and possibly other variables.
+In principle, the signature could also depend on other clinical or pathological variables in addition to the molecular measurements, but to simplify the discussion we will focus on signatures that depend on molecular feature data only <need to do this because you refer here and later to f as having domain Rp>.
+Let $S$ denote the development dataset which includes, for each of $n$ represented individuals, a feature vector $X$, an outcome $Y$, and a treatment $Z$.
 $S$ is a sample of size $n$ from distribution $\mathcal{P}$ with domain $\mathcal{S}$.
-Let $\mathcal{F}$ be a mapping from $\mathcal{X}$ to the space of continuous functions with domain $\mathbb{R}^p$ and range $\mathbb{R}$, $\mathcal{D}$.
-Thus $\mathcal{F}: \mathcal{X} \mapsto \mathcal{D}$ denotes the process or algorithm through which a particular $f$ is estimated.
+<Do you need this bit about domain $\mathcal{S}$?
+P is a distribution function so I don't think it is necessary to talk about its domain.> Let $\mathcal{F}$ be a mapping from $\mathcal{X}$ to the space of continuous functions, $\mathcal{D}$, with domain $\mathbb{R}^p$ and range $\mathbb{R}$.
+<The symbol $\mathcal(X) has not been defined.
+It is actually the space of development data sets that include not only X but also Y and Z for each of n individuals.
+Please use a letter different than X or readers be confused and think that it includes only the feature vectors.> <Here again you describe f as having domain Rp.>  Thus $\mathcal{F}: \mathcal{X} \mapsto \mathcal{D}$ denotes the process or algorithm through which a particular $f$ is estimated.
 We do not place any other restrictions on $\mathcal{F}$, it could be a clustering approach, a regression approach, a combination of both, or something else entirely.
@@
 
-Let $\phi: \mathcal{D} \times \mathcal{X} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC).
+Let $\phi: \mathcal{D} \times \mathcal{X} \mapsto \mathbb{R}$ <Change mathcal(X) here if you decide to follow my suggestion above to change X to a different letter.> denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC).
 This could also be a measure of association, such as an odds ratio, hazard ratio, or log-rank statistic.
 This is a function of both $f$ and $S$.
-We are interested in estimating $E_\mathcal{P}[\phi_{f^*}(S)]$, which is the expected error under the data generation mechanism, for a particular $f^* \in \mathcal{F}$.
+We are interested in estimating $E_\mathcal{P}[\phi_{f^*}(S)]$, which is the expected error under the data generation mechanism (distribution $\mathcal{P}$), for a particular $f^* \in \mathcal{F}$.
 This allows us to understand how the signature will perform on future observations generated from $\mathcal{P}$.
-We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ *for all* $f \in \mathcal{F}$, which is the generalization error for $f$ generated using mechanism $\mathcal{F}$.
+We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ *for all* $f \in \mathcal{F}$, which is the generalization performance for $f$ generated using mechanism $\mathcal{F}$.
 This doesn't guide outside researchers as to which specific $f$ to use, yet it is useful for development because it tells us how much signal is in the data.
@@
 
-A signature that **reliably predicts** an outcome $Y$ is one that has generalization error small enough for the clinical context.
-Such a signature may be useful for treatment selection, prognosis, or other type of clinical management.
-
 ## Overview of biomarker signature development
 
-The main goal is to produce a good signature.
-To establish this requires a valid estimate of the signature's performance.
-Assessment of the worth of a biomarker signature requires providing a valid estimate of the performance of $f \in \mathcal{F}$.
-Additionally, one can provide a specification of $f$ for others to use.
+The inherent statistical challenge is how to both develop a signature $f \in \mathcal{F}$ and obtain a valid assessment of its performance.
+Additionally, one should provide a specification of $f$ for others to use.
 Typically, a specific $f$ is estimated using $\mathcal{F}$ based on some training data.
-This can be done using a variety of different methods.
-In recent years there have been an explosion in the literature of computational approaches to classification and prediction, and we do not intend to summarize them all here.
-Some excellent reviews of such approaches are @hastie2009elements and @moons2012riskI.
-The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply, determining how to combine the feature measurements, and whether to apply thresholds/cutoffs to the resulting signature value.
+The class of methods $\mathcal{F}$ can comprise a variety of different computational approaches.
+In recent years there has been an explosion in the literature of computational approaches to classification and prediction, and we do not intend to summarize them all here.
+Some excellent reviews are provided by @hastie2009elements and @moons2012riskI.
+The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply to the feature measurements, determining how to combine the feature measurements, and determining what, if any, thresholds/cutoffs should be applied to the resulting signature value.
 
-A biomarker signature can inform clinical practice in a number of ways.
-In early-stage disease, a highly prognostic signature may identify a subpopulation of patients that has such a good chance of long term survival, that they do not require additional treatment beyond some standard base therapy.
-Therefore these good risk patients can be spared the risks and side-effects associated with additional therapy.
-In the context of a specific therapy that targets a particular molecular pathway, a signature may identify a subpopulation of patients that does or does not benefit from that therapy, thereby guiding the decision to treat or not.
+Our focus here will be mainly on the development of signatures using supervised methods, meaning methods that use information on an outcome variable it is desired to predict during the signature development process.
+Regression methods are a commonly used class of supervised methods.
+Oncotype DX [@paik2004multigene] is an example of a prognostic signature used for clinical decision making in early stage breast cancer that was developed using supervised methods.
+In the case of Oncotype DX, the outcome $Y$ was time to distant recurrence of breast cancer and the feature data were gene expression values measured in breast tumor specimens.
+An common supervised approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment.
+Supervised methods are contrasted with unsupervised methods which use only the feature data to derive a signature.
+An example of a signature developed by unsupervised methods is the signature that identifies biological subclasses of diffuse large B-cell lymphoma, which were originally developed using clustering methods and subsequently were found to be associated with prognosis [@alizadeh2000lymphoma].
+It is possible, and quite common in high-dimensional data settings, to combine multiple approaches when estimated a signature.
+For instance, a data-reduction step by variable selection or clustering may be performed before doing regression analysis on the resulting components.
+A review by Subramanian and Simon identified a large collection of gene expression–based prognostic signatures in lung cancer which had been developed using a wide variety of methods [@subramanian2010lung].
 
-Signatures can be estimated by identifying naturally occuring clusters, or intrinsic subtypes using $X$ alone without regard to the outcome $Y$.
-The PAM-50 gene signature was developed using clustering methods in an attempt to identify intrinsic subtypes with different prognosis [@parker2009supervised].
-Later, the intrinsic subtypes were shown to be strongly associated with clinical outcomes [@nielsen2010comparison].
-Supervised learning techniques can also be used to identify prognostic signatures, as was the case with Oncotype DX [@paik2004multigene], another signature developed for clinical decision making in early stage breast cancer.
-In this case, regression-based methods are used to estimate a model that is highly predictive for the outcome $Y$.
-In the case of Oncotype DX, the outcome in question was distant recurrence of breast cancer.
-
-An approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment.
-It is possible, and quite common in high-dimensional settings, to combine multiple approaches to estimating $f$.
-For instance, a data-reduction step by variable selection or clustering may be performed before doing regression analysis on the resulting components.
-
-No matter what the particular model building method is, our main concern and focus of this paper is with obtaining a valid estimate of its performance, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
-This depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used.
-An optional component of the development phase is to provide a specification of $f$ for others to use on independent data or in clinical practice.
-
-## Biomarker signatures in clinical practice
-
-Biomarker signatures are useful if they can correctly and reliably classify patients into distinct subgroups for which different treatment decisions would be made.
-There are two distinct but related statistical concepts involved here: calibration and discrimination.
-Signatures are often optimized to be well-calibrated, that is, accurate for predicting outcomes.
-However if the signature does not separate a population into distinct subgroups with therapeutic relevance, then it is unlikely to be informative enough to change clinical practice.
-
-In the development process it is important to evaluate both calibration and discrimination.
-It is not trivial to assess each of these in a valid manner when the data are used to define the signature itself.
-We illustrate the potential for bias through a series of examples, and we review some strategies to avoid the pitfalls that lead to bias.
-
-## Data analysis example
-
-Throughout this paper, we reanalyze data from @zhu2010prognostic.
-Briefly, the data of interest are from the JBR.10 trial, which was a randomized controlled trial of adjuvant vinorelbine/cisplatin (ACT) versus observation alone (OBS) in 482 participants with non small cell lung cancer (NSCLC).
-Of those 482 participants, 169 had frozen tissue collected, and of those samples, 133 (71 in ACT and 62 in OBS) had gene-expression profiling performed using U133A oligonucleotide microarrays (Affymetrix, Santa Clara, CA).
-
-The goal of the @zhu2010prognostic paper was to identify a multi-gene signature that strongly predicts prognosis, and the hypothesis was that the poor prognosis subgroup would benefit more from ACT compared to the good prognosis subgroup.
-The signature was trained to predict disease specific survival.
-The annotated gene expression data and clinical information are available from the Gene Expression Omnibus (identifier: GSE14814, @edgar2002gene).
-
-@zhu2010prognostic present results that mainly focus on the discrimination ability of their estimated signature.
-They do that by demonstrating that the two risk subgroups predicted by their signature (high risk and low risk) have separation in their survival curves and that the hazard ratio for their signature is large and significant even when adjusting for other risk factors.
-They do not directly address calibration, that is, whether their signature accurately predicts survival times.
-
-We used a similar approach to preprocessing as did @zhu2010prognostic, as we could not reproduce their workflow exactly due to outdated software.
-Batch effects were removed using the `ComBat` function in the `sva` `R` package [@leeksva] and then the gene expression values were centered by their means and scaled by their standard deviations.
-Our signature development approach is similar but not identical to that in @zhu2010prognostic.
-For purposes of illustrating the concepts we used a simplyfied approach to signature development that retains the main features of the original.
-The exact approach to signature development does not have a major impact on the main conclusions of our evaluation of the various approaches to signature performance assessment.
-
-Following the processing steps described above, we performed a gene selection step wherein we fit univariate Cox regression models with disease specific survival as the outcome and each gene as the single predictor.
-Genes with univariate p-values less than 0.005 were selected for further analysis.
-Then, each gene was weighted by its univariate Cox regression coefficient, and the resulting weighted gene expression values summed to form risk scores.
-Genes were selected in a forward selection manner, starting with the most significant genes, the gene that improved the concordance between survival times and the risk score was selected.
-If no gene improved the concordance, the process was stopped.
-The final list of selected genes were all included in a multivariable Cox regression model to fit the final risk score.
-The cutoff that yielded the smallest log-rank statistic p-value was used to dichotomize into two risk groups.
+The focus of this paper is how to obtain a valid estimate of signature performance as quantified by a metric $phi$, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
+The principles discussed apply no matter what signature development method is used.
+Performance depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used for development.
+The performance will reflect that expected of signatures developed applying the prescribed methods to data drawn from the distribution $mathcal{P}$.
+For signature deployment, one would also want a particular specification of $f$ for others to use on independent data (e.g., in clinical practice).
+It is not trivial to conduct this evaluation in a valid manner when there is a limited data set available to both define a signature and assess its performance.
+We illustrate the potential for bias to creep into the performance and review some strategies to avoid the pitfalls that lead to these biases.
+We evaluate various strategies through a limited simulation study and compare the performance of some of those strategies using a real data exmaple involving the developmnet of a signature for use in making treatment decisions for patients with lung cancerseries.
 
@@
 
-Recall that the main goal is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected value of a given statistic on future observations for $f \in \mathcal{F}$.
+The main goal of interest here is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected value of a given performance metric on future observations for $f \in \mathcal{F}$.
 This can be estimated with the in-sample empirical estimate: $\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)$ for a particular $f$.
 However, if $S$ is used to estimate $f$ then the estimate will be biased due to overfitting, that is, $|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|$ will be large.
-This bias results from the fact that $\phi$ depends on $f$, and thus the statistic $\phi$ is being adaptively defined based on the observed data $S$.
 Overfitting occurs when a model is fit to noise in the data.
 This often occurs when fitting a model that is overly complex relative to the amount of signal in the available data.
+Overfitting bias results from the fact that $\phi_f(S)$ depends on $f$ which depends on $S$, and thus the statistic $\phi_f(S)$ is being adaptively defined based on the observed data $S$.
+Any estimate of signature performance is potentially subject to bias if overfitting occurs, but such biases frequently go unrecognized in the medical literature.
 
-In many cases during signature development, the performance metric $\phi$ in question is a statistic that relates to calibration.
-While a biomarker signature may accurately predict a clinical outcome, that does not neccessarily imply that the signature is clinically useful.
-To assess discrimination, a different statistic $\phi$ may be used, such as the area under the ROC curve.
-Measures of association such as the odds ratio, hazard ratio, or difference in survival probabilities are examples of performance metrics that are commonly used, although their value for biomarker signatures has been debated [@pepe2004limitations].
-Evaluation of $\phi$ is subject to bias due to overfitting regardless of what $\phi$ is used, a fact that is commonly overlooked in the medical literature.
-
-In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups as defined by the JBL.10 signature $f$.
-The risk groups were determined by the signature $f$, which was estimated using $S$, the same data that was then used to produce the hazard ratio estimate of 15.
-They go on to assess the performance of the JBL.10 signature in a series of independent data sets, in which they estimate hazard ratios of approximately 2.
-Here we reanalyze the dataset and illustrate some remedies to avoid bias in estimating signature performance.
+Typically a performance metric $\phi$ used during signature development is degined to quantify either calibration or discrimination, or some combination of the two.
+<Define calibration and discrimination.> Calibration assesses <fill in>.
+To assess discrimination, a different metric $\phi$ may be used, such as the area under the ROC curve.
+Measures of association such as the odds ratio, hazard ratio, or difference in survival probabilities are also commonly used performance metrics, although their value for biomarker signatures has been debated [@pepe2004limitations].
+In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups.
+The risk groups were determined by the signature $f$, which was estimated using the JBL.10 observation arm data.
+The signature was then applied to the same data that were used to build it to define risk groups, producing a the hazard ratio estimate of 15 for a prognostic effect in the observation arm.
+Subsequent evaluations of the signature's performance on independent data resulted in hazard ratio estimates in the 2 - 3 range, substantially smaller than the original estimate of 15.
+Next we reanalyze the JBR.10 data and illustrate some methods to avoid bias in evaluation of signature performance.
 
@@
 Then, $S_h$ is hidden from the analyst while $\mathcal{F}$ is applied to $S_t$ to estimate the signature function $f_t$.
-For fixed $f_t$, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of both $E_{\mathcal{P}}[\phi_{f_t}(S)]$ and $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
-The specific form of $f_t$ that is fixed using the $S_t$ partition can be reported as the function for others to use, therefore the aforementioned estimator is also an estimate of the error for that specific $f_t$.
+For fixed $f_t$, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of $E_{\mathcal{P}}[\phi_{f_t}(S)]$ <Is this notation correct?> but it may underestimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$ which refers to a predictors built from a sample of size n.
+@dobbin2011optimally investigate how to optimally split a dataset into training and holdout partitions.
+As n, $n_t$, and $n_h$ all become very large the degree of underestimation will become negligible as the performance of all predcitors will approach that of a signature developed from a sample of infinite size from the distribution mathcal{P}.
+The specific form of $f_t$ that is fixed using $S_t$ can be reported as the function for others to use, therefore the aforementioned estimator applies for that specific $f_t$.
 The drawback of the split-sample approach is that the performance of $f_t$ is likely to be inferior to the performance of a function $f^*$ derived using the same approach applied to the entire data set.
-@dobbin2011optimally investigate how to optimally split a dataset into training and holdout partitions.
+@dobbin2011predictorsamplesize investigate how to choose the sample size n so that there is high probability that the signature developed will have expected performance within some specified margin of the the optimal signature that could be developed if the development data set had infinite sample size.
 
 Another approach to avoid overfitting is cross-validation, which is a resampling based approach.
-For a fixed integer $k$, which can be between 1 and $n$, we randomly select a partition of $k$ observations from $S$, denoted $S_k$.
-Then $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the subset of $S$ that is disjoint from $S_k$.
-Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an unbiased estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
-This process is repeated $K$ times to yield $K$ estimates.
-Each of these estimates is unbiased, but noisy, because typically $k$ is very small relative to $n$.
-Thus, we average over $K$ to get a less noisy estimate.
-This process is called "leave $k$ out" or "$n/k$ fold" cross-validation.
-Note that for each partition that is selected, we obtain a new estimate of $f$, therefore we are only estimating $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$ and not the performance for a single specified signature.
+For a fixed integer $k$, which can be between 1 and $n$, we randomly partition the full data set $S$ into subgroups of size $k$ (assume for simplicity here that n is evenly divisible by k so that n/k subgroups are formed).
+For each $k$, $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the subset of $S$ that is disjoint from $S_k$.
+Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$, but it will have some pessimistic bias because a signature developed from a sample of size n-k has less information to work with than one based on a sample of size n.
+When n is very large relative to k this bias is expected to be minimal.
+This process is repeated $K$ times to yield $K$ estimates and then these are averaged to obtain the performance estimate.
+<Did you just cycled through the K subgroups in the partition or did you randomly draw subsets?
+Hopefully it is the former, in which case K = n/k.
+This should be clarified in the text.> This process is called "leave $k$ out" or "$n/k$ fold" cross-validation.
+Note that for each subgroup in the partition, we obtain a new estimate of $f$, therefore we are only estimating $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$ and not the performance for a single specified signature.
 Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$ to yield $f^*$ as above.
+Importantly one should report the cross-validated estimate just described and not the naive esitimate that would be obtained by substituting the full data S into the performance metric calculation for the signature $f^*$.
+The problematic aspects of such naive "resubstitution estimates" of performance are discuss in more detail below.
 
@@
 Then $f_{b}$ is estimated and fixed by applying $\mathcal{F}$ to $S_{b}$.
-The performance metric $\phi$ is calculated on the subset of $S$ that is disjoint from $S_b$: $S_{-b}$ to yield an estimate of $E[\phi_{f_b}(S_{-b})]$.
+The performance metric $\phi$ is calculated on the subset of $S$ that is disjoint from $S_b$, which we denote by $S_{-b}$, to yield an estimate of $E[\phi_{f_b}(S_{-b})]$.
 This process is repeated $K$ times to yield $K$ estimates.
-These $K$ estimates of the performance metrics are averaged to obtain the mean over the bootstrap replicates.
+These $K$ performance metric estimates are averaged to obtain the mean over the bootstrap replicates.
 @efron1997improvements suggest a variation, the 0.632 estimate:
@@
 Another variation on all of these methods is the concept of pre-validation [@tibshirani2002pre].
-With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{-b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted predictor $\hat{f}(X_i)$ is estimated for $X_i \in S_{-b}$ where $\hat{f}$ is estimated using $S_{b}$.
-This process is repeated to obtain a set of pre-validated predictor estimates $\hat{f}$ which are then used to calculate $\phi$.
-For single-step holdout, this process is equivalent to what is described above.
+With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{-b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted signature $\hat{f}(X_i)$ is estimated for $X_i \in S_{-b}$ where $\hat{f}$ is estimated using $S_{b}$.
+This process is repeated to obtain a set of pre-validated signature estimates $\hat{f}$ which are then used to calculate $\phi$ <I think that we have been using phi as the function that defines the performance metric rather than the performance metric estimate, correct?>.
+For single-step holdout <what is this?>, this process is equivalent to what is described above.
 For cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on the held-out dataset.
+<Do you mean that this avoids the diffciulties in using cross-validation adn bootstrapping?>
 
@@
 
-To illustrate the different properties of these estimates and how they deal with overfitting, we conduct a simulation study.
-Data were generated with 1000 observations, each with a binary outcome $Y$ with prevalence 0.3, and 500 features sampled from the standard normal distribution.
+To illustrate some of the different properties of these estimates and how they help to avoid overfitting, we conduct a limited simulation study.
+Data were generated with 1000 observations, each with a binary outcome $Y$ with prevalence 0.3, and 500 mutually independent features sampled from the standard normal distribution.
 This is the null case where no features are associated with $Y$.
@@
 
-We compare each of the methods described above, split-sample holdout, cross-validation, bootstrap, and pre-validation, along with several commonly used but biased approaches.
-All of the approaches are described in \ref{descript}.
+We compare each of the methods described above: split-sample holdout, cross-validation, bootstrap, and pre-validation, along with several commonly used but biased approaches.
+The biased approaches are described in Table \ref{descript}.
 Two of the biased approaches use the full sample to select the features, followed by fitting the multivariable model on the holdout subset.
 This is referred to as "parital holdout" or "partial CV" when using split-sample holdout or cross-validation as the validation step, respectively.
-We also implemented the naive resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is trained on a holdout set and then evaluated on the combined complete dataset.
+We also implemented the naive resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is developed on a training subset but then evaluated on the combined training adn holdout data sets.
 Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
@@
 
-Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the OR estimate is over 2 times higher than it should be, on average.
+Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the <log?> OR estimate is over 2 times <that is on the log scale but the table is not log scale> higher than it should be, on average.
 Partial resubstitution, partial holdout, and partial cross-validation estimates do not ameliorate the bias very much.
-Investigators often feel as though partial holdout estimates are close to valid, as only half of the data are used to form the estimates, however here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.
+Investigators are tempted to use partial holdout estimates as it is more convneient to not have to carry along a large number of feature measurements into the modeling part of the signature development process, and they may think that the resulting perofrmance esitmates are still close to valid, as only half of the data are used to form the estimates;  However here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.
 
-All of the proposed remedies to avoid these biases are successful and are unbiased, with their mean AUCs being nearly 0.5 and the mean ORs being nearly 1.
+The split-sample holdout, cross-validation, bootstrap, and pre-validation methods all produce esentially unbiased estimates in the simulated example, with their mean AUCs being nearly 0.5 and the mean ORs being nearly 1.
 We can compare the spread of the distributions to get a sense of the differences in precision of the estimates.
-The bootstrap approach appears to be the most precise of the unbiased estimates, followed by the cross-validation, holdout, and finally the pre-validation.
+The bootstrap approach appears to be the most precise, followed by the cross-validation, holdout, and finally the pre-validation.
 The bootstrap, as intended, is a more efficient, smoothed version of the cross validation estimate (Figure \ref{fig1}, \ref{fig2}).
-It provides the best balance between allocating data to precisely train the signature and having independent data remaining to precisely estimate the statistic $\phi$.
+It provides the best balance between allocating data to <I still don't like the word "precisely" so I have deleted it.
+Could say "adequately."  Precisely gives the impression that there is a single underlying form of the predictor and it is just a matter of more precisely fitting model coefficients.> train the signature and having independent data remaining to precisely estimate the statistic $\phi$.
 
@@
 
+## Data analysis example
+
+We now illustrate the concepts and methods just discussed by reanalyzing data that had been used to build a previously published lung cancer prognostic signature @zhu2010prognostic.
+Briefly, the data of interest are from the JBR.10 trial, which was a randomized controlled trial of the adjuvant chemotherapy regimen vinorelbine/cisplatin (ACT) versus observation alone (OBS) in 482 participants with non small cell lung cancer (NSCLC) who had undergone surgery.
+Of those 482 participants, 169 had frozen tumor tissue collected, and of those tumor samples, 133 (71 in ACT and 62 in OBS) had gene-expression profiling performed using U133A oligonucleotide microarrays (Affymetrix, Santa Clara, CA).
+
+The goal of the @zhu2010prognostic paper was to identify a multi-gene signature that strongly predicts prognosis, and the hypothesis was that the poor prognosis subgroup would benefit more from ACT than the good prognosis subgroup.
+The signature was developed on a training data set ("trained") to predict disease specific survival.
+The annotated gene expression data and clinical information are available from the Gene Expression Omnibus database(identifier: GSE14814, @edgar2002gene).
+
+@zhu2010prognostic present results that mainly focus on the association of their signature with outcome, albeit a less than ideal measure of discrimination ability.
+They demonstrate that the two risk subgroups predicted by their signature (high risk and low risk) have separation in their survival curves and that the hazard ratio for their signature is large and significant even when adjusting for other risk factors.
+They do not directly address calibration, that is, whether their signature accurately predicts survival times.
+
+We used a similar approach to preprocessing as did @zhu2010prognostic, although we could not reproduce their workflow exactly due to outdated software.
+Batch effects were removed using the `ComBat` function in the `sva` `R` package [@leeksva] and then the gene expression values were centered by their means and scaled by their standard deviations.
+Our signature development approach is similar but not identical to that in @zhu2010prognostic.
+For purposes of illustrating the concepts we used a simplified approach to signature development that retains the main characteristics of the original method.
+The exact approach to signature development does not have a major impact on the main conclusions of our evaluation of the various approaches to signature performance assessment.
+
+After processing the data as described above, we performed a gene selection step wherein we fit univariate Cox regression models with disease specific survival as the outcome and each gene as the single predictor.
+Genes with univariate p-values less than 0.005 were preliminarily selected for further analysis.
+Then, each gene from the preliminary list was weighted by its univariate Cox regression coefficient, and the resulting weighted gene expression values were summed to form risk scores.
+Genes were selected for inclusion in the risk score in a forward selection manner.
+Starting with the most significant weighted gene, the gene that when added to the risk score improved the concordance between survival times and the risk score most was selected next.
+If no gene improved the concordance, the process was stopped.
+All genes on the final selected list were included in a multivariable Cox regression model to fit the final risk score <Did you throw away the original coefficients and refit all of them in the final model that included the genes on the final list?>.
+The cutoff applied to the continuous risk score that split the patients into two risk groups corresponding to the smallest log-rank statistic p-value was selected.
 The signature development procedure was described in the introduction.
@@
 On one hand, the overfit signature shows dramatic differences in treatment efficacy between the low risk and high risk groups.
-In fact it appears that the treatment is harmful in the low risk group, but highly beneficial in the high risk group, a very rare finding.
+In fact it appears that the treatment is harmful in the low risk group, but highly beneficial in the high risk group.
 The prevalidated signature on the other hand, shows differences that are much less dramatic.
 It appears that the treatment is mildly beneficial in both groups, possibly to a higher degree in the high risk group.
-This suggests that the dramatic predictive value of the signature was merely an artifact of the overfitting process on the OBS arm, as pointed out by @simon2011re.
+This suggests that the dramatic predictive value of the signature was merely an artifact of overfitting the signature to the OBS arm data.
+@simon2011re explain the flaws in this type of approach in which one attempts to develop a signature of differential treatment benefit ("predictive") by developing a prognostic signature on data from the control arm of a trial and then applying that signature to both the control and experimental arm data to establish predictiveness.
 
@@
 For the partial resubstitution approach, we find an extreme hazard ratio of nearly 40 for the prognostic effect, and a strong and significant treatment by signature interaction.
-Using the prevalidated signature, the effect estimates are much smaller and insignificant in comparison to the standard clinical features.
-Note that the inference from the pre-validated model is not exactly correct either, because the procedure induces a correlation among the observations.
+Using the prevalidated signature, the effect estimates are much smaller and of small magnitude in comparison to the standard clinical features.
+<You shoudl present a tabel witht the ful lresults of the mutivariable Cox regression fits.> Note that the inference from the pre-validated model is not exactly correct either, because the resampling procedure induces a correlation among the observations (the prevalidated signature values are estimated from heavily overlapping data).
 Despite this, these results are unimpressive and would not be considered promising for clinical use.
 
-## Additional Examples
-
-Examples of resubstitution estimates of performance are highly prevalent in the literature.
-@zhu2010prognostic was not unique in that sense, but rather they should be commended for their commitment to making their data, methods, and analysis code publically available, which allowed us to reproduce and reanalyze their study.
-More often, the methods are obfuscated through vague and non-standard descriptions, short methods sections, and hidden data analysis code.
-Most of the time, a resubstitution analysis can be discovered only through a very careful scrutiny of the methods and supplementary materials, if at all.
-Here we give some additional examples.
-
-@van2002gene used gene expression measurements to develop a signature and predict clinical outcome of patients with axillary lymph node-negative breast cancer (metastatic disease within 5 years versus disease-free at 5 years).
-The signature was developed using a stepwise selection and classification procedure, in which a subset of the genes was first selected using all of the subjects, then further subsets were selected using cross-validation until they arrived at a 70 gene signature.
-This is the classic form of the partial cross-validation as we've described above.
-The resubstitution estimate of the odds ratio to develop distant metastasis within 5-years was estimated to be 28, with $p < 0.0001$.
-A later study of a similar population used similar gene expression profiling approach, but with a valid split sample holdout validation approach [@wang2005gene].
-A univariate analysis of the locked-down signature evaluated on the holdout set gave a much more conservative odds ratio estimate of 5.7 for distant metastasis within 5 years.
-
-@zhang2001recursive describe a tree-based partitioning approach to use gene expression data to classify specimens as either cancer or normal tissue.
-Again, from a large number of candidate genes, a subset of three genes is identified using the entire set of 48 tissue specimens.
-Then, the cutoff values for the classifications are re-estimated using cross validation.
-This is yet again the classic form of partial cross validation that we have shown to be biased, contrary to their reports of "unbiased" estimates of 6-8% missclassification rates.
-This key detail of the analysis is only briefly described in the results section.
+@zhu2010prognostic should be commended for their commitment to making their data, methods, and analysis code publically available, which allowed us to reanalyze their study data.
+More often, methods used to derive signatures are vaguely and incompletely presented and a resubstitution analysis or other flawed approach goes undetected or can be discovered only through a very careful scrutiny of the methods and supplementary materials.
 
@@
 
-All statistics, whether they assess calibration or discrimination or something else, are subject to bias due to overfitting.
-Remedies to this type of bias are well-studied in the statistical literature and here we have demonstrated how they can be implemented in a real scenario.
+Examples of flawed approaches such as naive resubstitution, partial resubstitution, partial holdout, and partial CV remain very common in the published literature more than a decade after many high-dimensional molecular assays were first introduced.
+All performance metrics, whether they assess calibration or discrimination or something else, are subject to bias in estimation due to overfitting.
+Strategies to avoid this type of bias are well-studied in the statistical literature and here we have demonstrated how they can be implemented in a real clinical example.
+The ability to perform the kind of in-depth data reanalysis we performed on the example in this paper is usually limited by restrictions on data access and insufficient description of analysis methods.
 Sadly, reports of strong associations with overfit biomarker signatures are all too common in the medical literature.
 The amount of bias that is possible is not known and can be difficult to decipher based on study reports.
-It is imperative that investigators and journal editors take overfitting bias seriously to ensure that signature estimates are valid.
+It is imperative that investigators and journal editors take overfitting bias seriously to ensure that signature performance claims are valid.
 Unfortunately this often results in study reports that are far less optimistic than usual.
+
+I STOPPED HERE AND WILL HAVE TO WORK MORE ON THE DISCUSSION TOMORROW.
 

### 2016-05-18 Documents/paper-revised.Rmd
incorporated lisas edits, waiting on final discussion revisions

--- 2016-05-18 Documents/paper revised 5-10-16+LM.Rmd
@@
       pandoc_args: ["--bibliography", "paper.bib"]
-fontsize: 12pt
+header-includes:
+    - \usepackage{setspace}
+    - \doublespacing
+    - \usepackage{lineno}
+    - \linenumbers
+fontsize: 11pt
 geometry: margin=1in
@@
 library(tidyr)
+library(stargazer)
 
@@
 The signature result may be continuous, take multiple discrete values, or be dichotomous.
-In principle, the signature could also depend on other clinical or pathological variables in addition to the molecular measurements, but to simplify the discussion we will focus on signatures that depend on molecular feature data only <need to do this because you refer here and later to f as having domain Rp>.
+In principle, the signature could also depend on other clinical or pathological variables in addition to the molecular measurements, but to simplify the discussion we will focus on signatures that depend on molecular feature data only.
 Let $S$ denote the development dataset which includes, for each of $n$ represented individuals, a feature vector $X$, an outcome $Y$, and a treatment $Z$.
 $S$ is a sample of size $n$ from distribution $\mathcal{P}$ with domain $\mathcal{S}$.
-<Do you need this bit about domain $\mathcal{S}$?
-P is a distribution function so I don't think it is necessary to talk about its domain.> Let $\mathcal{F}$ be a mapping from $\mathcal{X}$ to the space of continuous functions, $\mathcal{D}$, with domain $\mathbb{R}^p$ and range $\mathbb{R}$.
-<The symbol $\mathcal(X) has not been defined.
-It is actually the space of development data sets that include not only X but also Y and Z for each of n individuals.
-Please use a letter different than X or readers be confused and think that it includes only the feature vectors.> <Here again you describe f as having domain Rp.>  Thus $\mathcal{F}: \mathcal{X} \mapsto \mathcal{D}$ denotes the process or algorithm through which a particular $f$ is estimated.
+Let $\mathcal{F}$ be a mapping from $\mathcal{S}$ to the space of continuous functions, $\mathcal{D}$, with domain $\mathbb{R}^p$ and range $\mathbb{R}$.
+Thus $\mathcal{F}: \mathcal{S} \mapsto \mathcal{D}$ denotes the process or algorithm through which a particular $f$ is estimated.
 We do not place any other restrictions on $\mathcal{F}$, it could be a clustering approach, a regression approach, a combination of both, or something else entirely.
@@
 
-Let $\phi: \mathcal{D} \times \mathcal{X} \mapsto \mathbb{R}$ <Change mathcal(X) here if you decide to follow my suggestion above to change X to a different letter.> denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC).
+Let $\phi: \mathcal{D} \times \mathcal{S} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC).
 This could also be a measure of association, such as an odds ratio, hazard ratio, or log-rank statistic.
@@
 This allows us to understand how the signature will perform on future observations generated from $\mathcal{P}$.
-We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ *for all* $f \in \mathcal{F}$, which is the generalization performance for $f$ generated using mechanism $\mathcal{F}$.
+We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ **for all** $f \in \mathcal{F}$, which is the generalization performance for $f$ generated using mechanism $\mathcal{F}$.
 This doesn't guide outside researchers as to which specific $f$ to use, yet it is useful for development because it tells us how much signal is in the data.
-As shorthand we will write this as $E_\mathcal{P}[\phi_\mathcal{F}(S)]$
+As shorthand we will write this as $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
 
@@
 Some excellent reviews are provided by @hastie2009elements and @moons2012riskI.
-The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply to the feature measurements, determining how to combine the feature measurements, and determining what, if any, thresholds/cutoffs should be applied to the resulting signature value.
+The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply to the feature measurements, determining how to combine the feature measurements, and determining what, if any, thresholds or cutoffs should be applied to the resulting signature value.
 
@@
 In the case of Oncotype DX, the outcome $Y$ was time to distant recurrence of breast cancer and the feature data were gene expression values measured in breast tumor specimens.
-An common supervised approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment.
+A common supervised approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment.
 Supervised methods are contrasted with unsupervised methods which use only the feature data to derive a signature.
-An example of a signature developed by unsupervised methods is the signature that identifies biological subclasses of diffuse large B-cell lymphoma, which were originally developed using clustering methods and subsequently were found to be associated with prognosis [@alizadeh2000lymphoma].
+An example of a signature developed by unsupervised methods is the signature that identifies biological subclasses of diffuse large B-cell lymphoma, which were originally developed using clustering methods and subsequently were found to be associated with prognosis [@alizadeh2000distinct].
 It is possible, and quite common in high-dimensional data settings, to combine multiple approaches when estimated a signature.
 For instance, a data-reduction step by variable selection or clustering may be performed before doing regression analysis on the resulting components.
-A review by Subramanian and Simon identified a large collection of gene expression–based prognostic signatures in lung cancer which had been developed using a wide variety of methods [@subramanian2010lung].
+A review by Subramanian and Simon identified a large collection of gene expression–based prognostic signatures in lung cancer which had been developed using a wide variety of methods [@subramanian2010gene].
 
@@
 Performance depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used for development.
-The performance will reflect that expected of signatures developed applying the prescribed methods to data drawn from the distribution $mathcal{P}$.
+The performance will reflect that expected of signatures developed applying the prescribed methods to data drawn from the distribution $\mathcal{P}$.
 For signature deployment, one would also want a particular specification of $f$ for others to use on independent data (e.g., in clinical practice).
 It is not trivial to conduct this evaluation in a valid manner when there is a limited data set available to both define a signature and assess its performance.
-We illustrate the potential for bias to creep into the performance and review some strategies to avoid the pitfalls that lead to these biases.
-We evaluate various strategies through a limited simulation study and compare the performance of some of those strategies using a real data exmaple involving the developmnet of a signature for use in making treatment decisions for patients with lung cancerseries.
+We illustrate the potential for bias to creep into the performance estimation and review some strategies to avoid the pitfalls that lead to these biases.
+We evaluate various strategies through a simulation study and compare the performance of some of those strategies using a real data exmaple involving the development of a signature for use in making treatment decisions for patients with lung cancer.
 
@@
 Typically a performance metric $\phi$ used during signature development is degined to quantify either calibration or discrimination, or some combination of the two.
-<Define calibration and discrimination.> Calibration assesses <fill in>.
+Calibration assesses how well the signature based predictions compare to the observed outcomes.
+Discrimination assesses how well the signature distinguishes between groups of patients who do or do not experience an event.
 To assess discrimination, a different metric $\phi$ may be used, such as the area under the ROC curve.
@@
 Then, $S_h$ is hidden from the analyst while $\mathcal{F}$ is applied to $S_t$ to estimate the signature function $f_t$.
-For fixed $f_t$, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of $E_{\mathcal{P}}[\phi_{f_t}(S)]$ <Is this notation correct?> but it may underestimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$ which refers to a predictors built from a sample of size n.
+For fixed $f_t$, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of $E_{\mathcal{P}}[\phi_{f_t}(S)]$.
 @dobbin2011optimally investigate how to optimally split a dataset into training and holdout partitions.
-As n, $n_t$, and $n_h$ all become very large the degree of underestimation will become negligible as the performance of all predcitors will approach that of a signature developed from a sample of infinite size from the distribution mathcal{P}.
 The specific form of $f_t$ that is fixed using $S_t$ can be reported as the function for others to use, therefore the aforementioned estimator applies for that specific $f_t$.
 The drawback of the split-sample approach is that the performance of $f_t$ is likely to be inferior to the performance of a function $f^*$ derived using the same approach applied to the entire data set.
-@dobbin2011predictorsamplesize investigate how to choose the sample size n so that there is high probability that the signature developed will have expected performance within some specified margin of the the optimal signature that could be developed if the development data set had infinite sample size.
+@dobbin2007sample investigate how to choose the sample size n so that there is high probability that the signature developed will have expected performance within some specified margin of the the optimal signature that could be developed if the development data set had infinite sample size.
 
 Another approach to avoid overfitting is cross-validation, which is a resampling based approach.
-For a fixed integer $k$, which can be between 1 and $n$, we randomly partition the full data set $S$ into subgroups of size $k$ (assume for simplicity here that n is evenly divisible by k so that n/k subgroups are formed).
+For a fixed integer $k$, which can be between 1 and $n$, we randomly partition the full data set $S$ into subgroups of size $k$ (assume for simplicity here that $n$ is evenly divisible by $k$ so that $n/k$ subgroups are formed).
 For each $k$, $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the subset of $S$ that is disjoint from $S_k$.
-Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$, but it will have some pessimistic bias because a signature developed from a sample of size n-k has less information to work with than one based on a sample of size n.
-When n is very large relative to k this bias is expected to be minimal.
+Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
 This process is repeated $K$ times to yield $K$ estimates and then these are averaged to obtain the performance estimate.
-<Did you just cycled through the K subgroups in the partition or did you randomly draw subsets?
-Hopefully it is the former, in which case K = n/k.
-This should be clarified in the text.> This process is called "leave $k$ out" or "$n/k$ fold" cross-validation.
+This process is called "leave $k$ out" or "$n/k$ fold" cross-validation.
 Note that for each subgroup in the partition, we obtain a new estimate of $f$, therefore we are only estimating $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$ and not the performance for a single specified signature.
 Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$ to yield $f^*$ as above.
-Importantly one should report the cross-validated estimate just described and not the naive esitimate that would be obtained by substituting the full data S into the performance metric calculation for the signature $f^*$.
+Importantly one should report the cross-validated estimate just described and not the naive esitimate that would be obtained by substituting the full data $S$ into the performance metric calculation for the signature $f^*$.
 The problematic aspects of such naive "resubstitution estimates" of performance are discuss in more detail below.
@@
 With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{-b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted signature $\hat{f}(X_i)$ is estimated for $X_i \in S_{-b}$ where $\hat{f}$ is estimated using $S_{b}$.
-This process is repeated to obtain a set of pre-validated signature estimates $\hat{f}$ which are then used to calculate $\phi$ <I think that we have been using phi as the function that defines the performance metric rather than the performance metric estimate, correct?>.
-For single-step holdout <what is this?>, this process is equivalent to what is described above.
-For cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on the held-out dataset.
-<Do you mean that this avoids the diffciulties in using cross-validation adn bootstrapping?>
+This process is repeated to obtain a set of pre-validated signature estimates $\hat{f}$ which are then used to calculate $\phi$.
+For single-step split sample training and validation, this process is equivalent to what is described above.
+For cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on each of the smaller held-out datasets.
 
@@
 This is referred to as "parital holdout" or "partial CV" when using split-sample holdout or cross-validation as the validation step, respectively.
-We also implemented the naive resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is developed on a training subset but then evaluated on the combined training adn holdout data sets.
+We also implemented the naïve resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is developed on a training subset but then evaluated on the combined training adn holdout data sets.
 Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
@@
 
-Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the <log?> OR estimate is over 2 times <that is on the log scale but the table is not log scale> higher than it should be, on average.
+Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the log OR estimate is over 2 times higher than it should be, on average.
 Partial resubstitution, partial holdout, and partial cross-validation estimates do not ameliorate the bias very much.
-Investigators are tempted to use partial holdout estimates as it is more convneient to not have to carry along a large number of feature measurements into the modeling part of the signature development process, and they may think that the resulting perofrmance esitmates are still close to valid, as only half of the data are used to form the estimates;  However here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.
+Investigators are tempted to use partial holdout estimates as it is more convenient to not have to carry along a large number of feature measurements into the modeling part of the signature development process, and they may think that the resulting performance estimates are still close to valid, as only half of the data are used to form the estimates.
+However, here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.
 
@@
 The bootstrap, as intended, is a more efficient, smoothed version of the cross validation estimate (Figure \ref{fig1}, \ref{fig2}).
-It provides the best balance between allocating data to <I still don't like the word "precisely" so I have deleted it.
-Could say "adequately."  Precisely gives the impression that there is a single underlying form of the predictor and it is just a matter of more precisely fitting model coefficients.> train the signature and having independent data remaining to precisely estimate the statistic $\phi$.
+It provides the best balance between allocating data to train the signature and having independent data remaining to precisely estimate the statistic $\phi$.
 
@@
 If no gene improved the concordance, the process was stopped.
-All genes on the final selected list were included in a multivariable Cox regression model to fit the final risk score <Did you throw away the original coefficients and refit all of them in the final model that included the genes on the final list?>.
-The cutoff applied to the continuous risk score that split the patients into two risk groups corresponding to the smallest log-rank statistic p-value was selected.
+All genes on the final selected list were included in a multivariable Cox regression model to fit the final risk score with new coefficients.
+We selected the cutoff that split the patients into two risk groups corresponding to the smallest log-rank statistic p-value when applied to the continuous risk score.
+
 The signature development procedure was described in the introduction.
@@
 This suggests that the dramatic predictive value of the signature was merely an artifact of overfitting the signature to the OBS arm data.
-@simon2011re explain the flaws in this type of approach in which one attempts to develop a signature of differential treatment benefit ("predictive") by developing a prognostic signature on data from the control arm of a trial and then applying that signature to both the control and experimental arm data to establish predictiveness.
+@simon2011re explain the flaws in this type of approach in which one attempts to develop a signature for treatment selection by developing a prognostic signature on data from the control arm of a trial and then applying that signature to both the control and experimental arm data to establish predictiveness.
 
@@
 The results are reported in Table \ref{adjhr}.
+The full model summaries are report in Table \ref{multimodel}.
 For the partial resubstitution approach, we find an extreme hazard ratio of nearly 40 for the prognostic effect, and a strong and significant treatment by signature interaction.
 Using the prevalidated signature, the effect estimates are much smaller and of small magnitude in comparison to the standard clinical features.
-<You shoudl present a tabel witht the ful lresults of the mutivariable Cox regression fits.> Note that the inference from the pre-validated model is not exactly correct either, because the resampling procedure induces a correlation among the observations (the prevalidated signature values are estimated from heavily overlapping data).
+Note that the inference from the pre-validated model is not exactly correct either, because the resampling procedure induces a correlation among the observations (the prevalidated signature values are estimated from heavily overlapping data).
 Despite this, these results are unimpressive and would not be considered promising for clinical use.
@@
 More often, methods used to derive signatures are vaguely and incompletely presented and a resubstitution analysis or other flawed approach goes undetected or can be discovered only through a very careful scrutiny of the methods and supplementary materials.
+
+```{r modtab, results = "asis"}
+stargazer(fit.train, fit.preval, fit.train.inter, fit.preval.inter, 
+          column.labels = rep(c("Resubstitution", "Prevalidation"), 2), 
+          covariate.labels = c("High Risk", "Trt: OBS", "Male Gender", "Stage > I", "Age in Years", "Histology: LCUC", 
+                               "Histology: SQCC", "High Risk * Trt:OBS"), 
+          dep.var.labels = "Disease-specific Survival", ci = TRUE, ci.separator = " to ", digits = 1, 
+          keep.stat = c("n", "logrank"), header = FALSE, apply.coef = exp,  apply.ci = exp, p.auto = FALSE,
+          ci.custom = lapply(list(fit.train, fit.preval, fit.train.inter, fit.preval.inter), confint),
+          title = "Complete Results of Multivariable Regression Models. Entries are hazard ratios and 95 percent confidence intervals.  \\label{multimodel}")
+```
 
@@
 Unfortunately this often results in study reports that are far less optimistic than usual.
-
-I STOPPED HERE AND WILL HAVE TO WORK MORE ON THE DISCUSSION TOMORROW.
 
@@
 
-All analysis code and the source files for this manuscript is available from the authors' webpage.
+All analysis code and the source files for this manuscript are available from the authors' webpage.
 

### 2016-05-18 Documents/paper-revised.tex
incorporated lisas edits, waiting on final discussion revisions

--- 2016-05-18 Documents/paper-revised.Rmd
@@
----
-title: "Issues in developing multivariable molecular signatures for guiding clinical care decisions"
-author: "Michael C Sachs and Lisa M McShane"
-date: "`r format(Sys.Date(), '%B %d, %Y')`"
-output: 
-    pdf_document:
-      citation_package: natbib
-      keep_tex: yes
-      fig_caption: yes
-      pandoc_args: ["--bibliography", "paper.bib"]
-header-includes:
-    - \usepackage{setspace}
-    - \doublespacing
-    - \usepackage{lineno}
-    - \linenumbers
-fontsize: 11pt
-geometry: margin=1in
----
+\documentclass[11pt,]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\usepackage[margin=1in]{geometry}
+\usepackage{natbib}
+\bibliographystyle{plainnat}
+\usepackage{longtable,booktabs}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={Michael C Sachs and Lisa M McShane},
+            pdftitle={Issues in developing multivariable molecular signatures for guiding clinical care decisions},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{0}
 
-# Abstract
+%%% Use protect on footnotes to avoid problems with footnotes in titles
+\let\rmarkdownfootnote\footnote%
+\def\footnote{\protect\rmarkdownfootnote}
 
-Omics technologies that generate a large amount of molecular data characterizing patient biospecimens have the potential to provide information about patients' disease characteristics above and beyond standard clinical and pathological features.
-By combining the information from a large number of molecular features into a multivariable model or decision algorithm, called a biomarker signature, there is the opportunity to identify distinct subgroups of patients for whom treatment decisions can be personalized.
-A biomarker signature can guide decisions to treat or not to treat and help identify the patients who are most likely to have a more favororable outcome or benefit from a particular therapy.
-The key challenge is to combine features from a high dimensional molecular assay to derive a signature with good clinical performance and appropriately characterize its performance.
-The inappropriate practice of using overlapping data to both build a signature and evaluate its performance can lead to severe over-optimism bias in performance estimates.
-We summarize the key statistical issues and methods for developing and validating biomarker signatures, using examples from the literature for illustration.
+%%% Change title format to be more compact
+\usepackage{titling}
 
-# Introduction
+% Create subtitle command for use in maketitle
+\newcommand{\subtitle}[1]{
+  \posttitle{
+    \begin{center}\large#1\end{center}
+    }
+}
 
-```{r setup, include = FALSE}
-library(dplyr)
-library(ggplot2)
-library(knitr)
-library(survival)
-library(ggkm)
-library(tidyr)
-library(stargazer)
+\setlength{\droptitle}{-2em}
+  \title{Issues in developing multivariable molecular signatures for guiding
+clinical care decisions}
+  \pretitle{\vspace{\droptitle}\centering\huge}
+  \posttitle{\par}
+  \author{Michael C Sachs and Lisa M McShane}
+  \preauthor{\centering\large\emph}
+  \postauthor{\par}
+  \predate{\centering\large\emph}
+  \postdate{\par}
+  \date{May 18, 2016}
 
-opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE, warning = FALSE)
-knit_hooks$set(plot = function(x, options) {
-  if ('tikz' %in% options$dev && !options$external) {
-    hook_plot_tex(x, options)
-  } else hook_plot_md(x, options)
-})
-```
+\usepackage{setspace}
+\doublespacing
+\usepackage{lineno}
+\linenumbers
 
-Omics technologies that generate a large amount of molecular data characterizing patient biospecimens have the potential to provide information about the patients' disease characteristics above and beyond standard clinical and pathological features.
-By combining the information from a large number of molecular features into a multivariable model, hereafter referred to as a _biomarker signature_, there is the opportunity to identify distinct subgroups of patients for whom clinical management decisions can be personalized.
-In early-stage disease, for example, a highly prognostic signature might identify a subpopulation of good risk patients that has such a high probability of long survival or disease-free survival that they do not require additional treatment beyond some standard base therapy.
-Therefore these good risk patients can be spared the risks and side-effects associated with additional therapy.
-In the context of a specific therapy that targets a particular molecular pathway, a signature may identify a subpopulation of patients that does or does not benefit from that therapy, thereby guiding the decision of whether or not to administer the therapy.
+\begin{document}
 
-The key challenge we address in this paper is how to combine feature mneasurements generated by a high dimensional multiplex molecular assay to derive a signature that is fit for a specified clinical and additionally provide valid estimates of its performance characteristics.
-A signature that has clinical value must have statistical performance (as quantified by one or more metrics) that is fit for the clinical context, and it must provide information that can be acted upon clinically to provide some benefit to a patient.
-For example, a biomarker signature that is well calibrated may accurately predict a clinical outcome, but that does not neccessarily imply that the prediction provides clinically useful information; the signature may divide patients into subgroups with different prognosis, but if there are no therapies available to improve the outcome of either subgroup, then the signature may have little clinical value.
-While the focus here is statistical estimation of a numerical performance metric, it should be kept in mind that clinical usefulness may depend on a variety of additional factors.
+\maketitle
 
-## Terminology and Notation
+\section{Abstract}\label{abstract}
 
-A **biomarker signature** is a transformation of multiple individual feature measurements, typically molecular characteristics measured on a multiplex assay, to a one-dimensional space.
-Specifically, let $X$ denote the set of $p$ feature measurements under consideration.
-The signature is an unknown function $f(X): \mathbb{R}^p \mapsto \mathbb{R}^1$.
-The signature result may be continuous, take multiple discrete values, or be dichotomous.
-In principle, the signature could also depend on other clinical or pathological variables in addition to the molecular measurements, but to simplify the discussion we will focus on signatures that depend on molecular feature data only.
-Let $S$ denote the development dataset which includes, for each of $n$ represented individuals, a feature vector $X$, an outcome $Y$, and a treatment $Z$.
-$S$ is a sample of size $n$ from distribution $\mathcal{P}$ with domain $\mathcal{S}$.
-Let $\mathcal{F}$ be a mapping from $\mathcal{S}$ to the space of continuous functions, $\mathcal{D}$, with domain $\mathbb{R}^p$ and range $\mathbb{R}$.
-Thus $\mathcal{F}: \mathcal{S} \mapsto \mathcal{D}$ denotes the process or algorithm through which a particular $f$ is estimated.
-We do not place any other restrictions on $\mathcal{F}$, it could be a clustering approach, a regression approach, a combination of both, or something else entirely.
-We will use $\mathcal{F}$ to denote the manner in which $f$ is estimated and will write $f \in \mathcal{F}$ to denote that $f$ is estimated with the class of methods $\mathcal{F}$.
+Omics technologies that generate a large amount of molecular data
+characterizing patient biospecimens have the potential to provide
+information about patients' disease characteristics above and beyond
+standard clinical and pathological features. By combining the
+information from a large number of molecular features into a
+multivariable model or decision algorithm, called a biomarker signature,
+there is the opportunity to identify distinct subgroups of patients for
+whom treatment decisions can be personalized. A biomarker signature can
+guide decisions to treat or not to treat and help identify the patients
+who are most likely to have a more favororable outcome or benefit from a
+particular therapy. The key challenge is to combine features from a high
+dimensional molecular assay to derive a signature with good clinical
+performance and appropriately characterize its performance. The
+inappropriate practice of using overlapping data to both build a
+signature and evaluate its performance can lead to severe over-optimism
+bias in performance estimates. We summarize the key statistical issues
+and methods for developing and validating biomarker signatures, using
+examples from the literature for illustration.
 
-Let $\phi: \mathcal{D} \times \mathcal{S} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC).
-This could also be a measure of association, such as an odds ratio, hazard ratio, or log-rank statistic.
-This is a function of both $f$ and $S$.
-We are interested in estimating $E_\mathcal{P}[\phi_{f^*}(S)]$, which is the expected error under the data generation mechanism (distribution $\mathcal{P}$), for a particular $f^* \in \mathcal{F}$.
-This allows us to understand how the signature will perform on future observations generated from $\mathcal{P}$.
-We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ **for all** $f \in \mathcal{F}$, which is the generalization performance for $f$ generated using mechanism $\mathcal{F}$.
-This doesn't guide outside researchers as to which specific $f$ to use, yet it is useful for development because it tells us how much signal is in the data.
-As shorthand we will write this as $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
+\section{Introduction}\label{introduction}
 
-## Overview of biomarker signature development
+Omics technologies that generate a large amount of molecular data
+characterizing patient biospecimens have the potential to provide
+information about the patients' disease characteristics above and beyond
+standard clinical and pathological features. By combining the
+information from a large number of molecular features into a
+multivariable model, hereafter referred to as a \emph{biomarker
+signature}, there is the opportunity to identify distinct subgroups of
+patients for whom clinical management decisions can be personalized. In
+early-stage disease, for example, a highly prognostic signature might
+identify a subpopulation of good risk patients that has such a high
+probability of long survival or disease-free survival that they do not
+require additional treatment beyond some standard base therapy.
+Therefore these good risk patients can be spared the risks and
+side-effects associated with additional therapy. In the context of a
+specific therapy that targets a particular molecular pathway, a
+signature may identify a subpopulation of patients that does or does not
+benefit from that therapy, thereby guiding the decision of whether or
+not to administer the therapy.
 
-The inherent statistical challenge is how to both develop a signature $f \in \mathcal{F}$ and obtain a valid assessment of its performance.
-Additionally, one should provide a specification of $f$ for others to use.
-Typically, a specific $f$ is estimated using $\mathcal{F}$ based on some training data.
-The class of methods $\mathcal{F}$ can comprise a variety of different computational approaches.
-In recent years there has been an explosion in the literature of computational approaches to classification and prediction, and we do not intend to summarize them all here.
-Some excellent reviews are provided by @hastie2009elements and @moons2012riskI.
-The main considerations in signature estimation are identifying the features to include, deciding what transformations to apply to the feature measurements, determining how to combine the feature measurements, and determining what, if any, thresholds or cutoffs should be applied to the resulting signature value.
+The key challenge we address in this paper is how to combine feature
+mneasurements generated by a high dimensional multiplex molecular assay
+to derive a signature that is fit for a specified clinical and
+additionally provide valid estimates of its performance characteristics.
+A signature that has clinical value must have statistical performance
+(as quantified by one or more metrics) that is fit for the clinical
+context, and it must provide information that can be acted upon
+clinically to provide some benefit to a patient. For example, a
+biomarker signature that is well calibrated may accurately predict a
+clinical outcome, but that does not neccessarily imply that the
+prediction provides clinically useful information; the signature may
+divide patients into subgroups with different prognosis, but if there
+are no therapies available to improve the outcome of either subgroup,
+then the signature may have little clinical value. While the focus here
+is statistical estimation of a numerical performance metric, it should
+be kept in mind that clinical usefulness may depend on a variety of
+additional factors.
 
-Our focus here will be mainly on the development of signatures using supervised methods, meaning methods that use information on an outcome variable it is desired to predict during the signature development process.
-Regression methods are a commonly used class of supervised methods.
-Oncotype DX [@paik2004multigene] is an example of a prognostic signature used for clinical decision making in early stage breast cancer that was developed using supervised methods.
-In the case of Oncotype DX, the outcome $Y$ was time to distant recurrence of breast cancer and the feature data were gene expression values measured in breast tumor specimens.
-A common supervised approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment.
-Supervised methods are contrasted with unsupervised methods which use only the feature data to derive a signature.
-An example of a signature developed by unsupervised methods is the signature that identifies biological subclasses of diffuse large B-cell lymphoma, which were originally developed using clustering methods and subsequently were found to be associated with prognosis [@alizadeh2000distinct].
-It is possible, and quite common in high-dimensional data settings, to combine multiple approaches when estimated a signature.
-For instance, a data-reduction step by variable selection or clustering may be performed before doing regression analysis on the resulting components.
-A review by Subramanian and Simon identified a large collection of gene expression–based prognostic signatures in lung cancer which had been developed using a wide variety of methods [@subramanian2010gene].
+\subsection{Terminology and Notation}\label{terminology-and-notation}
 
-The focus of this paper is how to obtain a valid estimate of signature performance as quantified by a metric $phi$, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
-The principles discussed apply no matter what signature development method is used.
-Performance depends on the true signal in the data and the specific algorithm $\mathcal{F}$ used for development.
-The performance will reflect that expected of signatures developed applying the prescribed methods to data drawn from the distribution $\mathcal{P}$.
-For signature deployment, one would also want a particular specification of $f$ for others to use on independent data (e.g., in clinical practice).
-It is not trivial to conduct this evaluation in a valid manner when there is a limited data set available to both define a signature and assess its performance.
-We illustrate the potential for bias to creep into the performance estimation and review some strategies to avoid the pitfalls that lead to these biases.
-We evaluate various strategies through a simulation study and compare the performance of some of those strategies using a real data exmaple involving the development of a signature for use in making treatment decisions for patients with lung cancer.
+A \textbf{biomarker signature} is a transformation of multiple
+individual feature measurements, typically molecular characteristics
+measured on a multiplex assay, to a one-dimensional space. Specifically,
+let \(X\) denote the set of \(p\) feature measurements under
+consideration. The signature is an unknown function
+\(f(X): \mathbb{R}^p \mapsto \mathbb{R}^1\). The signature result may be
+continuous, take multiple discrete values, or be dichotomous. In
+principle, the signature could also depend on other clinical or
+pathological variables in addition to the molecular measurements, but to
+simplify the discussion we will focus on signatures that depend on
+molecular feature data only. Let \(S\) denote the development dataset
+which includes, for each of \(n\) represented individuals, a feature
+vector \(X\), an outcome \(Y\), and a treatment \(Z\). \(S\) is a sample
+of size \(n\) from distribution \(\mathcal{P}\) with domain
+\(\mathcal{S}\). Let \(\mathcal{F}\) be a mapping from \(\mathcal{S}\)
+to the space of continuous functions, \(\mathcal{D}\), with domain
+\(\mathbb{R}^p\) and range \(\mathbb{R}\). Thus
+\(\mathcal{F}: \mathcal{S} \mapsto \mathcal{D}\) denotes the process or
+algorithm through which a particular \(f\) is estimated. We do not place
+any other restrictions on \(\mathcal{F}\), it could be a clustering
+approach, a regression approach, a combination of both, or something
+else entirely. We will use \(\mathcal{F}\) to denote the manner in which
+\(f\) is estimated and will write \(f \in \mathcal{F}\) to denote that
+\(f\) is estimated with the class of methods \(\mathcal{F}\).
 
-# Issues
+Let \(\phi: \mathcal{D} \times \mathcal{S} \mapsto \mathbb{R}\) denote
+the statistic that quantifies the performance of the function \(f\),
+such as predictive accuracy, mean squared error, or area under the
+receiver operating characteristic (ROC) curve (AUC). This could also be
+a measure of association, such as an odds ratio, hazard ratio, or
+log-rank statistic. This is a function of both \(f\) and \(S\). We are
+interested in estimating \(E_\mathcal{P}[\phi_{f^*}(S)]\), which is the
+expected error under the data generation mechanism (distribution
+\(\mathcal{P}\)), for a particular \(f^* \in \mathcal{F}\). This allows
+us to understand how the signature will perform on future observations
+generated from \(\mathcal{P}\). We may also be interested in estimating
+\(E_\mathcal{P}[\phi_f(S)]\) \textbf{for all} \(f \in \mathcal{F}\),
+which is the generalization performance for \(f\) generated using
+mechanism \(\mathcal{F}\). This doesn't guide outside researchers as to
+which specific \(f\) to use, yet it is useful for development because it
+tells us how much signal is in the data. As shorthand we will write this
+as \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\).
 
-The main goal of interest here is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected value of a given performance metric on future observations for $f \in \mathcal{F}$.
-This can be estimated with the in-sample empirical estimate: $\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)$ for a particular $f$.
-However, if $S$ is used to estimate $f$ then the estimate will be biased due to overfitting, that is, $|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|$ will be large.
-Overfitting occurs when a model is fit to noise in the data.
-This often occurs when fitting a model that is overly complex relative to the amount of signal in the available data.
-Overfitting bias results from the fact that $\phi_f(S)$ depends on $f$ which depends on $S$, and thus the statistic $\phi_f(S)$ is being adaptively defined based on the observed data $S$.
-Any estimate of signature performance is potentially subject to bias if overfitting occurs, but such biases frequently go unrecognized in the medical literature.
+\subsection{Overview of biomarker signature
+development}\label{overview-of-biomarker-signature-development}
 
-Typically a performance metric $\phi$ used during signature development is degined to quantify either calibration or discrimination, or some combination of the two.
-Calibration assesses how well the signature based predictions compare to the observed outcomes.
-Discrimination assesses how well the signature distinguishes between groups of patients who do or do not experience an event.
-To assess discrimination, a different metric $\phi$ may be used, such as the area under the ROC curve.
-Measures of association such as the odds ratio, hazard ratio, or difference in survival probabilities are also commonly used performance metrics, although their value for biomarker signatures has been debated [@pepe2004limitations].
-In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups.
-The risk groups were determined by the signature $f$, which was estimated using the JBL.10 observation arm data.
-The signature was then applied to the same data that were used to build it to define risk groups, producing a the hazard ratio estimate of 15 for a prognostic effect in the observation arm.
-Subsequent evaluations of the signature's performance on independent data resulted in hazard ratio estimates in the 2 - 3 range, substantially smaller than the original estimate of 15.
-Next we reanalyze the JBR.10 data and illustrate some methods to avoid bias in evaluation of signature performance.
+The inherent statistical challenge is how to both develop a signature
+\(f \in \mathcal{F}\) and obtain a valid assessment of its performance.
+Additionally, one should provide a specification of \(f\) for others to
+use. Typically, a specific \(f\) is estimated using \(\mathcal{F}\)
+based on some training data. The class of methods \(\mathcal{F}\) can
+comprise a variety of different computational approaches. In recent
+years there has been an explosion in the literature of computational
+approaches to classification and prediction, and we do not intend to
+summarize them all here. Some excellent reviews are provided by
+\citet{hastie2009elements} and \citet{moons2012riskI}. The main
+considerations in signature estimation are identifying the features to
+include, deciding what transformations to apply to the feature
+measurements, determining how to combine the feature measurements, and
+determining what, if any, thresholds or cutoffs should be applied to the
+resulting signature value.
 
-## Avoiding Overfitting
+Our focus here will be mainly on the development of signatures using
+supervised methods, meaning methods that use information on an outcome
+variable it is desired to predict during the signature development
+process. Regression methods are a commonly used class of supervised
+methods. Oncotype DX \citep{paik2004multigene} is an example of a
+prognostic signature used for clinical decision making in early stage
+breast cancer that was developed using supervised methods. In the case
+of Oncotype DX, the outcome \(Y\) was time to distant recurrence of
+breast cancer and the feature data were gene expression values measured
+in breast tumor specimens. A common supervised approach to identifying
+treatment-selection signatures is to use regression techniques to
+estimate a signature that has a strong interaction with a particular
+treatment. Supervised methods are contrasted with unsupervised methods
+which use only the feature data to derive a signature. An example of a
+signature developed by unsupervised methods is the signature that
+identifies biological subclasses of diffuse large B-cell lymphoma, which
+were originally developed using clustering methods and subsequently were
+found to be associated with prognosis \citep{alizadeh2000distinct}. It
+is possible, and quite common in high-dimensional data settings, to
+combine multiple approaches when estimated a signature. For instance, a
+data-reduction step by variable selection or clustering may be performed
+before doing regression analysis on the resulting components. A review
+by Subramanian and Simon identified a large collection of gene
+expression--based prognostic signatures in lung cancer which had been
+developed using a wide variety of methods \citep{subramanian2010gene}.
+
+The focus of this paper is how to obtain a valid estimate of signature
+performance as quantified by a metric \(phi\), that is, a good estimate
+of \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\). The principles discussed
+apply no matter what signature development method is used. Performance
+depends on the true signal in the data and the specific algorithm
+\(\mathcal{F}\) used for development. The performance will reflect that
+expected of signatures developed applying the prescribed methods to data
+drawn from the distribution \(\mathcal{P}\). For signature deployment,
+one would also want a particular specification of \(f\) for others to
+use on independent data (e.g., in clinical practice). It is not trivial
+to conduct this evaluation in a valid manner when there is a limited
+data set available to both define a signature and assess its
+performance. We illustrate the potential for bias to creep into the
+performance estimation and review some strategies to avoid the pitfalls
+that lead to these biases. We evaluate various strategies through a
+simulation study and compare the performance of some of those strategies
+using a real data exmaple involving the development of a signature for
+use in making treatment decisions for patients with lung cancer.
+
+\section{Issues}\label{issues}
+
+The main goal of interest here is to estimate
+\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\), the expected value of a given
+performance metric on future observations for \(f \in \mathcal{F}\).
+This can be estimated with the in-sample empirical estimate:
+\(\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)\) for a
+particular \(f\). However, if \(S\) is used to estimate \(f\) then the
+estimate will be biased due to overfitting, that is,
+\(|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|\) will be large.
+Overfitting occurs when a model is fit to noise in the data. This often
+occurs when fitting a model that is overly complex relative to the
+amount of signal in the available data. Overfitting bias results from
+the fact that \(\phi_f(S)\) depends on \(f\) which depends on \(S\), and
+thus the statistic \(\phi_f(S)\) is being adaptively defined based on
+the observed data \(S\). Any estimate of signature performance is
+potentially subject to bias if overfitting occurs, but such biases
+frequently go unrecognized in the medical literature.
+
+Typically a performance metric \(\phi\) used during signature
+development is degined to quantify either calibration or discrimination,
+or some combination of the two. Calibration assesses how well the
+signature based predictions compare to the observed outcomes.
+Discrimination assesses how well the signature distinguishes between
+groups of patients who do or do not experience an event. To assess
+discrimination, a different metric \(\phi\) may be used, such as the
+area under the ROC curve. Measures of association such as the odds
+ratio, hazard ratio, or difference in survival probabilities are also
+commonly used performance metrics, although their value for biomarker
+signatures has been debated \citep{pepe2004limitations}. In
+\citet{zhu2010prognostic}, \(\phi\) was the hazard ratio comparing the
+high risk and low risk groups. The risk groups were determined by the
+signature \(f\), which was estimated using the JBL.10 observation arm
+data. The signature was then applied to the same data that were used to
+build it to define risk groups, producing a the hazard ratio estimate of
+15 for a prognostic effect in the observation arm. Subsequent
+evaluations of the signature's performance on independent data resulted
+in hazard ratio estimates in the 2 - 3 range, substantially smaller than
+the original estimate of 15. Next we reanalyze the JBR.10 data and
+illustrate some methods to avoid bias in evaluation of signature
+performance.
+
+\subsection{Avoiding Overfitting}\label{avoiding-overfitting}
 
 A traditional method to avoid overfitting is the split sample approach.
-First, randomly partition $S$ into the training sample $S_t$ and the holdout sample $S_h$ with sample sizes $n_t$ and $n_h$, respectively.
-Then, $S_h$ is hidden from the analyst while $\mathcal{F}$ is applied to $S_t$ to estimate the signature function $f_t$.
-For fixed $f_t$, $\hat{E}[\phi_{f_t}(S_h)]$ is an unbiased estimator of $E_{\mathcal{P}}[\phi_{f_t}(S)]$.
-@dobbin2011optimally investigate how to optimally split a dataset into training and holdout partitions.
-The specific form of $f_t$ that is fixed using $S_t$ can be reported as the function for others to use, therefore the aforementioned estimator applies for that specific $f_t$.
-The drawback of the split-sample approach is that the performance of $f_t$ is likely to be inferior to the performance of a function $f^*$ derived using the same approach applied to the entire data set.
-@dobbin2007sample investigate how to choose the sample size n so that there is high probability that the signature developed will have expected performance within some specified margin of the the optimal signature that could be developed if the development data set had infinite sample size.
+First, randomly partition \(S\) into the training sample \(S_t\) and the
+holdout sample \(S_h\) with sample sizes \(n_t\) and \(n_h\),
+respectively. Then, \(S_h\) is hidden from the analyst while
+\(\mathcal{F}\) is applied to \(S_t\) to estimate the signature function
+\(f_t\). For fixed \(f_t\), \(\hat{E}[\phi_{f_t}(S_h)]\) is an unbiased
+estimator of \(E_{\mathcal{P}}[\phi_{f_t}(S)]\).
+\citet{dobbin2011optimally} investigate how to optimally split a dataset
+into training and holdout partitions. The specific form of \(f_t\) that
+is fixed using \(S_t\) can be reported as the function for others to
+use, therefore the aforementioned estimator applies for that specific
+\(f_t\). The drawback of the split-sample approach is that the
+performance of \(f_t\) is likely to be inferior to the performance of a
+function \(f^*\) derived using the same approach applied to the entire
+data set. \citet{dobbin2007sample} investigate how to choose the sample
+size n so that there is high probability that the signature developed
+will have expected performance within some specified margin of the the
+optimal signature that could be developed if the development data set
+had infinite sample size.
 
-Another approach to avoid overfitting is cross-validation, which is a resampling based approach.
-For a fixed integer $k$, which can be between 1 and $n$, we randomly partition the full data set $S$ into subgroups of size $k$ (assume for simplicity here that $n$ is evenly divisible by $k$ so that $n/k$ subgroups are formed).
-For each $k$, $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the subset of $S$ that is disjoint from $S_k$.
-Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
-This process is repeated $K$ times to yield $K$ estimates and then these are averaged to obtain the performance estimate.
-This process is called "leave $k$ out" or "$n/k$ fold" cross-validation.
-Note that for each subgroup in the partition, we obtain a new estimate of $f$, therefore we are only estimating $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$ and not the performance for a single specified signature.
-Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$ to yield $f^*$ as above.
-Importantly one should report the cross-validated estimate just described and not the naive esitimate that would be obtained by substituting the full data $S$ into the performance metric calculation for the signature $f^*$.
-The problematic aspects of such naive "resubstitution estimates" of performance are discuss in more detail below.
+Another approach to avoid overfitting is cross-validation, which is a
+resampling based approach. For a fixed integer \(k\), which can be
+between 1 and \(n\), we randomly partition the full data set \(S\) into
+subgroups of size \(k\) (assume for simplicity here that \(n\) is evenly
+divisible by \(k\) so that \(n/k\) subgroups are formed). For each
+\(k\), \(f_{-k}\) is estimated and fixed by appling \(\mathcal{F}\) on
+\(S_{-k}\) which is the subset of \(S\) that is disjoint from \(S_k\).
+Then, we get an estimate \(\hat{E}[\phi_{f_{-k}}(S_k)]\) which is an
+estimate of \(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). This process is
+repeated \(K\) times to yield \(K\) estimates and then these are
+averaged to obtain the performance estimate. This process is called
+``leave \(k\) out'' or ``\(n/k\) fold'' cross-validation. Note that for
+each subgroup in the partition, we obtain a new estimate of \(f\),
+therefore we are only estimating
+\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\) and not the performance for a
+single specified signature. Typically, if a specific form for \(f\) is
+desired, it would be estimated using the entire dataset \(S\) to yield
+\(f^*\) as above. Importantly one should report the cross-validated
+estimate just described and not the naive esitimate that would be
+obtained by substituting the full data \(S\) into the performance metric
+calculation for the signature \(f^*\). The problematic aspects of such
+naive ``resubstitution estimates'' of performance are discuss in more
+detail below.
 
-A variation on the cross-validation approach is bootstrapping.
-In that case, a sample $S_b$ of size $n$ is sampled _with replacement_ from $S$.
-Then $f_{b}$ is estimated and fixed by applying $\mathcal{F}$ to $S_{b}$.
-The performance metric $\phi$ is calculated on the subset of $S$ that is disjoint from $S_b$, which we denote by $S_{-b}$, to yield an estimate of $E[\phi_{f_b}(S_{-b})]$.
-This process is repeated $K$ times to yield $K$ estimates.
-These $K$ performance metric estimates are averaged to obtain the mean over the bootstrap replicates.
-@efron1997improvements suggest a variation, the 0.632 estimate:
-$$
+A variation on the cross-validation approach is bootstrapping. In that
+case, a sample \(S_b\) of size \(n\) is sampled \emph{with replacement}
+from \(S\). Then \(f_{b}\) is estimated and fixed by applying
+\(\mathcal{F}\) to \(S_{b}\). The performance metric \(\phi\) is
+calculated on the subset of \(S\) that is disjoint from \(S_b\), which
+we denote by \(S_{-b}\), to yield an estimate of
+\(E[\phi_{f_b}(S_{-b})]\). This process is repeated \(K\) times to yield
+\(K\) estimates. These \(K\) performance metric estimates are averaged
+to obtain the mean over the bootstrap replicates.
+\citet{efron1997improvements} suggest a variation, the 0.632 estimate:
+\[
 \hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
-$$
-where $\hat{E}[\phi_{f}(S)]$ is the naive estimate of $\phi_f$ using the entire dataset.
+\] where \(\hat{E}[\phi_{f}(S)]\) is the naive estimate of \(\phi_f\)
+using the entire dataset.
 
-Another variation on all of these methods is the concept of pre-validation [@tibshirani2002pre].
-With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{-b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted signature $\hat{f}(X_i)$ is estimated for $X_i \in S_{-b}$ where $\hat{f}$ is estimated using $S_{b}$.
-This process is repeated to obtain a set of pre-validated signature estimates $\hat{f}$ which are then used to calculate $\phi$.
-For single-step split sample training and validation, this process is equivalent to what is described above.
-For cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on each of the smaller held-out datasets.
+Another variation on all of these methods is the concept of
+pre-validation \citep{tibshirani2002pre}. With pre-validation, instead
+of computing the statistic \(\phi\) for each of the held-out subsets
+(\(S_{-b}\) for the bootstrap or \(S_{k}\) for cross-validation), the
+fitted signature \(\hat{f}(X_i)\) is estimated for \(X_i \in S_{-b}\)
+where \(\hat{f}\) is estimated using \(S_{b}\). This process is repeated
+to obtain a set of pre-validated signature estimates \(\hat{f}\) which
+are then used to calculate \(\phi\). For single-step split sample
+training and validation, this process is equivalent to what is described
+above. For cross-validation and the bootstrap, this process avoids the
+problem of having too few cases to estimate the statistic \(\phi\) on
+each of the smaller held-out datasets.
 
-## Simulation Study
+\subsection{Simulation Study}\label{simulation-study}
 
@@
 
-To illustrate some of the different properties of these estimates and how they help to avoid overfitting, we conduct a limited simulation study.
-Data were generated with 1000 observations, each with a binary outcome $Y$ with prevalence 0.3, and 500 mutually independent features sampled from the standard normal distribution.
-This is the null case where no features are associated with $Y$.
-The signature development procedure entails a feature selection step, in which each feature is regressed against $Y$ in a univariate logistic regression model.
-The 25 features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines the final signature.
+To illustrate some of the different properties of these estimates and
+how they help to avoid overfitting, we conduct a limited simulation
+study. Data were generated with 1000 observations, each with a binary
+outcome \(Y\) with prevalence 0.3, and 500 mutually independent features
+sampled from the standard normal distribution. This is the null case
+where no features are associated with \(Y\). The signature development
+procedure entails a feature selection step, in which each feature is
+regressed against \(Y\) in a univariate logistic regression model. The
+25 features with the smallest p-values are selected for inclusion in a
+multivariable logistic regression model which defines the final
+signature.
 
-We compare each of the methods described above: split-sample holdout, cross-validation, bootstrap, and pre-validation, along with several commonly used but biased approaches.
-The biased approaches are described in Table \ref{descript}.
-Two of the biased approaches use the full sample to select the features, followed by fitting the multivariable model on the holdout subset.
-This is referred to as "parital holdout" or "partial CV" when using split-sample holdout or cross-validation as the validation step, respectively.
-We also implemented the naïve resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is developed on a training subset but then evaluated on the combined training adn holdout data sets.
-Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
-In our simulation, we look at two different performance metrics, the area under the ROC curve (AUC) and the odds ratio for the outcome comparing the signature groups.
+We compare each of the methods described above: split-sample holdout,
+cross-validation, bootstrap, and pre-validation, along with several
+commonly used but biased approaches. The biased approaches are described
+in Table \ref{descript}. Two of the biased approaches use the full
+sample to select the features, followed by fitting the multivariable
+model on the holdout subset. This is referred to as ``parital holdout''
+or ``partial CV'' when using split-sample holdout or cross-validation as
+the validation step, respectively. We also implemented the naïve
+resubstitution approach, wherein the model is trained and evaluated on
+the same dataset, and the partial resubstitution approach wherein the
+model is developed on a training subset but then evaluated on the
+combined training adn holdout data sets. Our main interest is in
+comparing the bias and variance of the resulting estimates of
+\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). In our simulation, we look
+at two different performance metrics, the area under the ROC curve (AUC)
+and the odds ratio for the outcome comparing the signature groups.
 
-```{r cvsims, fig.cap = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) in the setting where a dataset is used to both develop the signature and evaluate its performance.
-The violin plots show mirrored density estimates for the AUC for 1000 replicates of the numerical experiment.
-In each replicate, there are 1000 observations and 500 features.
-The true value of the AUC is 0.5.
-CV = Cross validation.
-\\label{fig1}"}
-load("../Code/cvsim-result.RData")
-cvres <- do.call("rbind", cvsims)
-cvlong <- do.call("rbind", lapply(1:ncol(cvres), function(i){
+\begin{figure}[htbp]
+\centering
+\includegraphics{paper-revised_files/figure-latex/cvsims-1.pdf}
+\caption{Comparison of different approaches to estimating the Area Under
+the ROC Curve (AUC) in the setting where a dataset is used to both
+develop the signature and evaluate its performance. The violin plots
+show mirrored density estimates for the AUC for 1000 replicates of the
+numerical experiment. In each replicate, there are 1000 observations and
+500 features. The true value of the AUC is 0.5. CV = Cross validation.
+\label{fig1}}
+\end{figure}
 
-  cl <- cvres[, i]
-  splow <- strsplit(colnames(cvres)[i], ".", fixed = TRUE)
-  nm <- paste(unlist(sapply(splow, function(s) rev(rev(s)[-1]))), collapse = ".")
-  cls <- sapply(splow, function(s) rev(s)[1])
-  data.frame(value = cvres[, i], stat = cls, scen = nm, stringsAsFactors = FALSE)
+\begin{figure}[htbp]
+\centering
+\includegraphics{paper-revised_files/figure-latex/cvsims2-1.pdf}
+\caption{Comparison of different approaches to estimating the odds ratio
+(OR) in the setting where a dataset is used to both develop the
+signature and evaluate its performance. The violin plots show mirrored
+density estimates for the log OR for 1000 replicates of the numerical
+experiment. In each replicate, there are 1000 observations and 500
+features. The true value of the OR is 1.0. CV = Cross validation.
+\label{fig2}}
+\end{figure}
 
-}))
-cvlong$value[cvlong$stat == "OR"] <- exp(cvlong$value[cvlong$stat == "OR"])
+\begin{longtable}[c]{@{}lrrrrrr@{}}
+\caption{Comparison of different approaches to estimating the Area Under
+the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a
+dataset is used to both develop the signature and evaluate its
+performance. The true value of the AUC is 0.5 and the true value of the
+Log OR is 0.0. Estimates are based on 1000 replicates of the numerical
+experiment. In each replicate, there are 1000 observations and 500
+features. CV = Cross validation.}\tabularnewline
+\toprule
+Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR &
+Bias OR\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR &
+Bias OR\tabularnewline
+\midrule
+\endhead
+Resubstitution & 0.72 & 0.01 & 0.22 & 1.33 & 0.12 & 1.33\tabularnewline
+Partial CV & 0.68 & 0.01 & 0.18 & 1.02 & 0.12 & 1.02\tabularnewline
+Partial Holdout & 0.67 & 0.02 & 0.17 & 0.97 & 0.19 & 0.97\tabularnewline
+Partial Resubstitution & 0.65 & 0.02 & 0.15 & 0.88 & 0.12 &
+0.88\tabularnewline
+Pre-validation & 0.50 & 0.05 & 0.00 & -0.01 & 0.32 &
+-0.01\tabularnewline
+Leave 10 out CV & 0.50 & 0.04 & 0.00 & 0.00 & 0.25 & 0.00\tabularnewline
+Leave 100 out CV & 0.50 & 0.03 & 0.00 & 0.00 & 0.20 &
+0.00\tabularnewline
+30\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.24 & 0.00\tabularnewline
+50\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.20 & 0.00\tabularnewline
+Bootstrap & 0.50 & 0.01 & 0.00 & 0.00 & 0.08 & 0.00\tabularnewline
+\bottomrule
+\end{longtable}
 
-cvlong$scen <- factor(cvlong$scen, levels = c("naive", "zhu.cv", "zhu.hold", "zhu.hold2", 
-                                              "cv.preval", "cv.10", "cv.100", "holdout.3", 
-                                              "holdout.5", "boot"), ordered = TRUE)
+Not surprisingly, the resubstitution estimates are optimistically
+biased: the naive resubstitution estimate of the AUC is 44\% larger than
+it should be and the log OR estimate is over 2 times higher than it
+should be, on average. Partial resubstitution, partial holdout, and
+partial cross-validation estimates do not ameliorate the bias very much.
+Investigators are tempted to use partial holdout estimates as it is more
+convenient to not have to carry along a large number of feature
+measurements into the modeling part of the signature development
+process, and they may think that the resulting performance estimates are
+still close to valid, as only half of the data are used to form the
+estimates. However, here we see that these versions are still severely
+biased and should not be reported as valid assessments of the
+performance of biomarker signatures.
 
-labs <- c("Resubstitution", "Partial CV", 
-  "Partial Holdout", "Partial Resubstitution", 
-  "Pre-validation", "Leave 10 out CV", "Leave 100 out CV", 
-  "30% Holdout", "50% Holdout", "Bootstrap")
+The split-sample holdout, cross-validation, bootstrap, and
+pre-validation methods all produce esentially unbiased estimates in the
+simulated example, with their mean AUCs being nearly 0.5 and the mean
+ORs being nearly 1. We can compare the spread of the distributions to
+get a sense of the differences in precision of the estimates. The
+bootstrap approach appears to be the most precise, followed by the
+cross-validation, holdout, and finally the pre-validation. The
+bootstrap, as intended, is a more efficient, smoothed version of the
+cross validation estimate (Figure \ref{fig1}, \ref{fig2}). It provides
+the best balance between allocating data to train the signature and
+having independent data remaining to precisely estimate the statistic
+\(\phi\).
 
-names(labs) <- levels(cvlong$scen)
+\subsection{Data Analysis}\label{data-analysis}
 
-ggplot(subset(cvlong, stat == "AUC"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") + 
-  theme_bw(base_size = 13, base_family = "serif") + coord_flip() + 
-  scale_x_discrete("Estimation Approach", labels = labs) + 
-  ylab("Area Under the ROC Curve")
-```
+\subsection{Data analysis example}\label{data-analysis-example}
 
-```{r cvsims2, fig.cap = "Comparison of different approaches to estimating the odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance.
-The violin plots show mirrored density estimates for the log OR for 1000 replicates of the numerical experiment.
-In each replicate, there are 1000 observations and 500 features.
-The true value of the OR is 1.0.
-CV = Cross validation.
-\\label{fig2} "}
+We now illustrate the concepts and methods just discussed by reanalyzing
+data that had been used to build a previously published lung cancer
+prognostic signature \citet{zhu2010prognostic}. Briefly, the data of
+interest are from the JBR.10 trial, which was a randomized controlled
+trial of the adjuvant chemotherapy regimen vinorelbine/cisplatin (ACT)
+versus observation alone (OBS) in 482 participants with non small cell
+lung cancer (NSCLC) who had undergone surgery. Of those 482
+participants, 169 had frozen tumor tissue collected, and of those tumor
+samples, 133 (71 in ACT and 62 in OBS) had gene-expression profiling
+performed using U133A oligonucleotide microarrays (Affymetrix, Santa
+Clara, CA).
 
-ggplot(subset(cvlong, stat == "OR"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") + 
-  theme_bw(base_size = 13, base_family = "serif") + coord_flip() + 
-  scale_x_discrete("Estimation Approach", labels = labs) + scale_y_log10(breaks = c(.5, 1, 2, 4)) +
-  ylab("Odds Ratio (log scale)")
-```
+The goal of the \citet{zhu2010prognostic} paper was to identify a
+multi-gene signature that strongly predicts prognosis, and the
+hypothesis was that the poor prognosis subgroup would benefit more from
+ACT than the good prognosis subgroup. The signature was developed on a
+training data set (``trained'') to predict disease specific survival.
+The annotated gene expression data and clinical information are
+available from the Gene Expression Omnibus database(identifier:
+GSE14814, \citet{edgar2002gene}).
 
-```{r cvtab}
+\citet{zhu2010prognostic} present results that mainly focus on the
+association of their signature with outcome, albeit a less than ideal
+measure of discrimination ability. They demonstrate that the two risk
+subgroups predicted by their signature (high risk and low risk) have
+separation in their survival curves and that the hazard ratio for their
+signature is large and significant even when adjusting for other risk
+factors. They do not directly address calibration, that is, whether
+their signature accurately predicts survival times.
 
-cvlong$value[cvlong$stat == "OR"] <- log(cvlong$value[cvlong$stat == "OR"])
+We used a similar approach to preprocessing as did
+\citet{zhu2010prognostic}, although we could not reproduce their
+workflow exactly due to outdated software. Batch effects were removed
+using the \texttt{ComBat} function in the \texttt{sva} \texttt{R}
+package \citep{leeksva} and then the gene expression values were
+centered by their means and scaled by their standard deviations. Our
+signature development approach is similar but not identical to that in
+\citet{zhu2010prognostic}. For purposes of illustrating the concepts we
+used a simplified approach to signature development that retains the
+main characteristics of the original method. The exact approach to
+signature development does not have a major impact on the main
+conclusions of our evaluation of the various approaches to signature
+performance assessment.
 
-cvlong$Approach <- factor(labs[cvlong$scen], levels = labs, ordered = TRUE)
-cvlong$Truth <- ifelse(cvlong$stat == "AUC", .5, 0.0)
-cvlong$value[!is.finite(cvlong$value)] <- NA
-cvwide <- cvlong %>% group_by(stat, Approach) %>%
-  summarize(`mean` = mean(value, na.rm =TRUE), `std.dev` = sd(value, na.rm = TRUE), 
-            `Bias` = mean(value - Truth[1], na.rm = TRUE))
-
-cvwide2 <- left_join(cvwide[1:10, -1], cvwide[11:20, -1], by = c("Approach"))
-colnames(cvwide2) <- gsub(".x", " AUC", colnames(cvwide2))
-colnames(cvwide2) <- gsub(".y", " OR", colnames(cvwide2))
-
-cvwide2 %>%
-  kable(digits = 2, caption = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance.
-The true value of the AUC is 0.5 and the true value of the Log OR is 0.0.
-Estimates are based on 1000 replicates of the numerical experiment.
-In each replicate, there are 1000 observations and 500 features.
-CV = Cross validation.
-")
-
-```
-
-Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the log OR estimate is over 2 times higher than it should be, on average.
-Partial resubstitution, partial holdout, and partial cross-validation estimates do not ameliorate the bias very much.
-Investigators are tempted to use partial holdout estimates as it is more convenient to not have to carry along a large number of feature measurements into the modeling part of the signature development process, and they may think that the resulting performance estimates are still close to valid, as only half of the data are used to form the estimates.
-However, here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.
-
-The split-sample holdout, cross-validation, bootstrap, and pre-validation methods all produce esentially unbiased estimates in the simulated example, with their mean AUCs being nearly 0.5 and the mean ORs being nearly 1.
-We can compare the spread of the distributions to get a sense of the differences in precision of the estimates.
-The bootstrap approach appears to be the most precise, followed by the cross-validation, holdout, and finally the pre-validation.
-The bootstrap, as intended, is a more efficient, smoothed version of the cross validation estimate (Figure \ref{fig1}, \ref{fig2}).
-It provides the best balance between allocating data to train the signature and having independent data remaining to precisely estimate the statistic $\phi$.
-
-## Data Analysis
-
-```{r examp1}
-source("../Code/03-new-signature-zhu.R")
-```
-
-```{r exampresub, cache = TRUE}
-## substitution estimate using full dataset
-obsgrp <- subset(ldat, Post.Surgical.Treatment == "OBS")
-obsgrpw <- subset(gdat, Post.Surgical.Treatment == "OBS")
-fit.all <- fit.superpc(obsgrp, cldat)
-fitted <- predict.superpc(gdat, fit.all)
-
-evalall <- eval.predict(obsgrpw, fit.all)
-
-cldat[, "riskgrp"] <- fitted$riskgrp
-cldat$DSS <- as.numeric(cldat$DSS.status == "Dead")
-#ggplot(subset(cldat, Post.Surgical.Treatment == "OBS"), 
-#       aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km() + 
-#  theme_bw(base_size = 13, base_family = "serif") + 
-#  scale_linetype_discrete("Resubstitution", labels = c("Low risk", "High risk")) + 
-#  ylab("Disease Specific Survival") + xlab("Time in years since randomization")
-
-```
-
-## Data analysis example
-
-We now illustrate the concepts and methods just discussed by reanalyzing data that had been used to build a previously published lung cancer prognostic signature @zhu2010prognostic.
-Briefly, the data of interest are from the JBR.10 trial, which was a randomized controlled trial of the adjuvant chemotherapy regimen vinorelbine/cisplatin (ACT) versus observation alone (OBS) in 482 participants with non small cell lung cancer (NSCLC) who had undergone surgery.
-Of those 482 participants, 169 had frozen tumor tissue collected, and of those tumor samples, 133 (71 in ACT and 62 in OBS) had gene-expression profiling performed using U133A oligonucleotide microarrays (Affymetrix, Santa Clara, CA).
-
-The goal of the @zhu2010prognostic paper was to identify a multi-gene signature that strongly predicts prognosis, and the hypothesis was that the poor prognosis subgroup would benefit more from ACT than the good prognosis subgroup.
-The signature was developed on a training data set ("trained") to predict disease specific survival.
-The annotated gene expression data and clinical information are available from the Gene Expression Omnibus database(identifier: GSE14814, @edgar2002gene).
-
-@zhu2010prognostic present results that mainly focus on the association of their signature with outcome, albeit a less than ideal measure of discrimination ability.
-They demonstrate that the two risk subgroups predicted by their signature (high risk and low risk) have separation in their survival curves and that the hazard ratio for their signature is large and significant even when adjusting for other risk factors.
-They do not directly address calibration, that is, whether their signature accurately predicts survival times.
-
-We used a similar approach to preprocessing as did @zhu2010prognostic, although we could not reproduce their workflow exactly due to outdated software.
-Batch effects were removed using the `ComBat` function in the `sva` `R` package [@leeksva] and then the gene expression values were centered by their means and scaled by their standard deviations.
-Our signature development approach is similar but not identical to that in @zhu2010prognostic.
-For purposes of illustrating the concepts we used a simplified approach to signature development that retains the main characteristics of the original method.
-The exact approach to signature development does not have a major impact on the main conclusions of our evaluation of the various approaches to signature performance assessment.
-
-After processing the data as described above, we performed a gene selection step wherein we fit univariate Cox regression models with disease specific survival as the outcome and each gene as the single predictor.
-Genes with univariate p-values less than 0.005 were preliminarily selected for further analysis.
-Then, each gene from the preliminary list was weighted by its univariate Cox regression coefficient, and the resulting weighted gene expression values were summed to form risk scores.
-Genes were selected for inclusion in the risk score in a forward selection manner.
-Starting with the most significant weighted gene, the gene that when added to the risk score improved the concordance between survival times and the risk score most was selected next.
-If no gene improved the concordance, the process was stopped.
-All genes on the final selected list were included in a multivariable Cox regression model to fit the final risk score with new coefficients.
-We selected the cutoff that split the patients into two risk groups corresponding to the smallest log-rank statistic p-value when applied to the continuous risk score.
+After processing the data as described above, we performed a gene
+selection step wherein we fit univariate Cox regression models with
+disease specific survival as the outcome and each gene as the single
+predictor. Genes with univariate p-values less than 0.005 were
+preliminarily selected for further analysis. Then, each gene from the
+preliminary list was weighted by its univariate Cox regression
+coefficient, and the resulting weighted gene expression values were
+summed to form risk scores. Genes were selected for inclusion in the
+risk score in a forward selection manner. Starting with the most
+significant weighted gene, the gene that when added to the risk score
+improved the concordance between survival times and the risk score most
+was selected next. If no gene improved the concordance, the process was
+stopped. All genes on the final selected list were included in a
+multivariable Cox regression model to fit the final risk score with new
+coefficients. We selected the cutoff that split the patients into two
+risk groups corresponding to the smallest log-rank statistic p-value
+when applied to the continuous risk score.
 
 The signature development procedure was described in the introduction.
-There is both a feature selection step, and a multivariable estimation step.
-This results in a continuous signature which is the linear predictor of a Cox regression model.
-The signature is dichotomized by selecting the cutoff that yields the most significant log-rank statistic for comparing the resulting risk groups.
-Discrimination of the signature is assessed using the concordance statistic as implemented in the \texttt{survival} package in \texttt{R} [@survival].
-To paraphrase the help file: this is defined as the probability of agreement for any two randomly chosen observations, which in this case means that the observation with the shorter survival time also has the larger signature value.
-This is similar to an interpretation of the AUC for binary data.
+There is both a feature selection step, and a multivariable estimation
+step. This results in a continuous signature which is the linear
+predictor of a Cox regression model. The signature is dichotomized by
+selecting the cutoff that yields the most significant log-rank statistic
+for comparing the resulting risk groups. Discrimination of the signature
+is assessed using the concordance statistic as implemented in the
+\texttt{survival} package in \texttt{R} \citep{survival}. To paraphrase
+the help file: this is defined as the probability of agreement for any
+two randomly chosen observations, which in this case means that the
+observation with the shorter survival time also has the larger signature
+value. This is similar to an interpretation of the AUC for binary data.
 
 First we fit the signature using the entire observation cohort (n = 62).
-The signature was then evaluated on the same dataset.
-The survival plot on the left side of Figure \ref{fig2} shows extreme separation between the two risk groups (HR = `r round(exp(evalall[2]))`, p < 0.001), consistent with the reported JBL.10 signature, and the estimated concordance is `r round(evalall[1], 2)`.
-After correctly accounting for the selection process, our estimates of association and discrimination are much less impressive.
+The signature was then evaluated on the same dataset. The survival plot
+on the left side of Figure \ref{fig2} shows extreme separation between
+the two risk groups (HR = 20, p \textless{} 0.001), consistent with the
+reported JBL.10 signature, and the estimated concordance is 0.87. After
+correctly accounting for the selection process, our estimates of
+association and discrimination are much less impressive.
 
-```{r examppreval, cache = TRUE}
-set.seed(332)
-obssel <- unique(obsgrp$ID)
-# randomly reorder and select 8 groups of 6 and 2 groups of 7
-obssel <- obssel[sample(1:length(obssel), length(obssel), replace = FALSE)]
-sets <- c(rep(6, 8), 7, 7)
-scpet <- vector(mode = "list", length = 10)
-j <- 1
-for(i in 1:10){
-  scpet[[i]] <- obssel[j:(j + sets[i] - 1)]
-  j <- j + sets[i]
-}
+The right plot in Figure \ref{fig2} shows the survival curves for the
+two risk groups using the pre-validated estimates of the risk score. We
+partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
+Then for each group \(b\), we fit the model using \(S_{-b}\) and
+obtained prevalidated estimates for \(S_{b}\). The survival curves plot
+the survival times for comparing risk groups using the prevalidated
+estimates. The separation is much less impressive. The concordance
+between the prevalidated signature and the survival times is 0.61,
+indicating much worse discrimination.
 
-hat.f <- vector("list", 10)
-for(i in 1:10){
+\begin{figure}[htbp]
+\centering
+\includegraphics{paper-revised_files/figure-latex/examppresub-1.pdf}
+\caption{Comparison of survival by gene expression based risk signature.
+The left plot shows the resubstitution estimate, while the right plot
+shows the pre-validated estimate. \label{fig2}}
+\end{figure}
 
-  obsgrpb <- subset(ldat, Post.Surgical.Treatment == "OBS" & ID %in% setdiff(obssel, scpet[[i]]))
-  obsgrpwb <- subset(gdat, Post.Surgical.Treatment == "OBS" & ID %in% scpet[[i]])
-  fit.b <- fit.superpc(obsgrpb, cldat)
-  hat.f[[i]] <- cbind(ID = scpet[[i]], as.data.frame(predict.superpc(obsgrpwb, fit.b)))
+\begin{figure}[htbp]
+\centering
+\includegraphics{paper-revised_files/figure-latex/inter1-1.pdf}
+\caption{Survival curves comparing the treatment effect by gene
+expression signature risk group. The top set of plots shows the partial
+resubstitution based signature and the bottom row shows the
+pre-validated signature estimates. \label{fig3}}
+\end{figure}
 
-}
+\begin{longtable}[c]{@{}lllll@{}}
+\caption{Hazard ratios and 95\% confidence intervals from separate Cox
+regression models that adjust for tumor histologic subtype, stages, age,
+and sex. Rows labeled `High risk vs low risk' show the hazard ratio for
+the signature-based risk group comparison. The rows labeled `Trt/Risk
+interaction' show the hazard ratio for the interaction term of treatment
+by signature-based risk group. The partial substitution estimates are
+dramatically optimistically biased. \label{adjhr}}\tabularnewline
+\toprule
+Method & Comparison & Hazard Ratio & 95\% CI & Adjusted p\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+Method & Comparison & Hazard Ratio & 95\% CI & Adjusted p\tabularnewline
+\midrule
+\endhead
+\textbf{Partial Resubstitution} & High Risk vs Low Risk & 38.9 & 9.2 to
+164.7 & \textless{} 0.001\tabularnewline
+& Trt/Risk interaction & 14.7 & 3.2 to 67.0 & \textless{}
+0.001\tabularnewline
+\textbf{Prevalidation} & High Risk vs Low Risk & 1.9 & 0.8 to 4.3 &
+0.122\tabularnewline
+& Trt/Risk interaction & 1.8 & 0.5 to 6.5 & 0.395\tabularnewline
+\bottomrule
+\end{longtable}
 
-preval.dat <- merge(do.call("rbind", hat.f), cldat[, - which(colnames(cldat) %in% c("riskgrp", "lps"))], by = "ID", all.y = FALSE)
+We also show plots to assess the ability of the signature to be useful
+for treatment selection (Figure \ref{fig3}). These plots show the
+survival curves comparing treatment arms grouped in panels by the risk
+score. As described by \citet{polley2013statistical}, the idea is to
+determine whether the treatment is beneficial in one group and not
+benficial or harmful in another group, indicating that different
+treatment decisions would be made based on the signature. On one hand,
+the overfit signature shows dramatic differences in treatment efficacy
+between the low risk and high risk groups. In fact it appears that the
+treatment is harmful in the low risk group, but highly beneficial in the
+high risk group. The prevalidated signature on the other hand, shows
+differences that are much less dramatic. It appears that the treatment
+is mildly beneficial in both groups, possibly to a higher degree in the
+high risk group. This suggests that the dramatic predictive value of the
+signature was merely an artifact of overfitting the signature to the OBS
+arm data. \citet{simon2011re} explain the flaws in this type of approach
+in which one attempts to develop a signature for treatment selection by
+developing a prognostic signature on data from the control arm of a
+trial and then applying that signature to both the control and
+experimental arm data to establish predictiveness.
 
-# ggplot(preval.dat, 
-#        aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km() + 
-#   theme_bw(base_size = 13, base_family = "serif") + 
-#   scale_linetype_discrete("Prevalidated", labels = c("Low risk", "High risk")) + 
-#   ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization")
-```
+In a multivariable Cox model we observe similar trends when comparing
+the prevalidated signature to the overfit signature. We fit two
+regression models. In the first, the aim is to assess the prognostic
+value of the signature by estimating the hazard ratio for the high risk
+versus low risk groups, adjusted for tumor histologic subtype, stage,
+age, and sex. In the second model, the idea is to assess the predictive
+value of the signature by estimating the treatment by signature
+interaction effect, adjusting for the same clinical covariates. The
+results are reported in Table \ref{adjhr}. The full model summaries are
+report in Table \ref{multimodel}. For the partial resubstitution
+approach, we find an extreme hazard ratio of nearly 40 for the
+prognostic effect, and a strong and significant treatment by signature
+interaction. Using the prevalidated signature, the effect estimates are
+much smaller and of small magnitude in comparison to the standard
+clinical features. Note that the inference from the pre-validated model
+is not exactly correct either, because the resampling procedure induces
+a correlation among the observations (the prevalidated signature values
+are estimated from heavily overlapping data). Despite this, these
+results are unimpressive and would not be considered promising for
+clinical use.
 
-The right plot in Figure \ref{fig2} shows the survival curves for the two risk groups using the pre-validated estimates of the risk score.
-We partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
-Then for each group $b$, we fit the model using $S_{-b}$ and obtained prevalidated estimates for $S_{b}$.
-The survival curves plot the survival times for comparing risk groups using the prevalidated estimates.
-The separation is much less impressive.
-The concordance between the prevalidated signature and the survival times is `r round(survConcordance(Surv(DSS.time, DSS) ~ lps, data = preval.dat)$concordance, 2)`, indicating much worse discrimination.
+\citet{zhu2010prognostic} should be commended for their commitment to
+making their data, methods, and analysis code publically available,
+which allowed us to reanalyze their study data. More often, methods used
+to derive signatures are vaguely and incompletely presented and a
+resubstitution analysis or other flawed approach goes undetected or can
+be discovered only through a very careful scrutiny of the methods and
+supplementary materials.
 
-```{r examppresub, fig.height = 6.5, fig.width = 8, fig.cap = "Comparison of survival by gene expression based risk signature.
-The left plot shows the resubstitution estimate, while the right plot shows the pre-validated estimate.
-\\label{fig2}"}
-trtgrp <- subset(gdat, Post.Surgical.Treatment == "ACT")
+\begin{table}[!htbp] \centering 
+  \caption{Complete Results of Multivariable Regression Models. Entries are hazard ratios and 95 percent confidence intervals.  \label{multimodel}} 
+  \label{} 
+\begin{tabular}{@{\extracolsep{5pt}}lcccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+ & \multicolumn{4}{c}{\textit{Dependent variable:}} \\ 
+\cline{2-5} 
+\\[-1.8ex] & \multicolumn{4}{c}{Disease-specific Survival} \\ 
+ & Resubstitution & Prevalidation & Resubstitution & Prevalidation \\ 
+\\[-1.8ex] & (1) & (2) & (3) & (4)\\ 
+\hline \\[-1.8ex] 
+ High Risk & 38.9$^{***}$ & 1.9 & 0.7 & 1.0 \\ 
+  & (9.2 to 164.7) & (0.8 to 4.3) & (0.3 to 2.0) & (0.4 to 2.7) \\ 
+  & & & & \\ 
+ Trt: OBS &  &  & 0.5 & 1.3 \\ 
+  &  &  & (0.2 to 1.3) & (0.6 to 2.8) \\ 
+  & & & & \\ 
+ Male Gender & 1.3 & 1.4 & 1.2 & 1.4 \\ 
+  & (0.4 to 3.8) & (0.5 to 4.2) & (0.6 to 2.4) & (0.7 to 2.8) \\ 
+  & & & & \\ 
+ Stage > I & 2.1$^{*}$ & 2.7$^{**}$ & 2.0$^{**}$ & 2.3$^{***}$ \\ 
+  & (0.9 to 4.7) & (1.2 to 6.3) & (1.1 to 3.6) & (1.3 to 4.1) \\ 
+  & & & & \\ 
+ Age in Years & 1.0 & 1.1$^{**}$ & 1.0$^{*}$ & 1.1$^{***}$ \\ 
+  & (0.9 to 1.1) & (1.0 to 1.1) & (1.0 to 1.1) & (1.0 to 1.1) \\ 
+  & & & & \\ 
+ Histology: LCUC & 1.7 & 2.7 & 1.5 & 1.9 \\ 
+  & (0.5 to 6.3) & (0.8 to 10.0) & (0.6 to 3.7) & (0.7 to 4.7) \\ 
+  & & & & \\ 
+ Histology: SQCC & 3.4$^{*}$ & 0.4$^{**}$ & 0.8 & 0.4$^{***}$ \\ 
+  & (0.9 to 13.3) & (0.2 to 1.0) & (0.3 to 1.7) & (0.2 to 0.8) \\ 
+  & & & & \\ 
+ High Risk * Trt:OBS &  &  & 14.7$^{***}$ & 1.8 \\ 
+  &  &  & (3.2 to 67.0) & (0.5 to 6.5) \\ 
+  & & & & \\ 
+\hline \\[-1.8ex] 
+Observations & 62 & 62 & 133 & 133 \\ 
+Score (Logrank) Test & 69.6$^{***}$ (df = 6) & 21.5$^{***}$ (df = 6) & 84.5$^{***}$ (df = 8) & 32.0$^{***}$ (df = 8) \\ 
+\hline 
+\hline \\[-1.8ex] 
+\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
+\end{tabular} 
+\end{table}
 
-trt.dat <- merge(data.frame(ID = trtgrp$ID, lps = predict.superpc(trtgrp, fit.all)$lps, 
-                            riskgrp = predict.superpc(trtgrp, fit.all)$riskgrp), 
-                 cldat[, - which(colnames(cldat) %in% c("riskgrp", "lps"))], by = "ID", all.y = FALSE)
+\section{Discussion}\label{discussion}
 
-preval.dat2 <- rbind(preval.dat, trt.dat)
+Examples of flawed approaches such as naive resubstitution, partial
+resubstitution, partial holdout, and partial CV remain very common in
+the published literature more than a decade after many high-dimensional
+molecular assays were first introduced. All performance metrics, whether
+they assess calibration or discrimination or something else, are subject
+to bias in estimation due to overfitting. Strategies to avoid this type
+of bias are well-studied in the statistical literature and here we have
+demonstrated how they can be implemented in a real clinical example. The
+ability to perform the kind of in-depth data reanalysis we performed on
+the example in this paper is usually limited by restrictions on data
+access and insufficient description of analysis methods. Sadly, reports
+of strong associations with overfit biomarker signatures are all too
+common in the medical literature. The amount of bias that is possible is
+not known and can be difficult to decipher based on study reports. It is
+imperative that investigators and journal editors take overfitting bias
+seriously to ensure that signature performance claims are valid.
+Unfortunately this often results in study reports that are far less
+optimistic than usual.
 
-cldat.trtplot <- cldat
-cldat.trtplot$Post.Surgical.Treatment <- "Combined"
-cldat.trtplot <- rbind(cldat, cldat.trtplot)
+The \citet{zhu2010prognostic} paper used the approach of identifying a
+signature using the control arm of the trial, evaluating it using the
+combined control and treatment arms, and then hoping that the signature
+would be useful for treatment selection. This approach has been shown to
+be invalid \citep{simon2011re} for identifying a predictive signature,
+and indeed we show that the resubstitution aspect of the evaluation
+likely led to overstatments of the size of the signature effect.
+Properly doing cross-validation for estimating the calibration of the
+signature and doing pre-validation for assessment of the discrimination
+of the signature show that the associations are much more modest.
 
-row1 <- subset(cldat.trtplot, Post.Surgical.Treatment == "OBS")
+When we assess the signature on the independent treatment arm, we see
+that there is no significant difference between the risk groups. This
+raises the question of whether it is appropriate to use non-random
+splits of the dataset in order to obtain valid estimates of the
+calibration or discrimination. Instead of treatment arms, we could
+imagine a large multi-center study, and we could split the data into
+disjoint subgroups based on the center. Specifically, suppose we split
+the development dataset \(S\) into \(S_1\) and \(S_2\) according to a
+discrete covariate \(X\) that takes on levels 1 or 2. The we develop a
+signature \(f_{S_1}\) using \(S_1\) and evaluate it on \(S_2\) by
+estimating \(\hat{E}[\phi_{f_{S_1}}(S_2)]\). This is an estimate not of
+\(E_\mathcal{P}[\phi_f(S)]\), but rather
+\(E_{\mathcal{P}_2}[\phi_f(S)]\), where \(\mathcal{P}_2\) is the
+distribution for the sub-population with \(X = 2\) that \(S_2\) is a
+sample from. This estimate would only be recommended if the signature is
+intended for use in that specific subpopulation, and if that were the
+case, then it doesn't make much sense to develop the signature using the
+subpopulation \(\mathcal{P}_1\). If these groups differed substantially,
+then we would not expect the signature to perform adequately. The
+differences in performance may depend on many factors, including how
+\(X\) is associated with the distribution of the features and the
+outcome. If a signature becomes broadly used in clinics, these
+center-to-center differences would be important to assess as a part of
+signature efficacy surveillance.
 
-preval.dat2.trtplot <- preval.dat2
-preval.dat2.trtplot$Post.Surgical.Treatment <- "Combined"
-preval.dat2.trtplot <- rbind(preval.dat2, preval.dat2.trtplot)
-preval.dat2.trtplot$Risk.fac <- factor(preval.dat2.trtplot$riskgrp, levels = c(FALSE, TRUE), labels = c("Low risk", "High risk"))
+\section{Note}\label{note}
 
-row2 <- subset(preval.dat2.trtplot, Post.Surgical.Treatment == "OBS")
-row1$Set <- "Resubstitution"
-row2$Set <- "Pre-validation"
+All analysis code and the source files for this manuscript are available
+from the authors' webpage.
 
-rows <- bind_rows(row1, row2)
-rows$Set <- factor(rows$Set, levels = c("Resubstitution", "Pre-validation"), ordered = TRUE)
+\renewcommand\refname{References}
+\bibliography{paper}
 
-ggplot(rows, aes(time = DSS.time, status = DSS, linetype = riskgrp)) + geom_km() + geom_kmticks() + 
-  theme_bw(base_size = 13, base_family = "serif") + 
-  scale_linetype_discrete("Signature Group: ", labels = c("Low risk", "High risk")) + 
-  ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
-  facet_wrap(~ Set, ncol = 2) + theme(legend.position = "bottom")
-```
-
-```{r inter1, fig.height = 9, fig.width = 7.5, fig.cap = "Survival curves comparing the treatment effect by gene expression signature risk group.
-The top set of plots shows the partial resubstitution based signature and the bottom row shows the pre-validated signature estimates.
-\\label{fig3}"}
-cldat.trtplot$Risk.fac <- factor(cldat.trtplot$riskgrp, levels = c(FALSE, TRUE), labels = c("Low risk", "High risk"))
-
-library(gtable)
-library(grid)
-
-inter1 <- subset(cldat.trtplot, Post.Surgical.Treatment != "Combined") 
-inter2 <- subset(preval.dat2.trtplot, Post.Surgical.Treatment != "Combined") 
-inter1$Set <- "Resubstitution"
-inter2$Set <- "Pre-validation"
-
-inters <- bind_rows(inter1, inter2)
-inters$Set <- factor(inters$Set, levels = c("Resubstitution", "Pre-validation"), ordered = TRUE)
-
-p1 <- ggplot(inters, aes(time = DSS.time, status = DSS, linetype = Post.Surgical.Treatment)) + geom_km()+ geom_kmticks() + 
-  theme_bw(base_size = 13, base_family = "serif") + 
-  scale_linetype_discrete("Treatment") + 
-  ylab("Disease Specific Survival") + ylim(0, 1) + xlab("Time in years since randomization") + 
-  facet_grid( Set ~ Risk.fac)  + theme(strip.text.y = element_text(angle = 0))
-
-gt <- ggplotGrob(p1)
-# Get the positions of the panels in the layout: t = top, l = left, ...
-panels <-c(subset(gt$layout, name == "panel", select = t:r))
-
-# Add a row below the x-axis tick mark labels,
-# the same height as the strip
-gt <- gtable_add_rows(gt, gt$heights[3], 3)
-gt <- gtable_add_rows(gt, gt$heights[3], 6)
-
-# Get the strip grob
-stripGrob = gtable_filter(gt, "strip-right")
-
-# Insert the strip grob into the new row
-gt = gtable_add_grob(gt, stripGrob[[1]][[1]], t = 4, l = min(panels$l), r = max(panels$r))
-gt = gtable_add_grob(gt, stripGrob[[1]][[2]], t = 7, l = min(panels$l), r = max(panels$r))
-# remove the old strip
-gt = gt[, -(max(panels$r) + 1)]
-
-grid.newpage()
-grid.draw(gt)
-```
-
-```{r multivar, results = "asis"}
-cldat$Stage2 <- cldat$Stage %in% c("2A", "2B", "II")
-fit.train <- coxph(Surv(DSS.time, DSS) ~ riskgrp + Sex + Stage2 + age + Histology.type, 
-                   data = subset(cldat, Post.Surgical.Treatment == "OBS"))
-
-fit.train.inter <- coxph(Surv(DSS.time, DSS) ~ riskgrp * Post.Surgical.Treatment + Sex + Stage2 + age + Histology.type, 
-                   data = cldat)
-
-preval.dat2$Stage2 <- preval.dat2$Stage %in% c("2A", "2B", "II")
-fit.preval <- coxph(Surv(DSS.time, DSS) ~ riskgrp + Sex + Stage2 + age + Histology.type, 
-                    data = subset(preval.dat2, Post.Surgical.Treatment == "OBS"))
-
-fit.preval.inter <- coxph(Surv(DSS.time, DSS) ~ riskgrp * Post.Surgical.Treatment + Sex + Stage2 + age + Histology.type,
-                          data = preval.dat2)
-
-pCI <- function(x) {
-  sprintf("%.1f to %.1f", x[1], x[2])
-}
-
-myP <- function(x){
-  ifelse(x < .001, "< 0.001", paste(round(x, 3)))
-}
-
-HRmary <- data.frame(Comparison = "High Risk vs Low Risk", `Hazard Ratio` = round(exp(fit.train$coefficients[1]), 1), 
-           `95% CI` = pCI(exp(confint(fit.train)[1, ])), `Adjusted p` = myP(summary(fit.train)$coefficients[1, 5]), 
-           check.names = FALSE, stringsAsFactors = FALSE)
-
-HRmary <- rbind(HRmary, c("Trt/Risk interaction", round(exp(fit.train.inter$coefficients[8]), 1), 
-                          pCI(exp(confint(fit.train.inter)[8, ])), 
-                          myP(summary(fit.train.inter)$coefficients[8, 5])))
-
-HRmary <- rbind(HRmary, c("High Risk vs Low Risk", round(exp(fit.preval$coefficients[1]), 1), 
-                          pCI(exp(confint(fit.preval)[1, ])), 
-                          myP(summary(fit.preval)$coefficients[1, 5])),
-                c("Trt/Risk interaction", round(exp(fit.preval.inter$coefficients[8]), 1), 
-                          pCI(exp(confint(fit.preval.inter)[8, ])), 
-                          myP(summary(fit.preval.inter)$coefficients[8, 5])))
-
-rownames(HRmary) <- NULL
-
-HRmary <- cbind(Method = c("**Partial Resubstitution**", "", "**Prevalidation**", ""), HRmary)
-kable(HRmary, caption = "Hazard ratios and 95% confidence intervals from separate Cox regression models that adjust for tumor histologic subtype, stages, age, and sex.
-Rows labeled 'High risk vs low risk' show the hazard ratio for the signature-based risk group comparison.
-The rows labeled 'Trt/Risk interaction' show the hazard ratio for the interaction term of treatment by signature-based risk group.
-The partial substitution estimates are dramatically optimistically biased.
-\\label{adjhr}")
-```
-
-We also show plots to assess the ability of the signature to be useful for treatment selection (Figure \ref{fig3}).
-These plots show the survival curves comparing treatment arms grouped in panels by the risk score.
-As described by @polley2013statistical, the idea is to determine whether the treatment is beneficial in one group and not benficial or harmful in another group, indicating that different treatment decisions would be made based on the signature.
-On one hand, the overfit signature shows dramatic differences in treatment efficacy between the low risk and high risk groups.
-In fact it appears that the treatment is harmful in the low risk group, but highly beneficial in the high risk group.
-The prevalidated signature on the other hand, shows differences that are much less dramatic.
-It appears that the treatment is mildly beneficial in both groups, possibly to a higher degree in the high risk group.
-This suggests that the dramatic predictive value of the signature was merely an artifact of overfitting the signature to the OBS arm data.
-@simon2011re explain the flaws in this type of approach in which one attempts to develop a signature for treatment selection by developing a prognostic signature on data from the control arm of a trial and then applying that signature to both the control and experimental arm data to establish predictiveness.
-
-In a multivariable Cox model we observe similar trends when comparing the prevalidated signature to the overfit signature.
-We fit two regression models.
-In the first, the aim is to assess the prognostic value of the signature by estimating the hazard ratio for the high risk versus low risk groups, adjusted for tumor histologic subtype, stage, age, and sex.
-In the second model, the idea is to assess the predictive value of the signature by estimating the treatment by signature interaction effect, adjusting for the same clinical covariates.
-The results are reported in Table \ref{adjhr}.
-The full model summaries are report in Table \ref{multimodel}.
-For the partial resubstitution approach, we find an extreme hazard ratio of nearly 40 for the prognostic effect, and a strong and significant treatment by signature interaction.
-Using the prevalidated signature, the effect estimates are much smaller and of small magnitude in comparison to the standard clinical features.
-Note that the inference from the pre-validated model is not exactly correct either, because the resampling procedure induces a correlation among the observations (the prevalidated signature values are estimated from heavily overlapping data).
-Despite this, these results are unimpressive and would not be considered promising for clinical use.
-
-@zhu2010prognostic should be commended for their commitment to making their data, methods, and analysis code publically available, which allowed us to reanalyze their study data.
-More often, methods used to derive signatures are vaguely and incompletely presented and a resubstitution analysis or other flawed approach goes undetected or can be discovered only through a very careful scrutiny of the methods and supplementary materials.
-
-```{r modtab, results = "asis"}
-stargazer(fit.train, fit.preval, fit.train.inter, fit.preval.inter, 
-          column.labels = rep(c("Resubstitution", "Prevalidation"), 2), 
-          covariate.labels = c("High Risk", "Trt: OBS", "Male Gender", "Stage > I", "Age in Years", "Histology: LCUC", 
-                               "Histology: SQCC", "High Risk * Trt:OBS"), 
-          dep.var.labels = "Disease-specific Survival", ci = TRUE, ci.separator = " to ", digits = 1, 
-          keep.stat = c("n", "logrank"), header = FALSE, apply.coef = exp,  apply.ci = exp, p.auto = FALSE,
-          ci.custom = lapply(list(fit.train, fit.preval, fit.train.inter, fit.preval.inter), confint),
-          title = "Complete Results of Multivariable Regression Models. Entries are hazard ratios and 95 percent confidence intervals.  \\label{multimodel}")
-```
-
-# Discussion
-
-Examples of flawed approaches such as naive resubstitution, partial resubstitution, partial holdout, and partial CV remain very common in the published literature more than a decade after many high-dimensional molecular assays were first introduced.
-All performance metrics, whether they assess calibration or discrimination or something else, are subject to bias in estimation due to overfitting.
-Strategies to avoid this type of bias are well-studied in the statistical literature and here we have demonstrated how they can be implemented in a real clinical example.
-The ability to perform the kind of in-depth data reanalysis we performed on the example in this paper is usually limited by restrictions on data access and insufficient description of analysis methods.
-Sadly, reports of strong associations with overfit biomarker signatures are all too common in the medical literature.
-The amount of bias that is possible is not known and can be difficult to decipher based on study reports.
-It is imperative that investigators and journal editors take overfitting bias seriously to ensure that signature performance claims are valid.
-Unfortunately this often results in study reports that are far less optimistic than usual.
-
-The @zhu2010prognostic paper used the approach of identifying a signature using the control arm of the trial, evaluating it using the combined control and treatment arms, and then hoping that the signature would be useful for treatment selection.
-This approach has been shown to be invalid [@simon2011re] for identifying a predictive signature, and indeed we show that the resubstitution aspect of the evaluation likely led to overstatments of the size of the signature effect.
-Properly doing cross-validation for estimating the calibration of the signature and doing pre-validation for assessment of the discrimination of the signature show that the associations are much more modest.
-
-When we assess the signature on the independent treatment arm, we see that there is no significant difference between the risk groups.
-This raises the question of whether it is appropriate to use non-random splits of the dataset in order to obtain valid estimates of the calibration or discrimination.
-Instead of treatment arms, we could imagine a large multi-center study, and we could split the data into disjoint subgroups based on the center.
-Specifically, suppose we split the development dataset $S$ into $S_1$ and $S_2$ according to a discrete covariate $X$ that takes on levels 1 or 2.
-The we develop a signature $f_{S_1}$ using $S_1$ and evaluate it on $S_2$ by estimating $\hat{E}[\phi_{f_{S_1}}(S_2)]$.
-This is an estimate not of $E_\mathcal{P}[\phi_f(S)]$, but rather $E_{\mathcal{P}_2}[\phi_f(S)]$, where $\mathcal{P}_2$ is the distribution for the sub-population with $X = 2$ that $S_2$ is a sample from.
-This estimate would only be recommended if the signature is intended for use in that specific subpopulation, and if that were the case, then it doesn't make much sense to develop the signature using the subpopulation $\mathcal{P}_1$.
-If these groups differed substantially, then we would not expect the signature to perform adequately.
-The differences in performance may depend on many factors, including how $X$ is associated with the distribution of the features and the outcome.
-If a signature becomes broadly used in clinics, these center-to-center differences would be important to assess as a part of signature efficacy surveillance.
-
-# Note
-
-All analysis code and the source files for this manuscript are available from the authors' webpage.
-
-# References
+\end{document}

### 2016-07-05 Code/new sims/supplement.Rmd
new simulations

---
title: "Issues in developing multivariable molecular signatures for guiding clinical care decisions: supplemental simulation results"
author: "Michael C Sachs and Lisa M McShane"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
    pdf_document:
      citation_package: natbib
      keep_tex: yes
      fig_caption: yes
      pandoc_args: ["--bibliography", "paper.bib"]
header-includes:
    - \usepackage{setspace}
fontsize: 11pt
geometry: margin=1in
---

# Additional Simulation Results

In the tables and figures below, $N$ is to the number of samples, and $p$ is the number of variables per sample. All results are from 1000 simulation replicates.

```{r setup, include = FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(survival)
library(ggkm)
library(tidyr)
library(stargazer)

opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE, warning = FALSE)
knit_hooks$set(plot = function(x, options) {
  if ('tikz' %in% options$dev && !options$external) {
    hook_plot_tex(x, options)
  } else hook_plot_md(x, options)
})

source("newsims.R")

```

```{r cycle, echo = FALSE, results = "asis"}
files <- grep("*.txt", list.files(), value = TRUE)
results <- lapply(files, parse_sims)

for(i in results){

  cat(i$table)

}
```

```{r plotsauc, echo = FALSE, results = "asis"}
for(i in results){

  print(i$plot.auc)
  cat("\n\\clearpage\n\n")

}
```

```{r plots2, echo = FALSE, results = "asis"}
for(i in results){

  print(i$plot.or)
  cat("\n\\clearpage\n\n")

}
```

```{r overall, fig.height = 8.5}
alldat <- NULL
for(i in results){

  alldat <- rbind(alldat, i$datalong)

}

 labs <- c("Resubstitution", "Partial CV",
            "Partial Holdout", "Partial Resubstitution",
            "Pre-validation", "Leave 10 out CV", "Leave 100 out CV",
            "30% Holdout", "50% Holdout", "Bootstrap")

  names(labs) <- levels(alldat$scen)

ggplot(subset(alldat, stat == "AUC"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") +
    theme_bw(base_size = 13, base_family = "serif") + coord_flip() +
    scale_x_discrete("Estimation Approach", labels = labs) +
    ylab("Area Under the ROC Curve") + facet_grid(p ~ N, labeller = "label_both")
```

```{r, fig.height = 8.5}
alldat$value[alldat$stat == "OR"] <- exp(alldat$value[alldat$stat == "OR"])

ggplot(subset(alldat, stat == "OR"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") +
    theme_bw(base_size = 13, base_family = "serif") + coord_flip() +
    scale_x_discrete("Estimation Approach", labels = labs) + scale_y_log10(breaks = c(.5, 2, 5, 20, 80)) +
    ylab("Odds Ratio (log scale)") + facet_grid(p ~ N, labeller = "label_both")

```

### 2016-07-05 Code/new sims/supplement.tex
new simulations

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Issues in developing multivariable molecular signatures for guiding clinical care decisions: supplemental simulation results},
            pdfauthor={Michael C Sachs and Lisa M McShane},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Issues in developing multivariable molecular signatures for guiding
clinical care decisions: supplemental simulation results}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Michael C Sachs$^*$ and Lisa M McShane}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{June 30, 2016}

\usepackage{setspace}

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\begin{document}
\maketitle

\begin{center}
\textit{Biostatistics Branch, Biometric Research Program \\
National Cancer Institute \\
9609 Medical Center Drive, MSC 9735 \\
Bethesda, MD 20892-9735 \\
Phone: 240-276-6004 \\
Email: michael.sachs@nih.gov \\
*: Corresponding Author}

\end{center}

\clearpage

\section{Additional Simulation
Results}\label{additional-simulation-results}

In the tables and figures below, \(N\) is to the number of samples, and
\(p\) is the number of variables per sample. All results are from 1000
simulation replicates.

\clearpage

\begin{table}

\caption{Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance. The true value of the AUC is 0.5 and the true value of the Log OR is 0.0. Estimates are based on 1000 replicates of the numerical experiment. In each replicate,  N = 1000; p = 10 . CV = Cross validation. }
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR & Bias OR\\
\hline
Resubstitution & 0.56 & 0.01 & 0.06 & 0.34 & 0.12 & 0.34\\
\hline
Partial CV & 0.50 & 0.03 & 0.00 & -0.01 & 0.18 & -0.01\\
\hline
Partial Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.20 & 0.00\\
\hline
Partial Resubstitution & 0.54 & 0.02 & 0.04 & 0.24 & 0.13 & 0.24\\
\hline
Pre-validation & 0.49 & 0.03 & -0.01 & -0.02 & 0.18 & -0.02\\
\hline
Leave 10 out CV & 0.50 & 0.03 & 0.00 & 0.01 & 0.18 & 0.01\\
\hline
Leave 100 out CV & 0.50 & 0.03 & 0.00 & 0.00 & 0.17 & 0.00\\
\hline
30\% Holdout & 0.50 & 0.04 & 0.00 & -0.01 & 0.24 & -0.01\\
\hline
50\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.19 & 0.00\\
\hline
Bootstrap & 0.50 & 0.02 & 0.00 & 0.00 & 0.11 & 0.00\\
\hline
\end{tabular}
\end{table}

\begin{table}

\caption{Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance. The true value of the AUC is 0.5 and the true value of the Log OR is 0.0. Estimates are based on 1000 replicates of the numerical experiment. In each replicate,  N = 1000; p = 100 . CV = Cross validation. }
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR & Bias OR\\
\hline
Resubstitution & 0.66 & 0.01 & 0.16 & 0.95 & 0.12 & 0.95\\
\hline
Partial CV & 0.61 & 0.02 & 0.11 & 0.59 & 0.13 & 0.59\\
\hline
Partial Holdout & 0.59 & 0.03 & 0.09 & 0.51 & 0.19 & 0.51\\
\hline
Partial Resubstitution & 0.61 & 0.02 & 0.11 & 0.66 & 0.13 & 0.66\\
\hline
Pre-validation & 0.50 & 0.04 & 0.00 & -0.02 & 0.26 & -0.02\\
\hline
Leave 10 out CV & 0.50 & 0.04 & 0.00 & 0.00 & 0.21 & 0.00\\
\hline
Leave 100 out CV & 0.50 & 0.03 & 0.00 & 0.00 & 0.19 & 0.00\\
\hline
30\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.25 & 0.00\\
\hline
50\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.20 & 0.00\\
\hline
Bootstrap & 0.50 & 0.02 & 0.00 & 0.00 & 0.10 & 0.00\\
\hline
\end{tabular}
\end{table}

\begin{table}

\caption{Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance. The true value of the AUC is 0.5 and the true value of the Log OR is 0.0. Estimates are based on 1000 replicates of the numerical experiment. In each replicate,  N = 1000; p = 500 . CV = Cross validation. }
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR & Bias OR\\
\hline
Resubstitution & 0.72 & 0.01 & 0.22 & 1.33 & 0.11 & 1.33\\
\hline
Partial CV & 0.68 & 0.01 & 0.18 & 1.03 & 0.11 & 1.03\\
\hline
Partial Holdout & 0.66 & 0.02 & 0.16 & 0.96 & 0.18 & 0.96\\
\hline
Partial Resubstitution & 0.65 & 0.02 & 0.15 & 0.89 & 0.13 & 0.89\\
\hline
Pre-validation & 0.50 & 0.06 & 0.00 & -0.01 & 0.33 & -0.01\\
\hline
Leave 10 out CV & 0.50 & 0.04 & 0.00 & 0.00 & 0.25 & 0.00\\
\hline
Leave 100 out CV & 0.50 & 0.03 & 0.00 & 0.00 & 0.21 & 0.00\\
\hline
30\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.24 & 0.00\\
\hline
50\% Holdout & 0.50 & 0.03 & 0.00 & 0.00 & 0.20 & 0.00\\
\hline
Bootstrap & 0.50 & 0.01 & 0.00 & 0.00 & 0.08 & 0.00\\
\hline
\end{tabular}
\end{table}

\begin{table}

\caption{Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance. The true value of the AUC is 0.5 and the true value of the Log OR is 0.0. Estimates are based on 1000 replicates of the numerical experiment. In each replicate,  N = 200; p = 10 . CV = Cross validation. }
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR & Bias OR\\
\hline
Resubstitution & 0.63 & 0.03 & 0.13 & 0.78 & 0.27 & 0.78\\
\hline
Partial CV & 0.50 & 0.08 & 0.00 & -0.01 & 0.35 & -0.01\\
\hline
Partial Holdout & 0.50 & 0.06 & 0.00 & -0.02 & 0.43 & -0.02\\
\hline
Partial Resubstitution & 0.60 & 0.04 & 0.10 & 0.56 & 0.31 & 0.56\\
\hline
Pre-validation & 0.50 & 0.06 & 0.00 & 0.02 & 0.40 & 0.02\\
\hline
Leave 10 out CV & 0.50 & 0.08 & 0.00 & 0.00 & 0.34 & 0.00\\
\hline
Leave 100 out CV & 0.50 & 0.07 & 0.00 & 0.00 & 0.39 & 0.00\\
\hline
30\% Holdout & 0.50 & 0.08 & 0.00 & 0.02 & 0.53 & 0.02\\
\hline
50\% Holdout & 0.50 & 0.06 & 0.00 & 0.01 & 0.43 & 0.01\\
\hline
Bootstrap & 0.50 & 0.04 & 0.00 & -0.01 & 0.24 & -0.01\\
\hline
\end{tabular}
\end{table}

\begin{table}

\caption{Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance. The true value of the AUC is 0.5 and the true value of the Log OR is 0.0. Estimates are based on 1000 replicates of the numerical experiment. In each replicate,  N = 200; p = 100 . CV = Cross validation. }
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR & Bias OR\\
\hline
Resubstitution & 0.84 & 0.03 & 0.34 & 2.32 & 0.37 & 2.32\\
\hline
Partial CV & 0.71 & 0.04 & 0.21 & 1.21 & 0.34 & 1.21\\
\hline
Partial Holdout & 0.67 & 0.06 & 0.17 & 1.00 & 0.45 & 1.00\\
\hline
Partial Resubstitution & 0.72 & 0.04 & 0.22 & 1.40 & 0.33 & 1.40\\
\hline
Pre-validation & 0.50 & 0.07 & 0.00 & 0.01 & 0.41 & 0.01\\
\hline
Leave 10 out CV & 0.50 & 0.07 & 0.00 & 0.00 & 0.41 & 0.00\\
\hline
Leave 100 out CV & 0.50 & 0.07 & 0.00 & 0.03 & 0.43 & 0.03\\
\hline
30\% Holdout & 0.49 & 0.08 & -0.01 & -0.04 & 0.52 & -0.04\\
\hline
50\% Holdout & 0.50 & 0.06 & 0.00 & 0.00 & 0.44 & 0.00\\
\hline
Bootstrap & 0.50 & 0.04 & 0.00 & -0.01 & 0.21 & -0.01\\
\hline
\end{tabular}
\end{table}

\begin{table}

\caption{Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance. The true value of the AUC is 0.5 and the true value of the Log OR is 0.0. Estimates are based on 1000 replicates of the numerical experiment. In each replicate,  N = 200; p = 500 . CV = Cross validation. }
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR & Bias OR\\
\hline
Resubstitution & 0.92 & 0.02 & 0.42 & 3.42 & 0.58 & 3.42\\
\hline
Partial CV & 0.83 & 0.03 & 0.33 & 2.07 & 0.30 & 2.07\\
\hline
Partial Holdout & 0.76 & 0.07 & 0.26 & 1.77 & 0.54 & 1.77\\
\hline
Partial Resubstitution & 0.74 & 0.03 & 0.24 & 1.62 & 0.30 & 1.62\\
\hline
Pre-validation & 0.50 & 0.08 & 0.00 & 0.00 & 0.48 & 0.00\\
\hline
Leave 10 out CV & 0.50 & 0.07 & 0.00 & 0.02 & 0.45 & 0.02\\
\hline
Leave 100 out CV & 0.50 & 0.08 & 0.00 & 0.00 & 0.45 & 0.00\\
\hline
30\% Holdout & 0.50 & 0.08 & 0.00 & 0.00 & 0.54 & 0.00\\
\hline
50\% Holdout & 0.50 & 0.05 & 0.00 & 0.03 & 0.45 & 0.03\\
\hline
Bootstrap & 0.50 & 0.03 & 0.00 & 0.00 & 0.17 & 0.00\\
\hline
\end{tabular}
\end{table}

\begin{table}

\caption{Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) and the log odds ratio (OR) in the setting where a dataset is used to both develop the signature and evaluate its performance. The true value of the AUC is 0.5 and the true value of the Log OR is 0.0. Estimates are based on 1000 replicates of the numerical experiment. In each replicate,  N = 200; p = 5000 . CV = Cross validation. }
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r}
\hline
Approach & mean AUC & std.dev AUC & Bias AUC & mean OR & std.dev OR & Bias OR\\
\hline
Resubstitution & 0.97 & 0.01 & 0.47 & 4.86 & 0.77 & 4.86\\
\hline
Partial CV & 0.90 & 0.02 & 0.40 & 2.75 & 0.27 & 2.75\\
\hline
Partial Holdout & 0.80 & 0.06 & 0.30 & 2.53 & 0.68 & 2.53\\
\hline
Partial Resubstitution & 0.76 & 0.03 & 0.26 & 1.62 & 0.33 & 1.62\\
\hline
Pre-validation & 0.50 & 0.06 & 0.00 & 0.00 & 0.43 & 0.00\\
\hline
Leave 10 out CV & 0.50 & 0.07 & 0.00 & 0.00 & 0.44 & 0.00\\
\hline
Leave 100 out CV & 0.49 & 0.09 & -0.01 & -0.05 & 0.51 & -0.05\\
\hline
30\% Holdout & 0.50 & 0.06 & 0.00 & 0.03 & 0.55 & 0.03\\
\hline
50\% Holdout & 0.51 & 0.06 & 0.01 & 0.06 & 0.43 & 0.06\\
\hline
Bootstrap & 0.50 & 0.02 & 0.00 & -0.01 & 0.13 & -0.01\\
\hline
\end{tabular}
\end{table}

\clearpage

\includegraphics{supplement_files/figure-latex/plotsauc-1.pdf}
\clearpage

\includegraphics{supplement_files/figure-latex/plotsauc-2.pdf}
\clearpage

\includegraphics{supplement_files/figure-latex/plotsauc-3.pdf}
\clearpage

\includegraphics{supplement_files/figure-latex/plotsauc-4.pdf}
\clearpage

\includegraphics{supplement_files/figure-latex/plotsauc-5.pdf}
\clearpage

\includegraphics{supplement_files/figure-latex/plotsauc-6.pdf}
\clearpage

\includegraphics{supplement_files/figure-latex/plotsauc-7.pdf}
\clearpage

\includegraphics{supplement_files/figure-latex/plots2-1.pdf} \clearpage

\includegraphics{supplement_files/figure-latex/plots2-2.pdf} \clearpage

\includegraphics{supplement_files/figure-latex/plots2-3.pdf} \clearpage

\includegraphics{supplement_files/figure-latex/plots2-4.pdf} \clearpage

\includegraphics{supplement_files/figure-latex/plots2-5.pdf} \clearpage

\includegraphics{supplement_files/figure-latex/plots2-6.pdf} \clearpage

\includegraphics{supplement_files/figure-latex/plots2-7.pdf} \clearpage

\includegraphics{supplement_files/figure-latex/overall-1.pdf}

\includegraphics{supplement_files/figure-latex/unnamed-chunk-1-1.pdf}

\bibliography{paper.bib}

\end{document}

### 2016-07-05 Code/new sims/newsims.R

@@ -1,0 +1,78 @@
+
+library(ggplot2)
+
+parse_sims <- function(file, path = ""){
+
+  colnms <- c('holdout.5.AUC', 'holdout.5.OR', 'holdout.3.AUC', 'holdout.3.OR', 'cv.10.AUC', 'cv.10.OR', 'cv.preval.AUC', 'cv.preval.OR', 'cv.100.AUC', 'cv.100.OR', 'boot.AUC', 'boot.OR', 'zhu.hold.AUC', 'zhu.hold.OR', 'zhu.hold2.AUC', 'zhu.hold2.OR', 'zhu.cv.AUC', 'zhu.cv.OR', 'naive.AUC', 'naive.OR')
+
+
+  titles <- c("N = 1000; p = 10",
+              "N = 1000; p = 100",
+              "N = 1000; p = 500",
+              "N = 200; p = 10",
+              "N = 200; p = 100",
+              "N = 200; p = 500",
+              "N = 200; p = 5000"
+              )
+  names(titles) <- c("results100010.txt" , "results1000100.txt",
+                     "results1000500.txt", "results20010.txt" ,
+                     "results200100.txt",  "results200500.txt", "results2005000.txt")
+  tite <- titles[file]
+
+  cvres <- read.table(paste0(path, file))
+  colnames(cvres) <- colnms
+  cvlong <- do.call("rbind", lapply(1:ncol(cvres), function(i){
+
+    cl <- cvres[, i]
+    splow <- strsplit(colnames(cvres)[i], ".", fixed = TRUE)
+    nm <- paste(unlist(sapply(splow, function(s) rev(rev(s)[-1]))), collapse = ".")
+    cls <- sapply(splow, function(s) rev(s)[1])
+    data.frame(value = cvres[, i], stat = cls, scen = nm, stringsAsFactors = FALSE)
+
+  }))
+  cvlong$value[cvlong$stat == "OR"] <- exp(cvlong$value[cvlong$stat == "OR"])
+
+  cvlong$scen <- factor(cvlong$scen, levels = c("naive", "zhu.cv", "zhu.hold", "zhu.hold2",
+                                                "cv.preval", "cv.10", "cv.100", "holdout.3",
+                                                "holdout.5", "boot"), ordered = TRUE)
+  cvlong$N <- as.numeric(strsplit(tite, "( = |;)")[[1]][2])
+  cvlong$p <- as.numeric(strsplit(tite, "( = |;)")[[1]][4])
+
+  labs <- c("Resubstitution", "Partial CV",
[…]

### 2016-07-05 Documents/paper-revised-resub.Rmd
revisions

--- 2016-05-18 Documents/paper-revised.Rmd
@@
 
-The focus of this paper is how to obtain a valid estimate of signature performance as quantified by a metric $phi$, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
+The focus of this paper is how to obtain a valid estimate of signature performance as quantified by a metric $\phi$, that is, a good estimate of $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
 The principles discussed apply no matter what signature development method is used.
@@
 To illustrate some of the different properties of these estimates and how they help to avoid overfitting, we conduct a limited simulation study.
-Data were generated with 1000 observations, each with a binary outcome $Y$ with prevalence 0.3, and 500 mutually independent features sampled from the standard normal distribution.
+Data were generated with $N = 200$ or 1000 observations, each with a binary outcome $Y$ with prevalence 0.3, and $p = 10, 100, 500,$ or $5000$ mutually independent features sampled from the standard normal distribution.
 This is the null case where no features are associated with $Y$.
 The signature development procedure entails a feature selection step, in which each feature is regressed against $Y$ in a univariate logistic regression model.
-The 25 features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines the final signature.
+The 25 (10 in the case where $p = 10$) features with the smallest p-values are selected for inclusion in a multivariable logistic regression model which defines the final signature.
 
@@
 
-```{r cvsims, fig.cap = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) in the setting where a dataset is used to both develop the signature and evaluate its performance.
-The violin plots show mirrored density estimates for the AUC for 1000 replicates of the numerical experiment.
-In each replicate, there are 1000 observations and 500 features.
-The true value of the AUC is 0.5.
-CV = Cross validation.
-\\label{fig1}"}
+```{r simset, include = FALSE}
 load("../Code/cvsim-result.RData")
@@
 names(labs) <- levels(cvlong$scen)
+```
 
-ggplot(subset(cvlong, stat == "AUC"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") + 
+```{r cvsims, fig.cap = "Comparison of different approaches to estimating the Area Under the ROC Curve (AUC) in the setting where a dataset is used to both develop the signature and evaluate its performance.
+The violin plots show mirrored density estimates for the AUC for 1000 replicates of the numerical experiment.
+In each replicate, there are $N$ observations and $p$ features.
+The true value of the AUC is 0.5.
+CV = Cross validation.
+\\label{fig1}", fig.height = 8.5}
+source("../Code/new sims/newsims.R")
+
+files <- grep("*.txt", list.files("../Code/new sims/"), value = TRUE)
+results <- lapply(files, parse_sims)
+
+alldat <- NULL
+for(i in results){
+
+  alldat <- rbind(alldat, i$datalong)
+
+}
+
+ labs <- c("Resubstitution", "Partial CV",
+            "Partial Holdout", "Partial Resubstitution",
+            "Pre-validation", "Leave 10 out CV", "Leave 100 out CV",
+            "30% Holdout", "50% Holdout", "Bootstrap")
+
+  names(labs) <- levels(alldat$scen)
+
+ggplot(subset(alldat, stat == "AUC"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") +
     theme_bw(base_size = 13, base_family = "serif") + coord_flip() +
     scale_x_discrete("Estimation Approach", labels = labs) +
-  ylab("Area Under the ROC Curve")
+    ylab("Area Under the ROC Curve") + facet_grid(p ~ N, labeller = "label_both")
+
 ```
@@
 The violin plots show mirrored density estimates for the log OR for 1000 replicates of the numerical experiment.
-In each replicate, there are 1000 observations and 500 features.
+In each replicate, there are $N$ observations and $p$ features.
 The true value of the OR is 1.0.
@@
 
-ggplot(subset(cvlong, stat == "OR"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") + 
+alldat$value[alldat$stat == "OR"] <- exp(alldat$value[alldat$stat == "OR"])
+
+ggplot(subset(alldat, stat == "OR"), aes(y = value, x = scen)) + geom_violin(fill = "grey60") +
     theme_bw(base_size = 13, base_family = "serif") + coord_flip() +
-  scale_x_discrete("Estimation Approach", labels = labs) + scale_y_log10(breaks = c(.5, 1, 2, 4)) +
-  ylab("Odds Ratio (log scale)")
+    scale_x_discrete("Estimation Approach", labels = labs) + scale_y_log10(breaks = c(.1, .5, 1, 4, 10, 20)) +
+    ylab("Odds Ratio (log scale)") + facet_grid(p ~ N, labeller = "label_both")
+
 ```
@@
 CV = Cross validation.
+Additional results for different $N$ and $p$ are in the appendix.
 ")
@@
 
-Not surprisingly, the resubstitution estimates are optimistically biased: the naive resubstitution estimate of the AUC is 44% larger than it should be and the log OR estimate is over 2 times higher than it should be, on average.
+Not surprisingly, the resubstitution estimates are optimistically biased: for $N = 1000$ and $p = 500$ the naive resubstitution estimate of the AUC is 44% larger than it should be and the log OR estimate is over 2 times higher than it should be, on average.
 Partial resubstitution, partial holdout, and partial cross-validation estimates do not ameliorate the bias very much.
@@
 However, here we see that these versions are still severely biased and should not be reported as valid assessments of the performance of biomarker signatures.
+
+It is evident in Figures \ref{fig1} and \ref{fig2} that the amount of bias tends to increase as the number of features increases.
+Even with only 10 features to select from, the resubstitution estimates are still biased.
+A larger sample size does not help with the bias, though it does improve precision of the performance estimates.
+With $N = 200$ and $p = 5000$, which is the situation closest to our real data example, we see that the resubstitution estimate indicates nearly perfect prediction performance, when in fact there is no signal in the data.
 

### 2016-07-05 Documents/paper-revised-resub.tex
revisions

--- 2016-05-18 Documents/paper-revised.tex
@@
     \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
   \else
@@
   \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
   \newcommand{\euro}{€}
@@
 \usepackage[margin=1in]{geometry}
+\usepackage{hyperref}
+\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
+\hypersetup{unicode=true,
+            pdftitle={Issues in developing multivariable molecular signatures for guiding clinical care decisions},
+            pdfauthor={Michael C Sachs and Lisa M McShane},
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
 \usepackage{natbib}
@@
 \usepackage{longtable,booktabs}
-\usepackage{graphicx}
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={Michael C Sachs and Lisa M McShane},
-            pdftitle={Issues in developing multivariable molecular signatures for guiding clinical care decisions},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
 \setlength{\parindent}{0pt}
@@
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \setcounter{secnumdepth}{0}
@@
   \postdate{\par}
-  \date{May 18, 2016}
+  \date{June 30, 2016}
 
@@
 
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
 \begin{document}
-
 \maketitle
@@
 The focus of this paper is how to obtain a valid estimate of signature
-performance as quantified by a metric \(phi\), that is, a good estimate
+performance as quantified by a metric \(\phi\), that is, a good estimate
 of \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\). The principles discussed
@@
 how they help to avoid overfitting, we conduct a limited simulation
-study. Data were generated with 1000 observations, each with a binary
-outcome \(Y\) with prevalence 0.3, and 500 mutually independent features
-sampled from the standard normal distribution. This is the null case
-where no features are associated with \(Y\). The signature development
-procedure entails a feature selection step, in which each feature is
-regressed against \(Y\) in a univariate logistic regression model. The
-25 features with the smallest p-values are selected for inclusion in a
-multivariable logistic regression model which defines the final
-signature.
+study. Data were generated with \(N = 200\) or 1000 observations, each
+with a binary outcome \(Y\) with prevalence 0.3, and
+\(p = 10, 100, 500,\) or \(5000\) mutually independent features sampled
+from the standard normal distribution. This is the null case where no
+features are associated with \(Y\). The signature development procedure
+entails a feature selection step, in which each feature is regressed
+against \(Y\) in a univariate logistic regression model. The 25 (10 in
+the case where \(p = 10\)) features with the smallest p-values are
+selected for inclusion in a multivariable logistic regression model
+which defines the final signature.
 
@@
 \centering
-\includegraphics{paper-revised_files/figure-latex/cvsims-1.pdf}
+\includegraphics{paper-revised-resub_files/figure-latex/cvsims-1.pdf}
 \caption{Comparison of different approaches to estimating the Area Under
@@
 show mirrored density estimates for the AUC for 1000 replicates of the
-numerical experiment. In each replicate, there are 1000 observations and
-500 features. The true value of the AUC is 0.5. CV = Cross validation.
-\label{fig1}}
+numerical experiment. In each replicate, there are \(N\) observations
+and \(p\) features. The true value of the AUC is 0.5. CV = Cross
+validation. \label{fig1}}
 \end{figure}
@@
 \centering
-\includegraphics{paper-revised_files/figure-latex/cvsims2-1.pdf}
+\includegraphics{paper-revised-resub_files/figure-latex/cvsims2-1.pdf}
 \caption{Comparison of different approaches to estimating the odds ratio
@@
 density estimates for the log OR for 1000 replicates of the numerical
-experiment. In each replicate, there are 1000 observations and 500
+experiment. In each replicate, there are \(N\) observations and \(p\)
 features. The true value of the OR is 1.0. CV = Cross validation.
@@
 experiment. In each replicate, there are 1000 observations and 500
-features. CV = Cross validation.}\tabularnewline
+features. CV = Cross validation. Additional results for different \(N\)
+and \(p\) are in the appendix.}\tabularnewline
 \toprule
@@
 Not surprisingly, the resubstitution estimates are optimistically
-biased: the naive resubstitution estimate of the AUC is 44\% larger than
-it should be and the log OR estimate is over 2 times higher than it
-should be, on average. Partial resubstitution, partial holdout, and
-partial cross-validation estimates do not ameliorate the bias very much.
-Investigators are tempted to use partial holdout estimates as it is more
-convenient to not have to carry along a large number of feature
-measurements into the modeling part of the signature development
-process, and they may think that the resulting performance estimates are
-still close to valid, as only half of the data are used to form the
-estimates. However, here we see that these versions are still severely
-biased and should not be reported as valid assessments of the
-performance of biomarker signatures.
+biased: for \(N = 1000\) and \(p = 500\) the naive resubstitution
+estimate of the AUC is 44\% larger than it should be and the log OR
+estimate is over 2 times higher than it should be, on average. Partial
+resubstitution, partial holdout, and partial cross-validation estimates
+do not ameliorate the bias very much. Investigators are tempted to use
+partial holdout estimates as it is more convenient to not have to carry
+along a large number of feature measurements into the modeling part of
+the signature development process, and they may think that the resulting
+performance estimates are still close to valid, as only half of the data
+are used to form the estimates. However, here we see that these versions
+are still severely biased and should not be reported as valid
+assessments of the performance of biomarker signatures.
+
+It is evident in Figures \ref{fig1} and \ref{fig2} that the amount of
+bias tends to increase as the number of features increases. Even with
+only 10 features to select from, the resubstitution estimates are still
+biased. A larger sample size does not help with the bias, though it does
+improve precision of the performance estimates. With \(N = 200\) and
+\(p = 5000\), which is the situation closest to our real data example,
+we see that the resubstitution estimate indicates nearly perfect
+prediction performance, when in fact there is no signal in the data.
 
@@
 \centering
-\includegraphics{paper-revised_files/figure-latex/examppresub-1.pdf}
+\includegraphics{paper-revised-resub_files/figure-latex/examppresub-1.pdf}
 \caption{Comparison of survival by gene expression based risk signature.
@@
 \centering
-\includegraphics{paper-revised_files/figure-latex/inter1-1.pdf}
+\includegraphics{paper-revised-resub_files/figure-latex/inter1-1.pdf}
 \caption{Survival curves comparing the treatment effect by gene
@@
 \renewcommand\refname{References}
-\bibliography{paper}
+\bibliography{paper.bib}
 

### 2016-07-05 Documents/paper-revised-round2.Rmd
revisions

--- 2016-07-05 Documents/paper-revised-resub.Rmd
@@
     - \usepackage{setspace}
+    - \usepackage{tablefootnote}
     - \doublespacing
-    - \usepackage{lineno}
-    - \linenumbers
 fontsize: 11pt
@@
 
-Omics technologies that generate a large amount of molecular data characterizing patient biospecimens have the potential to provide information about patients' disease characteristics above and beyond standard clinical and pathological features.
-By combining the information from a large number of molecular features into a multivariable model or decision algorithm, called a biomarker signature, there is the opportunity to identify distinct subgroups of patients for whom treatment decisions can be personalized.
-A biomarker signature can guide decisions to treat or not to treat and help identify the patients who are most likely to have a more favororable outcome or benefit from a particular therapy.
-The key challenge is to combine features from a high dimensional molecular assay to derive a signature with good clinical performance and appropriately characterize its performance.
-The inappropriate practice of using overlapping data to both build a signature and evaluate its performance can lead to severe over-optimism bias in performance estimates.
+Omics technologies that generate a large amount of molecular data characterizing biospecimens have the potential to provide information about patients' disease characteristics beyond standard clinical features.
+By combining information from a large number of features into a multivariable model, called a biomarker signature, there is the opportunity to identify distinct subgroups of patients for whom treatment decisions can be personalized.
+The key challenge is to derive a signature with good performance and appropriately characterize its performance.
 We summarize the key statistical issues and methods for developing and validating biomarker signatures, using examples from the literature for illustration.
@@
 
-```{r setup, include = FALSE}
+```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
 library(dplyr)
@@
 
-The key challenge we address in this paper is how to combine feature mneasurements generated by a high dimensional multiplex molecular assay to derive a signature that is fit for a specified clinical and additionally provide valid estimates of its performance characteristics.
+The key challenge we address in this paper is how to combine feature mneasurements generated by a high dimensional multiplex molecular assay to derive a signature that is fit for a specified clinical purpose and additionally provide valid estimates of its performance characteristics.
 A signature that has clinical value must have statistical performance (as quantified by one or more metrics) that is fit for the clinical context, and it must provide information that can be acted upon clinically to provide some benefit to a patient.
@@
 
-Let $\phi: \mathcal{D} \times \mathcal{S} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$, such as predictive accuracy, mean squared error, or area under the receiver operating characteristic (ROC) curve (AUC).
-This could also be a measure of association, such as an odds ratio, hazard ratio, or log-rank statistic.
+Let $\phi: \mathcal{D} \times \mathcal{S} \mapsto \mathbb{R}$ denote the statistic that quantifies the performance of the function $f$.
 This is a function of both $f$ and $S$.
-We are interested in estimating $E_\mathcal{P}[\phi_{f^*}(S)]$, which is the expected error under the data generation mechanism (distribution $\mathcal{P}$), for a particular $f^* \in \mathcal{F}$.
+We are interested in estimating $E_\mathcal{P}[\phi_{f^*}(S)]$, which is the expected performance under the data generation mechanism (distribution $\mathcal{P}$), for a particular $f^* \in \mathcal{F}$ when $f^*$ is applied to $X$ for the purpose of predicting $Y$.
+This expectation is with respect to $\mathcal{P}$, the distribution of the random variable $S$.
 This allows us to understand how the signature will perform on future observations generated from $\mathcal{P}$.
+The difficulty in estimating this quantity is that generally only a single sample from $\mathcal{P}$ is available to both develop $f^*$ and to estimate its expected performance $E_\mathcal{P}[\phi_{f^*}(S)]$.
+
 We may also be interested in estimating $E_\mathcal{P}[\phi_f(S)]$ **for all** $f \in \mathcal{F}$, which is the generalization performance for $f$ generated using mechanism $\mathcal{F}$.
-This doesn't guide outside researchers as to which specific $f$ to use, yet it is useful for development because it tells us how much signal is in the data.
+This doesn't guide outside researchers as to which specific $f$ to use, yet it is useful for development because it tells us how much signal is in the data and how well it can be captured with $f \in \mathcal{F}$.
 As shorthand we will write this as $E_\mathcal{P}[\phi_\mathcal{F}(S)]$.
@@
 
-Our focus here will be mainly on the development of signatures using supervised methods, meaning methods that use information on an outcome variable it is desired to predict during the signature development process.
+Our focus here will be mainly on the development of signatures using supervised methods, meaning methods that use information about an outcome variable that it is desired to predict.
 Regression methods are a commonly used class of supervised methods.
@@
 In the case of Oncotype DX, the outcome $Y$ was time to distant recurrence of breast cancer and the feature data were gene expression values measured in breast tumor specimens.
-A common supervised approach to identifying treatment-selection signatures is to use regression techniques to estimate a signature that has a strong interaction with a particular treatment.
+Regression methods can also be used to develop treatment-selection (also called predictive) signatures that exhibit a strong interaction with a particular treatment.
+
 Supervised methods are contrasted with unsupervised methods which use only the feature data to derive a signature.
 An example of a signature developed by unsupervised methods is the signature that identifies biological subclasses of diffuse large B-cell lymphoma, which were originally developed using clustering methods and subsequently were found to be associated with prognosis [@alizadeh2000distinct].
-It is possible, and quite common in high-dimensional data settings, to combine multiple approaches when estimated a signature.
+It is possible, and quite common in high-dimensional data settings, to combine multiple approaches when estimating a signature.
 For instance, a data-reduction step by variable selection or clustering may be performed before doing regression analysis on the resulting components.
@@
 It is not trivial to conduct this evaluation in a valid manner when there is a limited data set available to both define a signature and assess its performance.
-We illustrate the potential for bias to creep into the performance estimation and review some strategies to avoid the pitfalls that lead to these biases.
+We illustrate the potential for bias to creep into the performance estimation and review some strategies to avoid the pitfalls that lead to those biases.
 We evaluate various strategies through a simulation study and compare the performance of some of those strategies using a real data exmaple involving the development of a signature for use in making treatment decisions for patients with lung cancer.
+All simulations and analyses were conducted using `R` version 3.2.4 [@arr].
 
@@
 
-The main goal of interest here is to estimate $E_\mathcal{P}[\phi_\mathcal{F}(S)]$, the expected value of a given performance metric on future observations for $f \in \mathcal{F}$.
-This can be estimated with the in-sample empirical estimate: $\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)$ for a particular $f$.
+$E_\mathcal{P}[\phi_\mathcal{F}(S)]$ can be estimated with the in-sample empirical estimate: $\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)$ for a particular $f$.
 However, if $S$ is used to estimate $f$ then the estimate will be biased due to overfitting, that is, $|E_\mathcal{P}[\phi_f(S)] - \hat{E}[\phi_f(S)]|$ will be large.
@@
 
-Typically a performance metric $\phi$ used during signature development is degined to quantify either calibration or discrimination, or some combination of the two.
-Calibration assesses how well the signature based predictions compare to the observed outcomes.
-Discrimination assesses how well the signature distinguishes between groups of patients who do or do not experience an event.
+Typically a performance metric $\phi$ used during signature development is designed to quantify either calibration or discrimination, or some combination of the two.
+Calibration assesses how well the signature-based predictions compare to the observed outcomes.
+Discrimination assesses how well the signature distinguishes between groups of patients who do or do not experience an event or have a particular phenotype.
 To assess discrimination, a different metric $\phi$ may be used, such as the area under the ROC curve.
 Measures of association such as the odds ratio, hazard ratio, or difference in survival probabilities are also commonly used performance metrics, although their value for biomarker signatures has been debated [@pepe2004limitations].
-In @zhu2010prognostic, $\phi$ was the hazard ratio comparing the high risk and low risk groups.
+In @zhu2010prognostic, $\phi$ was the hazard ratio comparing disease-specific survival between the groups identified as high risk and low risk.
 The risk groups were determined by the signature $f$, which was estimated using the JBL.10 observation arm data.
-The signature was then applied to the same data that were used to build it to define risk groups, producing a the hazard ratio estimate of 15 for a prognostic effect in the observation arm.
+The signature was then applied to the same data that were used to build it to define risk groups, producing the hazard ratio estimate of 15 for the prognostic effect comparing signature-defined risk groups in the observation arm.
 Subsequent evaluations of the signature's performance on independent data resulted in hazard ratio estimates in the 2 - 3 range, substantially smaller than the original estimate of 15.
-Next we reanalyze the JBR.10 data and illustrate some methods to avoid bias in evaluation of signature performance.
+This disparity is due, at least in part, to the bias in the original hazard ratio estimate of 15 resulting from reuse of the data used to build the signature.
+Next we discuss some methods to avoid bias in evaluation of signature performance.
 
@@
 
-A traditional method to avoid overfitting is the split sample approach.
+A traditional method to avoid overfitting is the split sample holdout approach.
 First, randomly partition $S$ into the training sample $S_t$ and the holdout sample $S_h$ with sample sizes $n_t$ and $n_h$, respectively.
@@
 The specific form of $f_t$ that is fixed using $S_t$ can be reported as the function for others to use, therefore the aforementioned estimator applies for that specific $f_t$.
-The drawback of the split-sample approach is that the performance of $f_t$ is likely to be inferior to the performance of a function $f^*$ derived using the same approach applied to the entire data set.
+Alternatively, $f^*$, the signature derived using the entire set $S$ can be reported, accompanied by the estimate of the performance of $f_t$.
+The drawback of this approach is that the performance of $f_t$ is likely to be inferior to the performance of a function $f^*$ derived using the same approach applied to the entire data set.
 @dobbin2007sample investigate how to choose the sample size n so that there is high probability that the signature developed will have expected performance within some specified margin of the the optimal signature that could be developed if the development data set had infinite sample size.
@@
 Another approach to avoid overfitting is cross-validation, which is a resampling based approach.
-For a fixed integer $k$, which can be between 1 and $n$, we randomly partition the full data set $S$ into subgroups of size $k$ (assume for simplicity here that $n$ is evenly divisible by $k$ so that $n/k$ subgroups are formed).
+For a fixed integer $k$, which can be between 1 and $n$, we randomly partition the full data set $S$ into subgroups of size $k$ (assume for simplicity here that $n$ is evenly divisible by $k$ so that $K = n/k$ subgroups are formed).
 For each $k$, $f_{-k}$ is estimated and fixed by appling $\mathcal{F}$ on $S_{-k}$ which is the subset of $S$ that is disjoint from $S_k$.
 Then, we get an estimate $\hat{E}[\phi_{f_{-k}}(S_k)]$ which is an estimate of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
-This process is repeated $K$ times to yield $K$ estimates and then these are averaged to obtain the performance estimate.
-This process is called "leave $k$ out" or "$n/k$ fold" cross-validation.
+Then one iterates through the $K$ subgroups to yield $K$ estimates and then these are averaged to obtain the performance estimate.
+This process is called "leave $k$ out" or "$K$ fold" cross-validation.
 Note that for each subgroup in the partition, we obtain a new estimate of $f$, therefore we are only estimating $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$ and not the performance for a single specified signature.
 Typically, if a specific form for $f$ is desired, it would be estimated using the entire dataset $S$ to yield $f^*$ as above.
-Importantly one should report the cross-validated estimate just described and not the naive esitimate that would be obtained by substituting the full data $S$ into the performance metric calculation for the signature $f^*$.
-The problematic aspects of such naive "resubstitution estimates" of performance are discuss in more detail below.
+Importantly one should report the cross-validated estimate just described and not the naïve estimate that would be obtained by substituting the full data $S$ into the performance metric calculation for the signature $f^*$.
+The problematic aspects of such naive "resubstitution estimates" of performance are discussed in more detail below.
 
@@
 $$
-where $\hat{E}[\phi_{f}(S)]$ is the naive estimate of $\phi_f$ using the entire dataset.
+where $\hat{E}[\phi_{f}(S)]$ is the naïve resubstitution estimate of $\phi_f$ using the entire dataset.
 
@@
 With pre-validation, instead of computing the statistic $\phi$ for each of the held-out subsets ($S_{-b}$ for the bootstrap or $S_{k}$ for cross-validation), the fitted signature $\hat{f}(X_i)$ is estimated for $X_i \in S_{-b}$ where $\hat{f}$ is estimated using $S_{b}$.
-This process is repeated to obtain a set of pre-validated signature estimates $\hat{f}$ which are then used to calculate $\phi$.
-For single-step split sample training and validation, this process is equivalent to what is described above.
-For cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on each of the smaller held-out datasets.
+This process is repeated to obtain a set of pre-validated signature estimates $\{\hat{f}_i, i = 1, \ldots, n\}$.
+Then some measure of the association between these signature estimates and the withheld true outcomes, $Y_i, i = 1, \ldots, n$, is estimated to serve as a measure of the signature performance.
+The association measure $\phi$ could be estimated as a coefficient in a linear, Cox proportional hazards, or logistic regression model, with or without adjustment for other covariates, as appropriate for the situation.
+The idea behind this pre-validation approach is that if the signature has good performance, then it should produce values that are strongly associated with the true outcomes.
+In contrast to cross-validation and the bootstrap, this process avoids the problem of having too few cases to estimate the statistic $\phi$ on each of the smaller held-out datasets.
+
+These resampling-based procedures described above provide unbiased estimates of predictive accuracy for the procedure, that is $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$, as opposed to the accuracy for a particular signature that others can use.
+Often, we will provide the signature developed on the entire dataset as the one for future use, as using the full data will estimate signature coefficients with the most precision.
+Still, it is important to avoid the temptation to report the estimate of performance on this full data signature, as it can be severly biased.
 
@@
 \hline
-Partial Holdout & Select features on full dataset $S$. Split data into $S_t$ and $S_h$. Build model on $S_t$ using only features pre-selected from full dataset $S$. Then test that model on $S_h$ \\
+Partial Holdout & Select features on full dataset $S$. Split data into $S_t$ and $S_h$. Build model on $S_t$ using only features pre-selected from full dataset $S$. Then test that model on $S_h$. \\
 Partial CV  & Select features on full dataset $S$.
@@
 \\
-Naive Resubstitution & Select features on full dataset $S$ and build model on $S$ using features pre-selected from $S$. Then test that model on $S_h$. \\
+Naïve Resubstitution & Select features on full dataset $S$ and build model on $S$ using features pre-selected from $S$. Then test that model on $S$. \\
 Partial Resubstitution & Split data into $S_t$ and $S_h$. Select features on $S_t$. Build model on $S_t$ using only features pre-selected from $S_t$. Then test that model on the full dataset $S$. \\
@@
 The biased approaches are described in Table \ref{descript}.
-Two of the biased approaches use the full sample to select the features, followed by fitting the multivariable model on the holdout subset.
+Two of the biased approaches use the full sample to select the features, followed by fitting the multivariable model on holdout subsets.
 This is referred to as "parital holdout" or "partial CV" when using split-sample holdout or cross-validation as the validation step, respectively.
-We also implemented the naïve resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is developed on a training subset but then evaluated on the combined training adn holdout data sets.
+We also implemented the naïve resubstitution approach, wherein the model is trained and evaluated on the same dataset, and the partial resubstitution approach wherein the model is developed on a training subset but then evaluated on the combined training and holdout data sets.
 Our main interest is in comparing the bias and variance of the resulting estimates of $E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]$.
-In our simulation, we look at two different performance metrics, the area under the ROC curve (AUC) and the odds ratio for the outcome comparing the signature groups.
+In our simulation, we look at two different performance metrics, the area under the ROC curve (AUC) and the odds ratio for the binary outcome comparing the signature groups.
 
@@
 CV = Cross validation.
-\\label{fig1}", fig.height = 8.5}
+\\label{fig1}", fig.height = 8.25}
 source("../Code/new sims/newsims.R")
@@
 files <- grep("*.txt", list.files("../Code/new sims/"), value = TRUE)
-results <- lapply(files, parse_sims)
+results <- lapply(files, parse_sims, path = "../Code/new sims/")
 
@@
 CV = Cross validation.
-\\label{fig2} "}
+\\label{fig2} ", fig.height = 8.25}
 
@@
     theme_bw(base_size = 13, base_family = "serif") + coord_flip() +
-    scale_x_discrete("Estimation Approach", labels = labs) + scale_y_log10(breaks = c(.1, .5, 1, 4, 10, 20)) +
+    scale_x_discrete("Estimation Approach", labels = labs) + scale_y_log10(breaks = c(.5, 2, 5, 20, 80)) +
     ylab("Odds Ratio (log scale)") + facet_grid(p ~ N, labeller = "label_both")
@@
 
-@zhu2010prognostic present results that mainly focus on the association of their signature with outcome, albeit a less than ideal measure of discrimination ability.
+@zhu2010prognostic present results that mainly focus on the association of their signature with outcome as quantified by a hazard ratio, albeit a less than ideal measure of discrimination ability.
 They demonstrate that the two risk subgroups predicted by their signature (high risk and low risk) have separation in their survival curves and that the hazard ratio for their signature is large and significant even when adjusting for other risk factors.
@@
 
-The signature development procedure was described in the introduction.
+Now recall the general framework for signature development that was described in the introduction.
 There is both a feature selection step, and a multivariable estimation step.
@@
 Discrimination of the signature is assessed using the concordance statistic as implemented in the \texttt{survival} package in \texttt{R} [@survival].
-To paraphrase the help file: this is defined as the probability of agreement for any two randomly chosen observations, which in this case means that the observation with the shorter survival time also has the larger signature value.
+To paraphrase the help file: the concordance statistic is defined as the probability of agreement for any two randomly chosen observations, which in this case means that the observation with the shorter survival time also has the larger signature value.
 This is similar to an interpretation of the AUC for binary data.
 
-First we fit the signature using the entire observation cohort (n = 62).
+First we fit the signature using the entire observation arm (n = 62).
 The signature was then evaluated on the same dataset.
-The survival plot on the left side of Figure \ref{fig2} shows extreme separation between the two risk groups (HR = `r round(exp(evalall[2]))`, p < 0.001), consistent with the reported JBL.10 signature, and the estimated concordance is `r round(evalall[1], 2)`.
+The survival plot on the left side of Figure \ref{fig3} shows extreme separation between the two risk groups (HR = `r round(exp(evalall[2]))`, p < 0.001), consistent with the reported JBL.10 signature, and the estimated concordance is `r round(evalall[1], 2)`.
 After correctly accounting for the selection process, our estimates of association and discrimination are much less impressive.
@@
 
-The right plot in Figure \ref{fig2} shows the survival curves for the two risk groups using the pre-validated estimates of the risk score.
+The right plot in Figure \ref{fig3} shows the survival curves for the two risk groups using the pre-validated estimates of the risk score.
 We partitioned the 62 observations into 8 groups of 6 and 2 groups of 7.
@@
 The left plot shows the resubstitution estimate, while the right plot shows the pre-validated estimate.
-\\label{fig2}"}
+\\label{fig3}"}
 trtgrp <- subset(gdat, Post.Surgical.Treatment == "ACT")
@@
 The top set of plots shows the partial resubstitution based signature and the bottom row shows the pre-validated signature estimates.
-\\label{fig3}"}
+\\label{fig4}"}
 cldat.trtplot$Risk.fac <- factor(cldat.trtplot$riskgrp, levels = c(FALSE, TRUE), labels = c("Low risk", "High risk"))
@@
 HRmary <- data.frame(Comparison = "High Risk vs Low Risk", `Hazard Ratio` = round(exp(fit.train$coefficients[1]), 1), 
-           `95% CI` = pCI(exp(confint(fit.train)[1, ])), `Adjusted p` = myP(summary(fit.train)$coefficients[1, 5]), 
+           `95% CI^*` = pCI(exp(confint(fit.train)[1, ])), `Adjusted p^*` = myP(summary(fit.train)$coefficients[1, 5]), 
            check.names = FALSE, stringsAsFactors = FALSE)
@@
 
-We also show plots to assess the ability of the signature to be useful for treatment selection (Figure \ref{fig3}).
+We also show plots to assess the ability of the signature to be useful for treatment selection (Figure \ref{fig4}).
 These plots show the survival curves comparing treatment arms grouped in panels by the risk score.
 As described by @polley2013statistical, the idea is to determine whether the treatment is beneficial in one group and not benficial or harmful in another group, indicating that different treatment decisions would be made based on the signature.
-On one hand, the overfit signature shows dramatic differences in treatment efficacy between the low risk and high risk groups.
+On one hand, the overfit resubstitution signature shows dramatic differences in treatment efficacy between the low risk and high risk groups.
 In fact it appears that the treatment is harmful in the low risk group, but highly beneficial in the high risk group.
@@
 This suggests that the dramatic predictive value of the signature was merely an artifact of overfitting the signature to the OBS arm data.
-@simon2011re explain the flaws in this type of approach in which one attempts to develop a signature for treatment selection by developing a prognostic signature on data from the control arm of a trial and then applying that signature to both the control and experimental arm data to establish predictiveness.
+
+The @zhu2010prognostic paper used the approach of identifying a signature using the control arm of the trial, evaluating it using the combined control and treatment arms, and then hoping that the signature would be useful for treatment selection.
+This approach has been shown to be invalid [@simon2011re] for identifying a predictive signature.
+Indeed we show that the resubstitution aspect of the evaluation likely led to overstatements of the size of the signature’s purported predictive effect for therapy selection.
 
@@
 Using the prevalidated signature, the effect estimates are much smaller and of small magnitude in comparison to the standard clinical features.
-Note that the inference from the pre-validated model is not exactly correct either, because the resampling procedure induces a correlation among the observations (the prevalidated signature values are estimated from heavily overlapping data).
+Note that the inferences from the pre-validated model (confidence intervals and p-values) are not exactly correct either, because the resampling procedure induces a correlation among the observations (the prevalidated signature values are estimated from heavily overlapping data).
 Despite this, these results are unimpressive and would not be considered promising for clinical use.
@@
           ci.custom = lapply(list(fit.train, fit.preval, fit.train.inter, fit.preval.inter), confint),
-          title = "Complete Results of Multivariable Regression Models. Entries are hazard ratios and 95 percent confidence intervals.  \\label{multimodel}")
+          title = "Complete Results of Multivariable Regression Models. Entries are hazard ratios and 95 percent confidence intervals.  \\label{multimodel}", 
+          notes = "CI's and p-values based on incorrect inference.")
 ```
@@
 
-Examples of flawed approaches such as naive resubstitution, partial resubstitution, partial holdout, and partial CV remain very common in the published literature more than a decade after many high-dimensional molecular assays were first introduced.
-All performance metrics, whether they assess calibration or discrimination or something else, are subject to bias in estimation due to overfitting.
-Strategies to avoid this type of bias are well-studied in the statistical literature and here we have demonstrated how they can be implemented in a real clinical example.
+Molecular signatures are being increasingly integrated into the practice of medicine to personalize clinical care decisions.
+Unlike new drugs which go through a rigorous development process that includes conduct of prospective clinical trials and review by regulatory authorities, the evidence base for molecular signatures is most often derived from retrospectively conducted studies using stored specimens and published literature is relied upon for evidence to assess the worth of a signature.
+More than a decade after many high-dimensional molecular assays were first introduced, there are many publications reporting signatures that have not been adequately evaluated for their performance.
+Many published reports contain highly exaggerated claims of signature performance due to use of inappropriate methods to develop them and assess their performance.
+It is important that use of inappropriate methods is recognized so that flawed or poor performing signatures are not adopted into clinical practice [@mcshane2013criteria; @mcshane2013criteriab].
+
+As an aid to researchers, journal editors, journal reviewers, regulatory and funding agencies, we have attempted to describe in this paper some of the flawed signature development and validation approaches that we have commonly encountered and to explain their impact on reported performance metrics.
+Many researchers have embarked on studies intended to develop and assess performance of molecular signatures, but many have not had much prior experience in this analytical area.
+The problematic approaches highlighted in the examples and simulations included naïve resubstitution, partial resubstitution, partial holdout, and partial CV.
+The simulations demonstrated how even in a simple null situation, the over-optimism bias in performance estimates is quite substantial, possibly enough to persuade some to adopt these signatures for clinical care prematurely.
+The magnitude of the bias could be even greater with more modest samples sizes or more convoluted signature development methods as are typical in the published literature.
+In our simulations and data example, we used a simple variable selection procedure follow by multivariable regression.
+More complex statistical methods for developing signatures such as the lasso, random forests, and others are not immune to this type of bias.
+In fact, more complex methods often involve tuning parameters, which give more opportunities for bias due to overfitting.
+Strategies to avoid this type of bias are well-studied in the statistical literature and here we demonstrated how they can be implemented in a real clinical example.
+
 The ability to perform the kind of in-depth data reanalysis we performed on the example in this paper is usually limited by restrictions on data access and insufficient description of analysis methods.
-Sadly, reports of strong associations with overfit biomarker signatures are all too common in the medical literature.
-The amount of bias that is possible is not known and can be difficult to decipher based on study reports.
-It is imperative that investigators and journal editors take overfitting bias seriously to ensure that signature performance claims are valid.
-Unfortunately this often results in study reports that are far less optimistic than usual.
-
-The @zhu2010prognostic paper used the approach of identifying a signature using the control arm of the trial, evaluating it using the combined control and treatment arms, and then hoping that the signature would be useful for treatment selection.
-This approach has been shown to be invalid [@simon2011re] for identifying a predictive signature, and indeed we show that the resubstitution aspect of the evaluation likely led to overstatments of the size of the signature effect.
-Properly doing cross-validation for estimating the calibration of the signature and doing pre-validation for assessment of the discrimination of the signature show that the associations are much more modest.
-
-When we assess the signature on the independent treatment arm, we see that there is no significant difference between the risk groups.
-This raises the question of whether it is appropriate to use non-random splits of the dataset in order to obtain valid estimates of the calibration or discrimination.
-Instead of treatment arms, we could imagine a large multi-center study, and we could split the data into disjoint subgroups based on the center.
-Specifically, suppose we split the development dataset $S$ into $S_1$ and $S_2$ according to a discrete covariate $X$ that takes on levels 1 or 2.
-The we develop a signature $f_{S_1}$ using $S_1$ and evaluate it on $S_2$ by estimating $\hat{E}[\phi_{f_{S_1}}(S_2)]$.
-This is an estimate not of $E_\mathcal{P}[\phi_f(S)]$, but rather $E_{\mathcal{P}_2}[\phi_f(S)]$, where $\mathcal{P}_2$ is the distribution for the sub-population with $X = 2$ that $S_2$ is a sample from.
-This estimate would only be recommended if the signature is intended for use in that specific subpopulation, and if that were the case, then it doesn't make much sense to develop the signature using the subpopulation $\mathcal{P}_1$.
-If these groups differed substantially, then we would not expect the signature to perform adequately.
-The differences in performance may depend on many factors, including how $X$ is associated with the distribution of the features and the outcome.
-If a signature becomes broadly used in clinics, these center-to-center differences would be important to assess as a part of signature efficacy surveillance.
+It is imperative that investigators and other stakeholders such as journal editors, funders, and regulators take overfitting bias seriously to ensure that signature performance claims are valid.
+This will require a concerted effort to ensure that reviewers are sufficiently informed of acceptable approaches to signature development and evaluation, data are made available for independent verification, and methods are reported in complete and transparent fashion.
 

### 2016-07-05 Documents/paper-revised-round2.tex
revisions

--- 2016-07-05 Documents/paper-revised-resub.tex
@@
   \posttitle{\par}
-  \author{Michael C Sachs and Lisa M McShane}
+  \author{Michael C Sachs$^*$ and Lisa M McShane}
   \preauthor{\centering\large\emph}
@@
   \postdate{\par}
-  \date{June 30, 2016}
+  \date{July 05, 2016}
 
 \usepackage{setspace}
+\usepackage{tablefootnote}
 \doublespacing
-\usepackage{lineno}
-\linenumbers
 
@@
 
+\begin{center}
+\textit{Biostatistics Branch, Biometric Research Program \\
+National Cancer Institute \\
+9609 Medical Center Drive, MSC 9735 \\
+Bethesda, MD 20892-9735 \\
+Phone: 240-276-6004 \\
+Email: michael.sachs@nih.gov \\
+*: Corresponding Author}
+
+\end{center}
+
 \section{Abstract}\label{abstract}
@@
 Omics technologies that generate a large amount of molecular data
-characterizing patient biospecimens have the potential to provide
-information about patients' disease characteristics above and beyond
-standard clinical and pathological features. By combining the
-information from a large number of molecular features into a
-multivariable model or decision algorithm, called a biomarker signature,
-there is the opportunity to identify distinct subgroups of patients for
-whom treatment decisions can be personalized. A biomarker signature can
-guide decisions to treat or not to treat and help identify the patients
-who are most likely to have a more favororable outcome or benefit from a
-particular therapy. The key challenge is to combine features from a high
-dimensional molecular assay to derive a signature with good clinical
-performance and appropriately characterize its performance. The
-inappropriate practice of using overlapping data to both build a
-signature and evaluate its performance can lead to severe over-optimism
-bias in performance estimates. We summarize the key statistical issues
-and methods for developing and validating biomarker signatures, using
-examples from the literature for illustration.
+characterizing biospecimens have the potential to provide information
+about patients' disease characteristics beyond standard clinical
+features. By combining information from a large number of features into
+a multivariable model, called a biomarker signature, there is the
+opportunity to identify distinct subgroups of patients for whom
+treatment decisions can be personalized. The key challenge is to derive
+a signature with good performance and appropriately characterize its
+performance. We summarize the key statistical issues and methods for
+developing and validating biomarker signatures, using examples from the
+literature for illustration. \\
+
+\textbf{keywords}: biomarkers; biomarker signatures; personalized medicine
+
+\clearpage
 
@@
 mneasurements generated by a high dimensional multiplex molecular assay
-to derive a signature that is fit for a specified clinical and
+to derive a signature that is fit for a specified clinical purpose and
 additionally provide valid estimates of its performance characteristics.
@@
 Let \(\phi: \mathcal{D} \times \mathcal{S} \mapsto \mathbb{R}\) denote
-the statistic that quantifies the performance of the function \(f\),
-such as predictive accuracy, mean squared error, or area under the
-receiver operating characteristic (ROC) curve (AUC). This could also be
-a measure of association, such as an odds ratio, hazard ratio, or
-log-rank statistic. This is a function of both \(f\) and \(S\). We are
-interested in estimating \(E_\mathcal{P}[\phi_{f^*}(S)]\), which is the
-expected error under the data generation mechanism (distribution
-\(\mathcal{P}\)), for a particular \(f^* \in \mathcal{F}\). This allows
-us to understand how the signature will perform on future observations
-generated from \(\mathcal{P}\). We may also be interested in estimating
-\(E_\mathcal{P}[\phi_f(S)]\) \textbf{for all} \(f \in \mathcal{F}\),
-which is the generalization performance for \(f\) generated using
-mechanism \(\mathcal{F}\). This doesn't guide outside researchers as to
-which specific \(f\) to use, yet it is useful for development because it
-tells us how much signal is in the data. As shorthand we will write this
-as \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\).
+the statistic that quantifies the performance of the function \(f\).
+This is a function of both \(f\) and \(S\). We are interested in
+estimating \(E_\mathcal{P}[\phi_{f^*}(S)]\), which is the expected
+performance under the data generation mechanism (distribution
+\(\mathcal{P}\)), for a particular \(f^* \in \mathcal{F}\) when \(f^*\)
+is applied to \(X\) for the purpose of predicting \(Y\). This
+expectation is with respect to \(\mathcal{P}\), the distribution of the
+random variable \(S\). This allows us to understand how the signature
+will perform on future observations generated from \(\mathcal{P}\). The
+difficulty in estimating this quantity is that generally only a single
+sample from \(\mathcal{P}\) is available to both develop \(f^*\) and to
+estimate its expected performance \(E_\mathcal{P}[\phi_{f^*}(S)]\).
+
+We may also be interested in estimating \(E_\mathcal{P}[\phi_f(S)]\)
+\textbf{for all} \(f \in \mathcal{F}\), which is the generalization
+performance for \(f\) generated using mechanism \(\mathcal{F}\). This
+doesn't guide outside researchers as to which specific \(f\) to use, yet
+it is useful for development because it tells us how much signal is in
+the data and how well it can be captured with \(f \in \mathcal{F}\). As
+shorthand we will write this as \(E_\mathcal{P}[\phi_\mathcal{F}(S)]\).
 
@@
 Our focus here will be mainly on the development of signatures using
-supervised methods, meaning methods that use information on an outcome
-variable it is desired to predict during the signature development
-process. Regression methods are a commonly used class of supervised
-methods. Oncotype DX \citep{paik2004multigene} is an example of a
-prognostic signature used for clinical decision making in early stage
-breast cancer that was developed using supervised methods. In the case
-of Oncotype DX, the outcome \(Y\) was time to distant recurrence of
-breast cancer and the feature data were gene expression values measured
-in breast tumor specimens. A common supervised approach to identifying
-treatment-selection signatures is to use regression techniques to
-estimate a signature that has a strong interaction with a particular
-treatment. Supervised methods are contrasted with unsupervised methods
-which use only the feature data to derive a signature. An example of a
-signature developed by unsupervised methods is the signature that
-identifies biological subclasses of diffuse large B-cell lymphoma, which
-were originally developed using clustering methods and subsequently were
+supervised methods, meaning methods that use information about an
+outcome variable that it is desired to predict. Regression methods are a
+commonly used class of supervised methods. Oncotype DX
+\citep{paik2004multigene} is an example of a prognostic signature used
+for clinical decision making in early stage breast cancer that was
+developed using supervised methods. In the case of Oncotype DX, the
+outcome \(Y\) was time to distant recurrence of breast cancer and the
+feature data were gene expression values measured in breast tumor
+specimens. Regression methods can also be used to develop
+treatment-selection (also called predictive) signatures that exhibit a
+strong interaction with a particular treatment.
+
+Supervised methods are contrasted with unsupervised methods which use
+only the feature data to derive a signature. An example of a signature
+developed by unsupervised methods is the signature that identifies
+biological subclasses of diffuse large B-cell lymphoma, which were
+originally developed using clustering methods and subsequently were
 found to be associated with prognosis \citep{alizadeh2000distinct}. It
 is possible, and quite common in high-dimensional data settings, to
-combine multiple approaches when estimated a signature. For instance, a
+combine multiple approaches when estimating a signature. For instance, a
 data-reduction step by variable selection or clustering may be performed
@@
 performance estimation and review some strategies to avoid the pitfalls
-that lead to these biases. We evaluate various strategies through a
+that lead to those biases. We evaluate various strategies through a
 simulation study and compare the performance of some of those strategies
 using a real data exmaple involving the development of a signature for
-use in making treatment decisions for patients with lung cancer.
+use in making treatment decisions for patients with lung cancer. All
+simulations and analyses were conducted using \texttt{R} version 3.2.4
+\citep{arr}.
 
@@
 
-The main goal of interest here is to estimate
-\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\), the expected value of a given
-performance metric on future observations for \(f \in \mathcal{F}\).
-This can be estimated with the in-sample empirical estimate:
+\(E_\mathcal{P}[\phi_\mathcal{F}(S)]\) can be estimated with the
+in-sample empirical estimate:
 \(\hat{E}[\phi_f(S)] = \frac{1}{n}\sum_{i=1}^n\phi_f(s_i)\) for a
@@
 Typically a performance metric \(\phi\) used during signature
-development is degined to quantify either calibration or discrimination,
-or some combination of the two. Calibration assesses how well the
-signature based predictions compare to the observed outcomes.
+development is designed to quantify either calibration or
+discrimination, or some combination of the two. Calibration assesses how
+well the signature-based predictions compare to the observed outcomes.
 Discrimination assesses how well the signature distinguishes between
-groups of patients who do or do not experience an event. To assess
-discrimination, a different metric \(\phi\) may be used, such as the
-area under the ROC curve. Measures of association such as the odds
-ratio, hazard ratio, or difference in survival probabilities are also
-commonly used performance metrics, although their value for biomarker
-signatures has been debated \citep{pepe2004limitations}. In
-\citet{zhu2010prognostic}, \(\phi\) was the hazard ratio comparing the
-high risk and low risk groups. The risk groups were determined by the
-signature \(f\), which was estimated using the JBL.10 observation arm
-data. The signature was then applied to the same data that were used to
-build it to define risk groups, producing a the hazard ratio estimate of
-15 for a prognostic effect in the observation arm. Subsequent
-evaluations of the signature's performance on independent data resulted
-in hazard ratio estimates in the 2 - 3 range, substantially smaller than
-the original estimate of 15. Next we reanalyze the JBR.10 data and
-illustrate some methods to avoid bias in evaluation of signature
-performance.
+groups of patients who do or do not experience an event or have a
+particular phenotype. To assess discrimination, a different metric
+\(\phi\) may be used, such as the area under the ROC curve. Measures of
+association such as the odds ratio, hazard ratio, or difference in
+survival probabilities are also commonly used performance metrics,
+although their value for biomarker signatures has been debated
+\citep{pepe2004limitations}. In \citet{zhu2010prognostic}, \(\phi\) was
+the hazard ratio comparing disease-specific survival between the groups
+identified as high risk and low risk. The risk groups were determined by
+the signature \(f\), which was estimated using the JBL.10 observation
+arm data. The signature was then applied to the same data that were used
+to build it to define risk groups, producing the hazard ratio estimate
+of 15 for the prognostic effect comparing signature-defined risk groups
+in the observation arm. Subsequent evaluations of the signature's
+performance on independent data resulted in hazard ratio estimates in
+the 2 - 3 range, substantially smaller than the original estimate of 15.
+This disparity is due, at least in part, to the bias in the original
+hazard ratio estimate of 15 resulting from reuse of the data used to
+build the signature. Next we discuss some methods to avoid bias in
+evaluation of signature performance.
 
@@
 
-A traditional method to avoid overfitting is the split sample approach.
-First, randomly partition \(S\) into the training sample \(S_t\) and the
-holdout sample \(S_h\) with sample sizes \(n_t\) and \(n_h\),
-respectively. Then, \(S_h\) is hidden from the analyst while
+A traditional method to avoid overfitting is the split sample holdout
+approach. First, randomly partition \(S\) into the training sample
+\(S_t\) and the holdout sample \(S_h\) with sample sizes \(n_t\) and
+\(n_h\), respectively. Then, \(S_h\) is hidden from the analyst while
 \(\mathcal{F}\) is applied to \(S_t\) to estimate the signature function
@@
 use, therefore the aforementioned estimator applies for that specific
-\(f_t\). The drawback of the split-sample approach is that the
+\(f_t\). Alternatively, \(f^*\), the signature derived using the entire
+set \(S\) can be reported, accompanied by the estimate of the
+performance of \(f_t\). The drawback of this approach is that the
 performance of \(f_t\) is likely to be inferior to the performance of a
@@
 subgroups of size \(k\) (assume for simplicity here that \(n\) is evenly
-divisible by \(k\) so that \(n/k\) subgroups are formed). For each
+divisible by \(k\) so that \(K = n/k\) subgroups are formed). For each
 \(k\), \(f_{-k}\) is estimated and fixed by appling \(\mathcal{F}\) on
@@
 Then, we get an estimate \(\hat{E}[\phi_{f_{-k}}(S_k)]\) which is an
-estimate of \(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). This process is
-repeated \(K\) times to yield \(K\) estimates and then these are
-averaged to obtain the performance estimate. This process is called
-``leave \(k\) out'' or ``\(n/k\) fold'' cross-validation. Note that for
-each subgroup in the partition, we obtain a new estimate of \(f\),
+estimate of \(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\). Then one
+iterates through the \(K\) subgroups to yield \(K\) estimates and then
+these are averaged to obtain the performance estimate. This process is
+called ``leave \(k\) out'' or ``\(K\) fold'' cross-validation. Note that
+for each subgroup in the partition, we obtain a new estimate of \(f\),
 therefore we are only estimating
@@
 \(f^*\) as above. Importantly one should report the cross-validated
-estimate just described and not the naive esitimate that would be
+estimate just described and not the naïve estimate that would be
 obtained by substituting the full data \(S\) into the performance metric
 calculation for the signature \(f^*\). The problematic aspects of such
-naive ``resubstitution estimates'' of performance are discuss in more
+naive ``resubstitution estimates'' of performance are discussed in more
 detail below.
@@
 \hat{E}^*[\phi_{\mathcal{F}}(S)] = .368 \hat{E}[\phi_{f}(S)] + 0.632 \hat{E}[\phi_{f_b}(S_{-b})],
-\] where \(\hat{E}[\phi_{f}(S)]\) is the naive estimate of \(\phi_f\)
-using the entire dataset.
+\] where \(\hat{E}[\phi_{f}(S)]\) is the naïve resubstitution estimate
+of \(\phi_f\) using the entire dataset.
 
@@
 where \(\hat{f}\) is estimated using \(S_{b}\). This process is repeated
-to obtain a set of pre-validated signature estimates \(\hat{f}\) which
-are then used to calculate \(\phi\). For single-step split sample
-training and validation, this process is equivalent to what is described
-above. For cross-validation and the bootstrap, this process avoids the
-problem of having too few cases to estimate the statistic \(\phi\) on
-each of the smaller held-out datasets.
+to obtain a set of pre-validated signature estimates
+\(\{\hat{f}_i, i = 1, \ldots, n\}\). Then some measure of the
+association between these signature estimates and the withheld true
+outcomes, \(Y_i, i = 1, \ldots, n\), is estimated to serve as a measure
+of the signature performance. The association measure \(\phi\) could be
+estimated as a coefficient in a linear, Cox proportional hazards, or
+logistic regression model, with or without adjustment for other
+covariates, as appropriate for the situation. The idea behind this
+pre-validation approach is that if the signature has good performance,
+then it should produce values that are strongly associated with the true
+outcomes. In contrast to cross-validation and the bootstrap, this
+process avoids the problem of having too few cases to estimate the
+statistic \(\phi\) on each of the smaller held-out datasets.
+
+These resampling-based procedures described above provide unbiased
+estimates of predictive accuracy for the procedure, that is
+\(E_{\mathcal{P}}[\phi_{\mathcal{F}}(S)]\), as opposed to the accuracy
+for a particular signature that others can use. Often, we will provide
+the signature developed on the entire dataset as the one for future use,
+as using the full data will estimate signature coefficients with the
+most precision. Still, it is important to avoid the temptation to report
+the estimate of performance on this full data signature, as it can be
+severly biased.
 
@@
 sample to select the features, followed by fitting the multivariable
-model on the holdout subset. This is referred to as ``parital holdout''
-or ``partial CV'' when using split-sample holdout or cross-validation as
+model on holdout subsets. This is referred to as ``parital holdout'' or
+``partial CV'' when using split-sample holdout or cross-validation as
 the validation step, respectively. We also implemented the naïve
@@
 model is developed on a training subset but then evaluated on the
-combined training adn holdout data sets. Our main interest is in
+combined training and holdout data sets. Our main interest is in
 comparing the bias and variance of the resulting estimates of
@@
 at two different performance metrics, the area under the ROC curve (AUC)
-and the odds ratio for the outcome comparing the signature groups.
+and the odds ratio for the binary outcome comparing the signature
+groups.
 
@@
 \centering
-\includegraphics{paper-revised-resub_files/figure-latex/cvsims-1.pdf}
+\includegraphics{paper-revised-round2_files/figure-latex/cvsims-1.pdf}
 \caption{Comparison of different approaches to estimating the Area Under
@@
 \centering
-\includegraphics{paper-revised-resub_files/figure-latex/cvsims2-1.pdf}
+\includegraphics{paper-revised-round2_files/figure-latex/cvsims2-1.pdf}
 \caption{Comparison of different approaches to estimating the odds ratio
@@
 \citet{zhu2010prognostic} present results that mainly focus on the
-association of their signature with outcome, albeit a less than ideal
-measure of discrimination ability. They demonstrate that the two risk
-subgroups predicted by their signature (high risk and low risk) have
-separation in their survival curves and that the hazard ratio for their
-signature is large and significant even when adjusting for other risk
-factors. They do not directly address calibration, that is, whether
-their signature accurately predicts survival times.
+association of their signature with outcome as quantified by a hazard
+ratio, albeit a less than ideal measure of discrimination ability. They
+demonstrate that the two risk subgroups predicted by their signature
+(high risk and low risk) have separation in their survival curves and
+that the hazard ratio for their signature is large and significant even
+when adjusting for other risk factors. They do not directly address
+calibration, that is, whether their signature accurately predicts
+survival times.
 
@@
 
-The signature development procedure was described in the introduction.
-There is both a feature selection step, and a multivariable estimation
-step. This results in a continuous signature which is the linear
-predictor of a Cox regression model. The signature is dichotomized by
-selecting the cutoff that yields the most significant log-rank statistic
-for comparing the resulting risk groups. Discrimination of the signature
-is assessed using the concordance statistic as implemented in the
-\texttt{survival} package in \texttt{R} \citep{survival}. To paraphrase
-the help file: this is defined as the probability of agreement for any
-two randomly chosen observations, which in this case means that the
-observation with the shorter survival time also has the larger signature
-value. This is similar to an interpretation of the AUC for binary data.
+Now recall the general framework for signature development that was
+described in the introduction. There is both a feature selection step,
+and a multivariable estimation step. This results in a continuous
+signature which is the linear predictor of a Cox regression model. The
+signature is dichotomized by selecting the cutoff that yields the most
+significant log-rank statistic for comparing the resulting risk groups.
+Discrimination of the signature is assessed using the concordance
+statistic as implemented in the \texttt{survival} package in \texttt{R}
+\citep{survival}. To paraphrase the help file: the concordance statistic
+is defined as the probability of agreement for any two randomly chosen
+observations, which in this case means that the observation with the
+shorter survival time also has the larger signature value. This is
+similar to an interpretation of the AUC for binary data.
 
-First we fit the signature using the entire observation cohort (n = 62).
+First we fit the signature using the entire observation arm (n = 62).
 The signature was then evaluated on the same dataset. The survival plot
-on the left side of Figure \ref{fig2} shows extreme separation between
+on the left side of Figure \ref{fig3} shows extreme separation between
 the two risk groups (HR = 20, p \textless{} 0.001), consistent with the
@@
 
-The right plot in Figure \ref{fig2} shows the survival curves for the
+The right plot in Figure \ref{fig3} shows the survival curves for the
 two risk groups using the pre-validated estimates of the risk score. We
@@
 \centering
-\includegraphics{paper-revised-resub_files/figure-latex/examppresub-1.pdf}
+\includegraphics{paper-revised-round2_files/figure-latex/examppresub-1.pdf}
 \caption{Comparison of survival by gene expression based risk signature.
 The left plot shows the resubstitution estimate, while the right plot
-shows the pre-validated estimate. \label{fig2}}
+shows the pre-validated estimate. \label{fig3}}
 \end{figure}
@@
 \centering
-\includegraphics{paper-revised-resub_files/figure-latex/inter1-1.pdf}
+\includegraphics{paper-revised-round2_files/figure-latex/inter1-1.pdf}
 \caption{Survival curves comparing the treatment effect by gene
@@
 resubstitution based signature and the bottom row shows the
-pre-validated signature estimates. \label{fig3}}
+pre-validated signature estimates. \label{fig4}}
 \end{figure}
@@
 \toprule
-Method & Comparison & Hazard Ratio & 95\% CI & Adjusted p\tabularnewline
+Method & Comparison & Hazard Ratio & 95\% CI$^*$ & Adjusted
+p$^*$\tabularnewline
 \midrule
@@
 \bottomrule
+& \multicolumn{4}{r}{CI's and p-values based on incorrect inference.} \\
 \end{longtable}
@@
 We also show plots to assess the ability of the signature to be useful
-for treatment selection (Figure \ref{fig3}). These plots show the
+for treatment selection (Figure \ref{fig4}). These plots show the
 survival curves comparing treatment arms grouped in panels by the risk
@@
 treatment decisions would be made based on the signature. On one hand,
-the overfit signature shows dramatic differences in treatment efficacy
-between the low risk and high risk groups. In fact it appears that the
-treatment is harmful in the low risk group, but highly beneficial in the
-high risk group. The prevalidated signature on the other hand, shows
-differences that are much less dramatic. It appears that the treatment
-is mildly beneficial in both groups, possibly to a higher degree in the
-high risk group. This suggests that the dramatic predictive value of the
-signature was merely an artifact of overfitting the signature to the OBS
-arm data. \citet{simon2011re} explain the flaws in this type of approach
-in which one attempts to develop a signature for treatment selection by
-developing a prognostic signature on data from the control arm of a
-trial and then applying that signature to both the control and
-experimental arm data to establish predictiveness.
+the overfit resubstitution signature shows dramatic differences in
+treatment efficacy between the low risk and high risk groups. In fact it
+appears that the treatment is harmful in the low risk group, but highly
+beneficial in the high risk group. The prevalidated signature on the
+other hand, shows differences that are much less dramatic. It appears
+that the treatment is mildly beneficial in both groups, possibly to a
+higher degree in the high risk group. This suggests that the dramatic
+predictive value of the signature was merely an artifact of overfitting
+the signature to the OBS arm data.
+
+The \citet{zhu2010prognostic} paper used the approach of identifying a
+signature using the control arm of the trial, evaluating it using the
+combined control and treatment arms, and then hoping that the signature
+would be useful for treatment selection. This approach has been shown to
+be invalid \citep{simon2011re} for identifying a predictive signature.
+Indeed we show that the resubstitution aspect of the evaluation likely
+led to overstatements of the size of the signature's purported
+predictive effect for therapy selection.
 
@@
 much smaller and of small magnitude in comparison to the standard
-clinical features. Note that the inference from the pre-validated model
-is not exactly correct either, because the resampling procedure induces
-a correlation among the observations (the prevalidated signature values
-are estimated from heavily overlapping data). Despite this, these
-results are unimpressive and would not be considered promising for
-clinical use.
+clinical features. Note that the inferences from the pre-validated model
+(confidence intervals and p-values) are not exactly correct either,
+because the resampling procedure induces a correlation among the
+observations (the prevalidated signature values are estimated from
+heavily overlapping data). Despite this, these results are unimpressive
+and would not be considered promising for clinical use.
 
@@
 \textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\
+ & \multicolumn{4}{r}{CI's and p-values based on incorrect inference.} \\
 \end{tabular}
@@
 
-Examples of flawed approaches such as naive resubstitution, partial
-resubstitution, partial holdout, and partial CV remain very common in
-the published literature more than a decade after many high-dimensional
-molecular assays were first introduced. All performance metrics, whether
-they assess calibration or discrimination or something else, are subject
-to bias in estimation due to overfitting. Strategies to avoid this type
-of bias are well-studied in the statistical literature and here we have
-demonstrated how they can be implemented in a real clinical example. The
-ability to perform the kind of in-depth data reanalysis we performed on
-the example in this paper is usually limited by restrictions on data
-access and insufficient description of analysis methods. Sadly, reports
-of strong associations with overfit biomarker signatures are all too
-common in the medical literature. The amount of bias that is possible is
-not known and can be difficult to decipher based on study reports. It is
-imperative that investigators and journal editors take overfitting bias
-seriously to ensure that signature performance claims are valid.
-Unfortunately this often results in study reports that are far less
-optimistic than usual.
+Molecular signatures are being increasingly integrated into the practice
+of medicine to personalize clinical care decisions. Unlike new drugs
+which go through a rigorous development process that includes conduct of
+prospective clinical trials and review by regulatory authorities, the
+evidence base for molecular signatures is most often derived from
+retrospectively conducted studies using stored specimens and published
+literature is relied upon for evidence to assess the worth of a
+signature. More than a decade after many high-dimensional molecular
+assays were first introduced, there are many publications reporting
+signatures that have not been adequately evaluated for their
+performance. Many published reports contain highly exaggerated claims of
+signature performance due to use of inappropriate methods to develop
+them and assess their performance. It is important that use of
+inappropriate methods is recognized so that flawed or poor performing
+signatures are not adopted into clinical practice
+\citep{mcshane2013criteria, mcshane2013criteriab}.
 
-The \citet{zhu2010prognostic} paper used the approach of identifying a
-signature using the control arm of the trial, evaluating it using the
-combined control and treatment arms, and then hoping that the signature
-would be useful for treatment selection. This approach has been shown to
-be invalid \citep{simon2011re} for identifying a predictive signature,
-and indeed we show that the resubstitution aspect of the evaluation
-likely led to overstatments of the size of the signature effect.
-Properly doing cross-validation for estimating the calibration of the
-signature and doing pre-validation for assessment of the discrimination
-of the signature show that the associations are much more modest.
+As an aid to researchers, journal editors, journal reviewers, regulatory
+and funding agencies, we have attempted to describe in this paper some
+of the flawed signature development and validation approaches that we
+have commonly encountered and to explain their impact on reported
+performance metrics. Many researchers have embarked on studies intended
+to develop and assess performance of molecular signatures, but many have
+not had much prior experience in this analytical area. The problematic
+approaches highlighted in the examples and simulations included naïve
+resubstitution, partial resubstitution, partial holdout, and partial CV.
+The simulations demonstrated how even in a simple null situation, the
+over-optimism bias in performance estimates is quite substantial,
+possibly enough to persuade some to adopt these signatures for clinical
+care prematurely. The magnitude of the bias could be even greater with
+more modest samples sizes or more convoluted signature development
+methods as are typical in the published literature. In our simulations
+and data example, we used a simple variable selection procedure follow
+by multivariable regression. More complex statistical methods for
+developing signatures such as the lasso, random forests, and others are
+not immune to this type of bias. In fact, more complex methods often
+involve tuning parameters, which give more opportunities for bias due to
+overfitting. Strategies to avoid this type of bias are well-studied in
+the statistical literature and here we demonstrated how they can be
+implemented in a real clinical example.
 
-When we assess the signature on the independent treatment arm, we see
-that there is no significant difference between the risk groups. This
-raises the question of whether it is appropriate to use non-random
-splits of the dataset in order to obtain valid estimates of the
-calibration or discrimination. Instead of treatment arms, we could
-imagine a large multi-center study, and we could split the data into
-disjoint subgroups based on the center. Specifically, suppose we split
-the development dataset \(S\) into \(S_1\) and \(S_2\) according to a
-discrete covariate \(X\) that takes on levels 1 or 2. The we develop a
-signature \(f_{S_1}\) using \(S_1\) and evaluate it on \(S_2\) by
-estimating \(\hat{E}[\phi_{f_{S_1}}(S_2)]\). This is an estimate not of
-\(E_\mathcal{P}[\phi_f(S)]\), but rather
-\(E_{\mathcal{P}_2}[\phi_f(S)]\), where \(\mathcal{P}_2\) is the
-distribution for the sub-population with \(X = 2\) that \(S_2\) is a
-sample from. This estimate would only be recommended if the signature is
-intended for use in that specific subpopulation, and if that were the
-case, then it doesn't make much sense to develop the signature using the
-subpopulation \(\mathcal{P}_1\). If these groups differed substantially,
-then we would not expect the signature to perform adequately. The
-differences in performance may depend on many factors, including how
-\(X\) is associated with the distribution of the features and the
-outcome. If a signature becomes broadly used in clinics, these
-center-to-center differences would be important to assess as a part of
-signature efficacy surveillance.
+The ability to perform the kind of in-depth data reanalysis we performed
+on the example in this paper is usually limited by restrictions on data
+access and insufficient description of analysis methods. It is
+imperative that investigators and other stakeholders such as journal
+editors, funders, and regulators take overfitting bias seriously to
+ensure that signature performance claims are valid. This will require a
+concerted effort to ensure that reviewers are sufficiently informed of
+acceptable approaches to signature development and evaluation, data are
+made available for independent verification, and methods are reported in
+complete and transparent fashion.
 
@@
 All analysis code and the source files for this manuscript are available
-from the authors' webpage.
+from the authors' webpage: https://github.com/sachsmc/signature-tutorial
 

### 2016-07-05 Code/11-simfun.R
new simulations R code

@@ -1,0 +1,262 @@
+
+auc <- function(lll, ddd){
+
+  ROCR::performance(ROCR::prediction(lll, ddd), "auc")@y.values[[1]]
+
+}
+
+or <- function(sc, resp){
+
+  sc01 <- sc > median(sc, na.rm = TRUE)
+  tab <- table(sc01, resp) + .5
+  log(tab[1 ,1]) + log(tab[2, 2]) - (log(tab[1, 2]) + log(tab[2, 1]))
+
+}
+
+createnosignaldata <- function(n, d){
+
+  data.frame(Y = rbinom(n, 1, p = .3), X = matrix(rnorm(n * d), nrow = n))
+
+}
+
+selectvars <- function(data0){
+
+  pees <- sapply(data0[, -1], function(v1){
+
+    fit0 <- glm(data0$Y ~ v1, family = "binomial", model = FALSE, y = FALSE)
+    summary(fit0)$coefficients[2, 4]
+
+  })
+
+  order(pees)[1:25]
+
+}
+
+runclassifier <- function(data0, selected){
+
+  dat2 <- data0[, c(1, selected + 1)]
+  fit1 <- glm(Y ~ ., data = dat2, family = "binomial", model = FALSE, y = FALSE)
+
+  fit1
+
+}
+
+fitclassifier <- function(fit1, data1){
+
+  scores <- predict(fit1, newdata = data1, type = "response")
+  response <- as.factor(data1$Y)
+
+  c(AUC = auc(scores, response), OR = or(scores, response))
+
+}
+
+naiveest <- function(n = 2000, d = 500){
+
+  dat0 <- createnosignaldata(n, d)
+
+  train <- dat0
+
+  selecttr <- selectvars(train)
+  fit <- runclassifier(train, selecttr)
+  fitclassifier(fit, train)
+
+}
+
+zhutest.hold <- function(trratio = .5, n = 2000, d = 500){
+
+  dat0 <- createnosignaldata(n, d)
+  npart.tr <- floor(trratio * n)
+  tr.dex <- sample(1:n, npart.tr)
+  ho.dex <- setdiff(1:n, tr.dex)
+
+  train <- dat0[tr.dex, ]
+  hold <- dat0[ho.dex, ]
+
+  selecttr <- selectvars(dat0) ## selection performed on entire dataset
+  fit <- runclassifier(train, selecttr)
[…]

### 2016-07-05 Code/cv-sims.R

@@ -1,175 +1,0 @@
-
-args <- commandArgs(TRUE)
-print(args)
-ntasks <- as.numeric(args[1])
-
-library(glmnet)
-library(snow)
-cl <- makeCluster(ntasks, type = "MPI")
-
-clusterEvalQ(cl, {
-
-createnosignaldata <- function(n, d){
-
-  data.frame(Y = rbinom(n, 1, p = .3), X = matrix(rnorm(n * d), nrow = n))
-
-}
-
-selectvars <- function(data0){
-
-  pees <- sapply(data0[, -1], function(v1){
-
-    fit0 <- glm(data0$Y ~ v1, family = "binomial", model = FALSE, y = FALSE)
-    summary(fit0)$coefficients[2, 4]
-
-  })
-
-  order(pees)[1:25]
-
-}
-
-runclassifier <- function(data0, selected){
-
-  dat2 <- data0[, c(1, selected + 1)]
-  fit1 <- glm(Y ~ ., data = dat2, family = "binomial", model = FALSE, y = FALSE)
-
-  fit1
-
-}
-
-fitclassifier <- function(fit1, data1){
-
-  scores <- predict(fit1, newdata = data1, type = "response")
-  response <- as.factor(data1$Y)
-
-  AUC::auc(AUC::roc(scores, response))
-
-}
-
-zhutest.hold <- function(trratio = .5, n = 2000, d = 500){
-
-  dat0 <- createnosignaldata(n, d)
-  npart.tr <- floor(trratio * n)
-  tr.dex <- sample(1:n, npart.tr)
-  ho.dex <- setdiff(1:n, tr.dex)
-
-  train <- dat0[tr.dex, ]
-  hold <- dat0[ho.dex, ]
-
-  selecttr <- selectvars(dat0) ## selection performed on entire dataset
-  fit <- runclassifier(train, selecttr)
-  fitclassifier(fit, hold)
-
-}
-
-
-holdoutest <- function(trratio = .5, n = 2000, d = 500){
-
-  dat0 <- createnosignaldata(n, d)
-  npart.tr <- floor(trratio * n)
-  tr.dex <- sample(1:n, npart.tr)
-  ho.dex <- setdiff(1:n, tr.dex)
-
-  train <- dat0[tr.dex, ]
-  hold <- dat0[ho.dex, ]
-
-  selecttr <- selectvars(train)
[…]
